---
title: "Filter Samples (*`r params$species.name`*)"
author: "Bucao, Christabel Floi"
output: html_document
date: "`r Sys.Date()`"
params:
  counts.path: "../../data/counts"
  sample.metadata.path: "../../data/sample_metadata"
  gene.metadata.path: "../../data/gene_metadata"
  output.rds.path: "../../results/01_Filter_Samples"
  umi.counts: TRUE
  species: "DRE"
  species.name: "Danio rerio"
  min.genes: 10000
  min.reads: 500000
  min.replicates: 4
  min.cpm: 1.0
  normalize.by.condition: TRUE
  set.scale: TRUE
---

# Setup {.tabset}

## Parameters
```{r print-params}
print(params)
```

## Libraries
```{r import-libraries}
# Install sva
packages <- c("dplyr","sva")

for (p in packages) {
  if (!require(p, character.only=TRUE)) {
    print(paste("Installing", p))
    install.packages(p, character.only=TRUE)
    renv::snapshot()
  }
  print(paste("Loading", p))
  library(p, character.only=TRUE)
}
```

## Functions
```{r source-functions}
source("../functions/helper_functions.R")
source("../functions/filter_samples.R")
source("../functions/filter_samples_plots.R")
```

## Session Info
```{r print-session-info}
sessionInfo()
```

## Data

### Metadata
```{r import-metadata}
# Sample metadata
metadata <- 
  read.delim(file.path(params$sample.metadata.path, 
                       paste(params$species, "sample_annotation.tsv", sep="_")), 
             sep="\t") %>%
  dplyr::arrange(Tissue, IndID, .by_group=TRUE)

# Gene metadata
biotype <- 
  read.delim(file.path(params$gene.metadata.path, 
                       paste(params$species, "biotype_Ensembl_105.txt", sep="_")), 
             sep="\t")
```

### UMI counts
```{r import-counts}
# Load counts
counts <- 
  vector("list", length(levels(as.factor(metadata$ProjectSeqBatch))))
names(counts) <- levels(as.factor(metadata$ProjectSeqBatch))

if (params$umi.counts==TRUE) {
  counts <- lapply(setNames(names(counts),names(counts)), function(batch) {
    read.delim(file.path(params$counts.path, 
                         paste(params$species, batch, "UMI_counts.txt", sep="_")), 
               sep="\t")
  })  
} else {
  counts <- lapply(setNames(names(counts),names(counts)), function(batch) {
    read.delim(file.path(params$counts.path, 
                         paste(params$species, batch, "counts.txt", sep="_")), 
               sep="\t")
  })
}

# Map barcodes to sample names
counts <- 
  lapply(setNames(names(counts), names(counts)), function(batch) {
    map_barcodes_to_sample_names(counts[[batch]], 
                                 metadata[metadata$ProjectSeqBatch==batch,])
  }) %>%
  Reduce(function(...) merge(..., by="GeneID"), .) %>%
  tibble::column_to_rownames("GeneID") %>%
  as.matrix()

counts <- counts[,metadata$SampleName]

data.init <- list("metadata"=metadata, "counts"=counts)
rm(counts, metadata)
```

# Main {.tabset}

## Before filtering

Summarize number of mapped reads and number of detected genes (counts > 0)
```{r summarize-sequencing-results}
data.init$metadata <- 
  dplyr::inner_join(data.init$metadata, 
                    summarize_sequencing_results(data.init$counts), by="SampleName")
```

### Plot - Detected genes x Mapped reads
```{r plot-detected-genes-vs-mapped-reads, fig.align='center', fig.height=5}
plot_detected_genes_vs_mapped_reads(
  data.init$metadata, params$min.genes, params$min.reads, params$set.scale)
```

### Initial gene filtering and normalization

#### No batch correction
```{r initial-gene-filtering-and-normalization}
data.init$counts.gf <-
  filter_genes_by_cpm_per_condition(data.init$counts, data.init$metadata, min.cpm=params$min.cpm)

# Normalize across all conditions
data.init$tmm.cpm.gf <- 
  transform_counts_to_tmm_cpm(data.init$counts.gf, data.init$metadata, log2=FALSE)

data.init$log2.tmm.cpm.gf <- 
  transform_counts_to_tmm_cpm(data.init$counts.gf, data.init$metadata, 
                              log2=TRUE, shift.min=TRUE)
```

#### With batch correction
```{r}
data.init.combat <- vector(mode="list", length=6)
names(data.init.combat) <- c("conditions","metadata","counts","counts.gf","tmm.cpm.gf","log2.tmm.cpm.gf")

data.init.combat$conditions <- data.init$conditions
data.init.combat$metadata <- data.init$metadata

# Batch correction on raw count matrix
#data.init.combat$counts <- ComBat_seq(counts=data.init$counts, batch=data.init$metadata$Batch, group=NULL)
if ("Sex" %in% colnames(data.init$metadata)) {
  data.init.combat$counts <- 
    ComBat_seq(counts=data.init$counts,
               batch=data.init$metadata$Batch,
               group=paste(data.init$metadata$Tissue, data.init$metadata$Sex, sep="_"))  
} else {
  data.init.combat$counts <- 
    ComBat_seq(counts=data.init$counts,
               batch=data.init$metadata$Batch,
               group=data.init$metadata$Tissue)  
}

data.init.combat$counts.gf <-
  filter_genes_by_cpm_per_condition(data.init.combat$counts, data.init.combat$metadata, min.cpm=params$min.cpm)

# Normalize across all conditions
data.init.combat$tmm.cpm.gf <- 
  transform_counts_to_tmm_cpm(data.init.combat$counts.gf, data.init.combat$metadata, log2=FALSE)

data.init.combat$log2.tmm.cpm.gf <- 
  transform_counts_to_tmm_cpm(data.init.combat$counts.gf, data.init.combat$metadata, 
                              log2=TRUE, shift.min=TRUE)
```


### Plot - PCA {.tabset}

Check sample clustering before removing low quality samples
```{r plot-initial-pca, fig.align='center'}
data.init$pca <- run_pca(data.init$log2.tmm.cpm.gf)
data.init.combat$pca <- run_pca(data.init.combat$log2.tmm.cpm.gf)
```

#### By mapped reads
```{r plot-initial-pca-mapped-reads, fig.align='center'}
plot_pca_by_mapped_reads(data.init$pca, data.init$metadata, params$min.reads)
plot_pca_by_mapped_reads(data.init.combat$pca, data.init.combat$metadata, params$min.reads)

plot_pca_by_mapped_reads_gradient(data.init$pca, data.init$metadata)
plot_pca_by_mapped_reads_gradient(data.init.combat$pca, data.init.combat$metadata)
```

#### By detected genes
```{r plot-initial-pca-detected-genes, fig.align='center'}
plot_pca_by_detected_genes(data.init$pca, data.init$metadata, params$min.genes)
plot_pca_by_detected_genes(data.init.combat$pca, data.init.combat$metadata, params$min.genes)

plot_pca_by_detected_genes_gradient(data.init$pca, data.init$metadata)
plot_pca_by_detected_genes_gradient(data.init.combat$pca, data.init.combat$metadata)
```

#### By batch
```{r plot-initial-pca-batch, fig.align='center'}
plot_pca_by_batch(data.init$pca, data.init$metadata)
plot_pca_by_batch(data.init.combat$pca, data.init.combat$metadata)
```

#### By condition
```{r plot-initial-pca-condition, fig.align='center'}
plot_pca_by_condition(data.init$pca, data.init$metadata)
plot_pca_by_condition(data.init.combat$pca, data.init.combat$metadata)

plot_pca_by_condition_pc34(data.init$pca, data.init$metadata)
plot_pca_by_condition_pc34(data.init.combat$pca, data.init.combat$metadata)
```

## Filter 1: Sequencing quality

### Remove samples with less than minimum detected genes OR less than minimum mapped reads

#### No batch correction
```{r filter-samples-by-detected-genes-mapped-reads, fig.align='center', fig.height=5}
data.qc1.seq <- 
  filter_samples_by_detected_genes_mapped_reads(
    data.init$counts, data.init$metadata, params$min.genes, params$min.reads)

# Remove tissue-sex conditions with few replicates remaining
data.qc1.seq <-
  filter_conditions_by_min_replicates(
    data.qc1.seq$counts, data.qc1.seq$metadata, params$min.replicates)

data.qc1.seq$counts.gf <-
  filter_genes_by_cpm_per_condition(data.qc1.seq$counts, data.qc1.seq$metadata, min.cpm=params$min.cpm)

# Normalize across all conditions
data.qc1.seq$tmm.cpm.gf <- 
  transform_counts_to_tmm_cpm(data.qc1.seq$counts.gf, data.qc1.seq$metadata, log2=FALSE)

data.qc1.seq$log2.tmm.cpm.gf <- 
  transform_counts_to_tmm_cpm(data.qc1.seq$counts.gf, data.qc1.seq$metadata,
                              log2=TRUE, shift.min=TRUE)

plot_detected_genes_vs_mapped_reads(data.qc1.seq$metadata, params$min.genes, params$min.reads)
```

#### With batch correction
```{r filter-samples-by-detected-genes-mapped-reads-combat, fig.align='center', fig.height=5}
data.qc1.seq.combat <- list()
data.qc1.seq.combat$conditions <- data.qc1.seq$conditions
data.qc1.seq.combat$metadata <- data.qc1.seq$metadata

# Batch correction
#data.qc1.seq.combat$counts <- ComBat_seq(counts=data.qc1.seq$counts,
#                                         batch=data.qc1.seq$metadata$Batch)
if ("Sex" %in% colnames(data.qc1.seq$metadata)) {
  data.qc1.seq.combat$counts <- 
    ComBat_seq(counts=data.qc1.seq$counts,
               batch=data.qc1.seq$metadata$Batch,
               group=paste(data.qc1.seq$metadata$Tissue, data.qc1.seq$metadata$Sex, sep="_"))  
} else {
  data.qc1.seq.combat$counts <- 
    ComBat_seq(counts=data.qc1.seq$counts,
               batch=data.qc1.seq$metadata$Batch,
               group=data.qc1.seq$metadata$Tissue)    
}

data.qc1.seq.combat$counts.gf <-
  filter_genes_by_cpm_per_condition(data.qc1.seq.combat$counts, data.qc1.seq.combat$metadata, min.cpm=params$min.cpm)

# Normalize across all conditions
data.qc1.seq.combat$tmm.cpm.gf <- 
  transform_counts_to_tmm_cpm(data.qc1.seq.combat$counts.gf, data.qc1.seq.combat$metadata, log2=FALSE)

data.qc1.seq.combat$log2.tmm.cpm.gf <- 
  transform_counts_to_tmm_cpm(data.qc1.seq.combat$counts.gf, data.qc1.seq.combat$metadata,
                              log2=TRUE, shift.min=TRUE)

plot_detected_genes_vs_mapped_reads(data.qc1.seq.combat$metadata, params$min.genes, params$min.reads)
```

### Plot - PCA {.tabset}
```{r plot-pca-sequencing-quality-filter, fig.align='center'}
data.qc1.seq$pca <- run_pca(data.qc1.seq$log2.tmm.cpm.gf)
data.qc1.seq.combat$pca <- run_pca(data.qc1.seq.combat$log2.tmm.cpm.gf)
```

#### By mapped reads
```{r plot-pca-sequencing-quality-filter-mapped-reads, fig.align='center'}
plot_pca_by_mapped_reads_gradient(data.qc1.seq$pca, data.qc1.seq$metadata)
plot_pca_by_mapped_reads_gradient(data.qc1.seq.combat$pca, data.qc1.seq.combat$metadata)

plot_pca_by_detected_genes_gradient(data.qc1.seq$pca, data.qc1.seq$metadata)
plot_pca_by_detected_genes_gradient(data.qc1.seq.combat$pca, data.qc1.seq.combat$metadata)
```

#### By detected genes
```{r plot-pca-sequencing-quality-filter-detected-genes, fig.align='center'}
plot_pca_by_detected_genes_gradient(data.qc1.seq$pca, data.qc1.seq$metadata)
plot_pca_by_detected_genes_gradient(data.qc1.seq.combat$pca, data.qc1.seq.combat$metadata)
```

#### By batch
```{r plot-pca-sequencing-quality-filter-batch, fig.align='center'}
plot_pca_by_batch(data.qc1.seq$pca, data.qc1.seq$metadata)
plot_pca_by_batch(data.qc1.seq.combat$pca, data.qc1.seq.combat$metadata)
```

#### By condition
```{r plot-pca-sequencing-quality-filter-condition, fig.align='center'}
plot_pca_by_condition(data.qc1.seq$pca, data.qc1.seq$metadata)
plot_pca_by_condition(data.qc1.seq.combat$pca, data.qc1.seq.combat$metadata)

plot_pca_by_condition_pc34(data.qc1.seq$pca, data.qc1.seq$metadata)
plot_pca_by_condition_pc34(data.qc1.seq.combat$pca, data.qc1.seq.combat$metadata)
```

### Plot - Correlation heatmap
```{r plot-heatmap-sequencing-quality-filter, fig.align='center', fig.width = 12, fig.height = 8}
data.qc1.seq$cor.matrix <- cor(data.qc1.seq$log2.tmm.cpm.gf, method="pearson")
plot_correlation_heatmap(data.qc1.seq$cor.matrix, data.qc1.seq$metadata, clust=TRUE)

data.qc1.seq.combat$cor.matrix <- cor(data.qc1.seq.combat$log2.tmm.cpm.gf, method="pearson")
plot_correlation_heatmap(data.qc1.seq.combat$cor.matrix, data.qc1.seq.combat$metadata, clust=TRUE)
```

## Filter 2: Within-tissue correlation

### Iterative removal of low correlation samples

#### No batch correction
```{r}
data.qc2.cor <- 
  iterate_filter_samples_by_correlation(
    data.qc1.seq$counts, data.qc1.seq$metadata, min.cpm=params$min.cpm, round=2)

# Remove tissue-sex conditions with few replicates remaining
data.qc2.cor <- 
  filter_conditions_by_min_replicates(
    data.qc2.cor$counts, data.qc2.cor$metadata, params$min.replicates)

data.qc2.cor$counts.gf <-
  filter_genes_by_cpm_per_condition(
    data.qc2.cor$counts, data.qc2.cor$metadata, min.cpm=params$min.cpm)

# Normalize across all conditions
data.qc2.cor$tmm.cpm.gf <- 
  transform_counts_to_tmm_cpm(data.qc2.cor$counts.gf, data.qc2.cor$metadata, log2=FALSE)

data.qc2.cor$log2.tmm.cpm.gf <- 
  transform_counts_to_tmm_cpm(data.qc2.cor$counts.gf, data.qc2.cor$metadata, 
                              log2=TRUE, shift.min=TRUE)
```

#### With batch correction
```{r filter-samples-by-correlation}
data.qc2.cor.combat <- list()
data.qc2.cor.combat$conditions <- data.qc2.cor$conditions
data.qc2.cor.combat$metadata <- data.qc2.cor$metadata

# Batch correction
#data.qc2.cor.combat$counts <-
#  ComBat_seq(counts=data.qc2.cor$counts,
#             batch=data.qc2.cor$metadata$Batch)
if ("Sex" %in% colnames(data.qc2.cor$metadata)) {
  data.qc2.cor.combat$counts <- 
    ComBat_seq(counts=data.qc2.cor$counts,
               batch=data.qc2.cor$metadata$Batch,
               group=paste(data.qc2.cor$metadata$Tissue, data.qc2.cor$metadata$Sex, sep="_"))  
} else {
  data.qc2.cor.combat$counts <- 
    ComBat_seq(counts=data.qc2.cor$counts,
               batch=data.qc2.cor$metadata$Batch,
               group=data.qc2.cor$metadata$Tissue)    
}

data.qc2.cor.combat$counts.gf <-
  filter_genes_by_cpm_per_condition(
    data.qc2.cor.combat$counts, data.qc2.cor.combat$metadata, min.cpm=params$min.cpm)

# Normalize across all conditions
data.qc2.cor.combat$tmm.cpm.gf <- 
  transform_counts_to_tmm_cpm(data.qc2.cor.combat$counts.gf, data.qc2.cor.combat$metadata, log2=FALSE)

data.qc2.cor.combat$log2.tmm.cpm.gf <- 
  transform_counts_to_tmm_cpm(data.qc2.cor.combat$counts.gf, data.qc2.cor.combat$metadata, 
                              log2=TRUE, shift.min=TRUE)
```

### Plot - PCA {.tabset}
```{r plot-pca-correlation-filter, fig.align='center'}
data.qc2.cor$pca <- run_pca(data.qc2.cor$log2.tmm.cpm.gf)
data.qc2.cor.combat$pca <- run_pca(data.qc2.cor.combat$log2.tmm.cpm.gf)
```

#### By batch
```{r plot-pca-correlation-filter-batch, fig.align='center'}
plot_pca_by_batch(data.qc2.cor$pca, data.qc2.cor$metadata)
plot_pca_by_batch(data.qc2.cor.combat$pca, data.qc2.cor.combat$metadata)
```

#### By condition
```{r plot-pca-correlation-filter-condition, fig.align='center'}
plot_pca_by_condition(data.qc2.cor$pca, data.qc2.cor$metadata)
plot_pca_by_condition(data.qc2.cor.combat$pca, data.qc2.cor.combat$metadata)

plot_pca_by_condition_pc34(data.qc2.cor$pca, data.qc2.cor$metadata)
plot_pca_by_condition_pc34(data.qc2.cor.combat$pca, data.qc2.cor.combat$metadata)
```

### Plot - Correlation heatmap
```{r plot-heatmap-correlation-filter, fig.align='center', fig.width = 12, fig.height = 8}
data.qc2.cor$cor.matrix <- cor(data.qc2.cor$log2.tmm.cpm.gf, method="pearson")
plot_correlation_heatmap(data.qc2.cor$cor.matrix, data.qc2.cor$metadata, clust=TRUE)

data.qc2.cor.combat$cor.matrix <- cor(data.qc2.cor.combat$log2.tmm.cpm.gf, method="pearson")
plot_correlation_heatmap(data.qc2.cor.combat$cor.matrix, data.qc2.cor.combat$metadata, clust=TRUE)
```

### Boxplot - Detected genes per organ
```{r boxplot-detected-genes-correlation-filter, fig.width=18, fig.height=10}
boxplot_detected_genes(data.qc2.cor$metadata)
boxplot_detected_genes(data.qc2.cor.combat$metadata)
```


## Gene biotype

Split genes by biotype

### No batch correction
```{r split-by-gene-biotype}
data.qc2.cor$coding$log2.tmm.cpm.gf <-
  data.qc2.cor$log2.tmm.cpm.gf[
    rownames(data.qc2.cor$log2.tmm.cpm.gf) 
        %in% biotype$Gene.stable.ID[biotype$Gene.type=="protein_coding"],]

data.qc2.cor$lncrna$log2.tmm.cpm.gf <-
  data.qc2.cor$log2.tmm.cpm.gf[
    rownames(data.qc2.cor$log2.tmm.cpm.gf) 
        %in% biotype$Gene.stable.ID[biotype$Gene.type=="lncRNA" | 
                                      biotype$Gene.type=="lincRNA"],]

# Protein-coding
print(nrow(data.qc2.cor$coding$log2.tmm.cpm.gf))
# lncRNA
print(nrow(data.qc2.cor$lncrna$log2.tmm.cpm.gf))
# Neither
nrow(data.qc2.cor$log2.tmm.cpm.gf) - 
  (nrow(data.qc2.cor$coding$log2.tmm.cpm.gf) + nrow(data.qc2.cor$lncrna$log2.tmm.cpm.gf))
```

### With batch correction
```{r}
data.qc2.cor.combat$coding$log2.tmm.cpm.gf <-
  data.qc2.cor.combat$log2.tmm.cpm.gf[
    rownames(data.qc2.cor.combat$log2.tmm.cpm.gf) 
        %in% biotype$Gene.stable.ID[biotype$Gene.type=="protein_coding"],]

data.qc2.cor.combat$lncrna$log2.tmm.cpm.gf <-
  data.qc2.cor.combat$log2.tmm.cpm.gf[
    rownames(data.qc2.cor.combat$log2.tmm.cpm.gf) 
        %in% biotype$Gene.stable.ID[biotype$Gene.type=="lncRNA" | 
                                      biotype$Gene.type=="lincRNA"],]

# Protein-coding
print(nrow(data.qc2.cor.combat$coding$log2.tmm.cpm.gf))
# lncRNA
print(nrow(data.qc2.cor.combat$lncrna$log2.tmm.cpm.gf))
# Neither
nrow(data.qc2.cor.combat$log2.tmm.cpm.gf) - 
  (nrow(data.qc2.cor.combat$coding$log2.tmm.cpm.gf) + nrow(data.qc2.cor.combat$lncrna$log2.tmm.cpm.gf))
```


### Protein-coding

#### Plot - PCA {.tabset}
```{r plot-pca-protein-coding, fig.align='center'}
data.qc2.cor$coding$pca <- run_pca(data.qc2.cor$coding$log2.tmm.cpm.gf)
data.qc2.cor.combat$coding$pca <- run_pca(data.qc2.cor.combat$coding$log2.tmm.cpm.gf)
```

##### By batch
```{r plot-pca-protein-coding-batch, fig.align='center'}
plot_pca_by_batch(data.qc2.cor$coding$pca, data.qc2.cor$metadata)
plot_pca_by_batch(data.qc2.cor.combat$coding$pca, data.qc2.cor.combat$metadata)
```

##### By condition
```{r plot-pca-protein-coding-condition, fig.align='center'}
plot_pca_by_condition(data.qc2.cor$coding$pca, data.qc2.cor$metadata)
plot_pca_by_condition(data.qc2.cor.combat$coding$pca, data.qc2.cor.combat$metadata)

plot_pca_by_condition_pc34(data.qc2.cor$coding$pca, data.qc2.cor$metadata)
plot_pca_by_condition_pc34(data.qc2.cor.combat$coding$pca, data.qc2.cor.combat$metadata)
```

#### Plot - Correlation heatmap
```{r plot-heatmap-protein-coding, fig.align='center', fig.width = 12, fig.height = 8}
data.qc2.cor$coding$cor.matrix <- 
  cor(data.qc2.cor$coding$log2.tmm.cpm.gf, method="pearson")
plot_correlation_heatmap(data.qc2.cor$coding$cor.matrix, 
                         data.qc2.cor$metadata, clust=TRUE)

data.qc2.cor.combat$coding$cor.matrix <-
  cor(data.qc2.cor.combat$coding$log2.tmm.cpm.gf, method="pearson")
plot_correlation_heatmap(data.qc2.cor.combat$coding$cor.matrix,
                         data.qc2.cor.combat$metadata, clust=TRUE)
```

### lncRNA

#### Plot - PCA {.tabset}
```{r plot-pca-lncrna, fig.align='center'}
data.qc2.cor$lncrna$pca <- run_pca(data.qc2.cor$lncrna$log2.tmm.cpm.gf)
data.qc2.cor.combat$lncrna$pca <- run_pca(data.qc2.cor.combat$lncrna$log2.tmm.cpm.gf)
```

##### By batch
```{r plot-pca-lncrna-batch, fig.align='center'}
plot_pca_by_batch(data.qc2.cor$lncrna$pca, data.qc2.cor$metadata)
plot_pca_by_batch(data.qc2.cor.combat$lncrna$pca, data.qc2.cor.combat$metadata)
```

##### By condition
```{r plot-pca-lncrna-condition, fig.align='center'}
plot_pca_by_condition(data.qc2.cor$lncrna$pca, data.qc2.cor$metadata)
plot_pca_by_condition(data.qc2.cor.combat$lncrna$pca, data.qc2.cor.combat$metadata)

plot_pca_by_condition_pc34(data.qc2.cor$lncrna$pca, data.qc2.cor$metadata)
plot_pca_by_condition_pc34(data.qc2.cor.combat$lncrna$pca, data.qc2.cor.combat$metadata)
```

#### Plot - Correlation heatmap
```{r plot-heatmap-lncrna, fig.align='center', fig.width = 12, fig.height = 8}
data.qc2.cor$lncrna$cor.matrix <- 
  cor(data.qc2.cor$lncrna$log2.tmm.cpm.gf, method="pearson")
plot_correlation_heatmap(data.qc2.cor$lncrna$cor.matrix, 
                         data.qc2.cor$metadata, clust=TRUE)

data.qc2.cor.combat$lncrna$cor.matrix <-
  cor(data.qc2.cor.combat$lncrna$log2.tmm.cpm.gf, method="pearson")
plot_correlation_heatmap(data.qc2.cor.combat$lncrna$cor.matrix,
                         data.qc2.cor.combat$metadata, clust=TRUE)
```

## Reduced (all nonzero) expression matrix

### No batch correction
```{r keep-all-nonzero-counts-only}
data.qc2.cor.nonzero <- data.qc2.cor[c("conditions","metadata","counts")]

# Only keep genes with no zero counts across all samples
# to set the same number of 'expressed genes' per condition
data.qc2.cor.nonzero$counts.gf <-
  keep_full_nonzero_expression_matrix(data.qc2.cor.nonzero$counts)

# Normalize across all conditions
data.qc2.cor.nonzero$tmm.cpm.gf <-
  transform_counts_to_tmm_cpm(data.qc2.cor.nonzero$counts.gf, data.qc2.cor.nonzero$metadata, log2=FALSE)

data.qc2.cor.nonzero$log2.tmm.cpm.gf <- 
  transform_counts_to_tmm_cpm(data.qc2.cor.nonzero$counts.gf, data.qc2.cor.nonzero$metadata, 
                              log2=TRUE, shift.min=FALSE)
```

### With batch correction
```{r keep-all-nonzero-counts-only-combat}
data.qc2.cor.nonzero.combat <- data.qc2.cor.combat[c("conditions","metadata","counts")]

# Only keep genes with no zero counts across all samples
# to set the same number of 'expressed genes' per condition
data.qc2.cor.nonzero.combat$counts.gf <-
  keep_full_nonzero_expression_matrix(data.qc2.cor.nonzero.combat$counts)

# Normalize across all conditions
data.qc2.cor.nonzero.combat$tmm.cpm.gf <-
  transform_counts_to_tmm_cpm(data.qc2.cor.nonzero.combat$counts.gf, data.qc2.cor.nonzero.combat$metadata, log2=FALSE)

data.qc2.cor.nonzero.combat$log2.tmm.cpm.gf <- 
  transform_counts_to_tmm_cpm(data.qc2.cor.nonzero.combat$counts.gf, data.qc2.cor.nonzero.combat$metadata, 
                              log2=TRUE, shift.min=FALSE)
```

### Plot - PCA {.tabset}
```{r plot-pca-correlation-filter-nonzero, fig.align='center'}
data.qc2.cor.nonzero$pca <- run_pca(data.qc2.cor.nonzero$log2.tmm.cpm.gf)
data.qc2.cor.nonzero.combat$pca <- run_pca(data.qc2.cor.nonzero.combat$log2.tmm.cpm.gf)
```

#### By batch
```{r plot-pca-correlation-filter-nonzero-batch, fig.align='center'}
plot_pca_by_batch(data.qc2.cor.nonzero$pca, data.qc2.cor.nonzero$metadata)
plot_pca_by_batch(data.qc2.cor.nonzero.combat$pca, data.qc2.cor.nonzero.combat$metadata)
```

#### By condition
```{r plot-pca-correlation-filter-nonzero-condition, fig.align='center'}
plot_pca_by_condition(data.qc2.cor.nonzero$pca, data.qc2.cor.nonzero$metadata)
plot_pca_by_condition(data.qc2.cor.nonzero.combat$pca, data.qc2.cor.nonzero.combat$metadata)

plot_pca_by_condition_pc34(data.qc2.cor.nonzero$pca, data.qc2.cor.nonzero$metadata)
plot_pca_by_condition_pc34(data.qc2.cor.nonzero.combat$pca, data.qc2.cor.nonzero.combat$metadata)
```


### Plot - Correlation heatmap
```{r plot-heatmap-correlation-filter-nonzero, fig.align='center', fig.width = 12, fig.height = 8}
data.qc2.cor.nonzero$cor.matrix <- cor(data.qc2.cor.nonzero$log2.tmm.cpm.gf, method="pearson")
plot_correlation_heatmap(data.qc2.cor.nonzero$cor.matrix, data.qc2.cor.nonzero$metadata, clust=TRUE)

data.qc2.cor.nonzero.combat$cor.matrix <- cor(data.qc2.cor.nonzero.combat$log2.tmm.cpm.gf, method="pearson")
plot_correlation_heatmap(data.qc2.cor.nonzero.combat$cor.matrix, data.qc2.cor.nonzero.combat$metadata, clust=TRUE)
```


## Normalization per condition

### No batch correction
```{r normalize-per-condition, eval=params$normalize.by.condition}
data.conditions <- vector(mode="list", length=6)
names(data.conditions) <- 
  c("conditions","metadata","counts","counts.gf","tmm.cpm.gf","log2.tmm.cpm.gf")

data.conditions$conditions <- data.qc2.cor$conditions
data.conditions$metadata <- data.qc2.cor$metadata

Tissue <- as.factor(data.qc2.cor$metadata$Tissue)

if ("Sex" %in% colnames(data.qc2.cor$metadata)) {
  Sex <- as.factor(data.qc2.cor$metadata$Sex)
  
  data.conditions$counts <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        subset_expr_matrix(data.qc2.cor$counts, tissue=t, sex=s)
      })
    })

  data.conditions$counts.gf <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        filter_genes_by_cpm(data.conditions$counts[[t]][[s]], 
                            min.cpm=params$min.cpm, tmm=FALSE)
      })
    })
  
  data.conditions$tmm.cpm.gf <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        transform_counts_to_tmm_cpm(
          data.conditions$counts.gf[[t]][[s]],
          data.conditions$metadata[data.conditions$metadata$Tissue==t &
                                  data.conditions$metadata$Sex==s,],
          log2=FALSE)
      })
    })
    
  data.conditions$log2.tmm.cpm.gf <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        transform_counts_to_tmm_cpm(
          data.conditions$counts.gf[[t]][[s]],
          data.conditions$metadata[data.conditions$metadata$Tissue==t &
                                  data.conditions$metadata$Sex==s,],
          log2=TRUE, shift.min=TRUE)
      })
    })
  
} else {
  data.conditions$counts <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      subset_expr_matrix(data.qc2.cor$counts, tissue=t)
    })
 
  data.conditions$counts.gf <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      filter_genes_by_cpm(data.conditions$counts[[t]], 
                          min.cpm=params$min.cpm, tmm=FALSE)
    })
  
  data.conditions$tmm.cpm.gf <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      transform_counts_to_tmm_cpm(
        data.conditions$counts.gf[[t]],
          data.conditions$metadata[data.conditions$metadata$Tissue==t,],
          log2=FALSE)
    })
  
  data.conditions$log2.tmm.cpm.gf <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      transform_counts_to_tmm_cpm(
        data.conditions$counts.gf[[t]],
          data.conditions$metadata[data.conditions$metadata$Tissue==t,],
          log2=TRUE, shift.min=TRUE)
    })
}
```

### With batch correction
```{r normalize-per-condition-combat}
data.conditions.combat <- vector(mode="list", length=6)
names(data.conditions.combat) <- 
  c("conditions","metadata","counts","counts.gf","tmm.cpm.gf","log2.tmm.cpm.gf")

data.conditions.combat$conditions <- data.qc2.cor.combat$conditions
data.conditions.combat$metadata <- data.qc2.cor.combat$metadata

Tissue <- as.factor(data.qc2.cor.combat$metadata$Tissue)

if ("Sex" %in% colnames(data.qc2.cor.combat$metadata)) {
  Sex <- as.factor(data.qc2.cor.combat$metadata$Sex)
  
  data.conditions.combat$counts <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        subset_expr_matrix(data.qc2.cor.combat$counts, tissue=t, sex=s)
      })
    })

  data.conditions.combat$counts.gf <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        filter_genes_by_cpm(data.conditions.combat$counts[[t]][[s]], 
                            min.cpm=params$min.cpm, tmm=FALSE)
      })
    })
  
  data.conditions.combat$tmm.cpm.gf <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        transform_counts_to_tmm_cpm(
          data.conditions.combat$counts.gf[[t]][[s]],
          data.conditions.combat$metadata[data.conditions.combat$metadata$Tissue==t &
                                  data.conditions.combat$metadata$Sex==s,],
          log2=FALSE)
      })
    })
    
  data.conditions.combat$log2.tmm.cpm.gf <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        transform_counts_to_tmm_cpm(
          data.conditions.combat$counts.gf[[t]][[s]],
          data.conditions.combat$metadata[data.conditions.combat$metadata$Tissue==t &
                                  data.conditions.combat$metadata$Sex==s,],
          log2=TRUE, shift.min=TRUE)
      })
    })
  
} else {
  data.conditions.combat$counts <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      subset_expr_matrix(data.qc2.cor.combat$counts, tissue=t)
    })
 
  data.conditions.combat$counts.gf <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      filter_genes_by_cpm(data.conditions.combat$counts[[t]], 
                          min.cpm=params$min.cpm, tmm=FALSE)
    })
  
  data.conditions.combat$tmm.cpm.gf <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      transform_counts_to_tmm_cpm(
        data.conditions.combat$counts.gf[[t]],
          data.conditions.combat$metadata[data.conditions.combat$metadata$Tissue==t,],
          log2=FALSE)
    })
  
  data.conditions.combat$log2.tmm.cpm.gf <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      transform_counts_to_tmm_cpm(
        data.conditions.combat$counts.gf[[t]],
          data.conditions.combat$metadata[data.conditions.combat$metadata$Tissue==t,],
          log2=TRUE, shift.min=TRUE)
    })
}
```


### Reduced (all nonzero) expression matrix

#### No batch correction
```{r normalize-per-condition-nonzero, eval=params$normalize.by.condition}
data.conditions.nonzero <- vector(mode="list", length=6)
names(data.conditions.nonzero) <- 
  c("conditions","metadata","counts","counts.gf","tmm.cpm.gf","log2.tmm.cpm.gf")

data.conditions.nonzero$conditions <- data.qc2.cor.nonzero$conditions
data.conditions.nonzero$metadata <- data.qc2.cor.nonzero$metadata

Tissue <- as.factor(data.qc2.cor.nonzero$metadata$Tissue)

if ("Sex" %in% colnames(data.qc2.cor.nonzero$metadata)) {
  Sex <- as.factor(data.qc2.cor.nonzero$metadata$Sex)
  
  data.conditions.nonzero$counts <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        subset_expr_matrix(data.qc2.cor.nonzero$counts, tissue=t, sex=s)
      })
    })

  data.conditions.nonzero$counts.gf <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        filter_genes_by_cpm(data.conditions.nonzero$counts[[t]][[s]], 
                            min.cpm=params$min.cpm, tmm=FALSE)
      })
    })
  
  data.conditions.nonzero$tmm.cpm.gf <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        transform_counts_to_tmm_cpm(
          data.conditions.nonzero$counts.gf[[t]][[s]],
          data.conditions.nonzero$metadata[data.conditions.nonzero$metadata$Tissue==t &
                                  data.conditions.nonzero$metadata$Sex==s,],
          log2=FALSE)
      })
    })
    
  data.conditions.nonzero$log2.tmm.cpm.gf <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        transform_counts_to_tmm_cpm(
          data.conditions.nonzero$counts.gf[[t]][[s]],
          data.conditions.nonzero$metadata[data.conditions.nonzero$metadata$Tissue==t &
                                  data.conditions.nonzero$metadata$Sex==s,],
          log2=TRUE, shift.min=TRUE)
      })
    })
  
} else {
  data.conditions.nonzero$counts <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      subset_expr_matrix(data.qc2.cor.nonzero$counts, tissue=t)
    })
 
  data.conditions.nonzero$counts.gf <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      filter_genes_by_cpm(data.conditions.nonzero$counts[[t]], 
                          min.cpm=params$min.cpm, tmm=FALSE)
    })
  
  data.conditions.nonzero$tmm.cpm.gf <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      transform_counts_to_tmm_cpm(
        data.conditions.nonzero$counts.gf[[t]],
          data.conditions.nonzero$metadata[data.conditions.nonzero$metadata$Tissue==t,],
          log2=FALSE)
    })
  
  data.conditions.nonzero$log2.tmm.cpm.gf <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      transform_counts_to_tmm_cpm(
        data.conditions.nonzero$counts.gf[[t]],
          data.conditions.nonzero$metadata[data.conditions.nonzero$metadata$Tissue==t,],
          log2=TRUE, shift.min=TRUE)
    })
}
```

### With batch correction
```{r normalize-per-condition-nonzero-combat}
data.conditions.nonzero.combat <- vector(mode="list", length=6)
names(data.conditions.nonzero.combat) <- 
  c("conditions","metadata","counts","counts.gf","tmm.cpm.gf","log2.tmm.cpm.gf")

data.conditions.nonzero.combat$conditions <- data.qc2.cor.nonzero.combat$conditions
data.conditions.nonzero.combat$metadata <- data.qc2.cor.nonzero.combat$metadata

Tissue <- as.factor(data.qc2.cor.nonzero.combat$metadata$Tissue)

if ("Sex" %in% colnames(data.qc2.cor.nonzero.combat$metadata)) {
  Sex <- as.factor(data.qc2.cor.nonzero.combat$metadata$Sex)
  
  data.conditions.nonzero.combat$counts <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        subset_expr_matrix(data.qc2.cor.nonzero.combat$counts, tissue=t, sex=s)
      })
    })

  data.conditions.nonzero.combat$counts.gf <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        filter_genes_by_cpm(data.conditions.nonzero.combat$counts[[t]][[s]], 
                            min.cpm=params$min.cpm, tmm=FALSE)
      })
    })
  
  data.conditions.nonzero.combat$tmm.cpm.gf <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        transform_counts_to_tmm_cpm(
          data.conditions.nonzero.combat$counts.gf[[t]][[s]],
          data.conditions.nonzero.combat$metadata[data.conditions.nonzero.combat$metadata$Tissue==t &
                                  data.conditions.nonzero.combat$metadata$Sex==s,],
          log2=FALSE)
      })
    })
    
  data.conditions.nonzero.combat$log2.tmm.cpm.gf <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        transform_counts_to_tmm_cpm(
          data.conditions.nonzero.combat$counts.gf[[t]][[s]],
          data.conditions.nonzero.combat$metadata[data.conditions.nonzero.combat$metadata$Tissue==t &
                                  data.conditions.nonzero.combat$metadata$Sex==s,],
          log2=TRUE, shift.min=TRUE)
      })
    })
  
} else {
  data.conditions.nonzero.combat$counts <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      subset_expr_matrix(data.qc2.cor.nonzero.combat$counts, tissue=t)
    })
 
  data.conditions.nonzero.combat$counts.gf <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      filter_genes_by_cpm(data.conditions.nonzero.combat$counts[[t]], 
                          min.cpm=params$min.cpm, tmm=FALSE)
    })
  
  data.conditions.nonzero.combat$tmm.cpm.gf <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      transform_counts_to_tmm_cpm(
        data.conditions.nonzero.combat$counts.gf[[t]],
          data.conditions.nonzero.combat$metadata[data.conditions.nonzero.combat$metadata$Tissue==t,],
          log2=FALSE)
    })
  
  data.conditions.nonzero.combat$log2.tmm.cpm.gf <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      transform_counts_to_tmm_cpm(
        data.conditions.nonzero.combat$counts.gf[[t]],
          data.conditions.nonzero.combat$metadata[data.conditions.nonzero.combat$metadata$Tissue==t,],
          log2=TRUE, shift.min=TRUE)
    })
}
```


# Save
```{r save-objects-1}
if (!dir.exists(params$output.rds.path)) { dir.create(params$output.rds.path, showWarnings=FALSE) }

# Normalized across all conditions
saveRDS(data.qc2.cor, 
        file.path(params$output.rds.path, 
                  paste(params$species, 
                        params$min.replicates, "reps",
                        "data.qc2.cor.rds", sep="_")))

saveRDS(data.qc2.cor.combat, 
        file.path(params$output.rds.path, 
                  paste(params$species, 
                        params$min.replicates, "reps",
                        "data.qc2.cor.combat.rds", sep="_")))

saveRDS(data.qc2.cor.nonzero, 
        file.path(params$output.rds.path, 
                  paste(params$species, 
                        params$min.replicates, "reps",
                        "data.qc2.cor.nonzero.rds", sep="_")))

saveRDS(data.qc2.cor.nonzero.combat, 
        file.path(params$output.rds.path, 
                  paste(params$species, 
                        params$min.replicates, "reps",
                        "data.qc2.cor.nonzero.combat.rds", sep="_")))
```

```{r save-objects-2, eval=params$normalize.by.condition}
# Normalized per condition
saveRDS(data.conditions, 
        file.path(params$output.rds.path, 
                  paste(params$species,
                        params$min.replicates, "reps",
                        "data.conditions.rds", sep="_")))

saveRDS(data.conditions.combat, 
        file.path(params$output.rds.path, 
                  paste(params$species,
                        params$min.replicates, "reps",
                        "data.conditions.combat.rds", sep="_")))

saveRDS(data.conditions.nonzero, 
        file.path(params$output.rds.path, 
                  paste(params$species,
                        params$min.replicates, "reps",
                        "data.conditions.nonzero.rds", sep="_")))

saveRDS(data.conditions.nonzero.combat, 
        file.path(params$output.rds.path, 
                  paste(params$species,
                        params$min.replicates, "reps",
                        "data.conditions.nonzero.combat.rds", sep="_")))
```