---
title: "Variability - No Jackknife (*`r params$species.name`*)"
author: "Bucao, Christabel Floi"
output: html_document
date: "`r Sys.Date()`"
params:
  gene.metadata.path: "../../data/gene_metadata"  
  input.rds.path: "../../results/01_Filter_Samples"
  output.rds.path: "../../results/02b_Variability_NoJack"
  species: "DRE"
  species.name: "Danio rerio"
  min.percentile: 0.00
  max.percentile: 0.95
  win.size: 100
  qc1.only: FALSE
  all.nonzero.matrix: FALSE
  combat: FALSE
  run.example: FALSE
  ex.tissue: "brain"
  ex.sex: "F"
---

# Setup {.tabset}

## Parameters
```{r print-params}
print(params)
```

## Libraries
```{r import-libraries}
library(dplyr)
library(parallel)
```

## Functions
```{r source-functions}
source("../functions/variability_jackknife.R")
```

## Session Info
```{r print-session-info}
sessionInfo()
```


## Data

### Gene metadata
```{r import-metadata}
biotype <- 
  read.delim(file.path(params$gene.metadata.path, 
                       paste(params$species, "biotype_Ensembl_105.txt", sep="_")), 
             sep="\t")
```

### Processed expression data
```{r import-filtered-samples}
if (!params$combat) {
  if (!params$qc1.only) {
    input.path <- file.path(params$input.rds.path, "no_combat", "full_qc")
  } else {
    input.path <- file.path(params$input.rds.path, "no_combat", "qc1_only")    
  }
  
} else {
  if (!params$qc1.only) {
    input.path <- file.path(params$input.rds.path, "combat", "full_qc")    
  } else {
    input.path <- file.path(params$input.rds.path, "combat", "qc1_only")    
  }

}

# Load saved data from 01_Filter_Samples.Rmd
# Normalized by condition
if (!params$all.nonzero.matrix) {
  if (!params$qc1.only) {
    filtered.samples <- readRDS(
      file.path(input.path,
                paste(params$species, "4_reps_data.conditions.rds", sep="_")))        
  } else {
    filtered.samples <- readRDS(
      file.path(input.path,
                paste(params$species, "4_reps_data.qc1.conditions.rds", sep="_")))            
  } 


} else {
    filtered.samples <- readRDS(
      file.path(input.path,
                paste(params$species, "4_reps_data.conditions.nonzero.rds", sep="_")))    

}

Tissue <- as.factor(filtered.samples$metadata$Tissue)
if ("Sex" %in% colnames(filtered.samples$metadata)) { 
  Sex <- as.factor(filtered.samples$metadata$Sex) 
}
```

# Main

## Example: Compute variability metrics without jackknifing {.tabset}
```{r run-example, eval=params$run.example}
t <- params$ex.tissue
s <- params$ex.sex

# See variability_jackknife.R for descriptions of functions

# Different variability metrics
## Adjusted SD
var.adj.sd <- adjusted_sd(filtered.samples$log2.tmm.cpm.gf[[t]][[s]])
### Variability percentile rank within a sliding window of length win.size
var.adj.sd$rank.adj.sd <- 
  compute_local_ev_rank(var.adj.sd$mean, var.adj.sd$log2.adj.sd, win.size=100)

## Residual log2 (CV)^2
var.resid.cv <- residual_log2cv(filtered.samples$log2.tmm.cpm.gf[[t]][[s]])
var.resid.cv$rank.resid.log2.cv <- 
  compute_local_ev_rank(var.resid.cv$mean, var.resid.cv$resid.log2.cv, win.size=100)

## Residual log2 SD
var.resid.sd <- residual_log2sd(filtered.samples$log2.tmm.cpm.gf[[t]][[s]])
var.resid.sd$rank.resid.log2.sd <- 
  compute_local_ev_rank(var.resid.sd$mean, var.resid.sd$resid.log2.sd, win.size=100)

## Residual MAD
var.resid.mad <- residual_mad(filtered.samples$log2.tmm.cpm.gf[[t]][[s]])
var.resid.mad$rank.resid.mad <-
  compute_local_ev_rank(var.resid.mad$mean, var.resid.mad$resid.mad, win.size=100)
```

### Correlation between methods
```{r compare-variability-metrics, eval=params$run.example}
# Correlation between methods

## Adjusted SD vs residual log2 (CV)^2
cor.test(var.adj.sd$log2.adj.sd, var.resid.cv$resid.log2.cv)
cor.test(var.adj.sd$rank.adj.sd, var.resid.cv$rank.resid.log2.cv)

## Residual log2 (CV)^2 vs residual MAD
cor.test(var.resid.cv$resid.log2.cv, var.resid.mad$resid.mad)
cor.test(var.resid.cv$rank.resid.log2.cv, var.resid.mad$rank.resid.mad)

## Residual log2 SD vs residual MAD
cor.test(var.resid.sd$resid.log2.sd, var.resid.mad$resid.mad)
cor.test(var.resid.sd$rank.resid.log2.sd, var.resid.mad$rank.resid.mad)
```

## Compute variability metrics without jackknifing

### Adjusted SD
```{r run-adj-sd}
if ("Sex" %in% colnames(filtered.samples$metadata)) {
  adj.sd <- lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
    lapply(setNames(levels(Sex), levels(Sex)), function(s) {
      # Check if tissue-sex condition is included
      if (ncol(filtered.samples$log2.tmm.cpm.gf[[t]][[s]])==0) { NULL } 
      else {
        adjusted_sd(filtered.samples$log2.tmm.cpm.gf[[t]][[s]])
      }
    })
  })
} else {
  adj.sd <- lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
    adjusted_sd(filtered.samples$log2.tmm.cpm.gf[[t]])
  })
}

# Bind into data frame
if ("Sex" %in% colnames(filtered.samples$metadata)) {
  adj.sd.summary <- lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
    lapply(setNames(levels(Sex), levels(Sex)), function(s) {
      # Check if tissue-sex condition is included
      if (ncol(filtered.samples$log2.tmm.cpm.gf[[t]][[s]])==0) { NULL }
      else {
        data.frame("Mean"=adj.sd[[t]][[s]]$mean,
                   "SD"=adj.sd[[t]][[s]]$sd,
                   "AdjSD"=adj.sd[[t]][[s]]$adj.sd,
                   "Log2AdjSD"=adj.sd[[t]][[s]]$log2.adj.sd,
                   "Rank_Log2AdjSD"=compute_local_ev_rank(adj.sd[[t]][[s]]$mean,
                                                          adj.sd[[t]][[s]]$log2.adj.sd,
                                                          win.size=100),
                   row.names=rownames(filtered.samples$log2.tmm.cpm.gf[[t]][[s]]))
      }
    })
  })
} else {
  adj.sd.summary <- lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
    data.frame("Mean"=adj.sd[[t]]$mean,
               "SD"=adj.sd[[t]]$sd,
               "AdjSD"=adj.sd[[t]]$adj.sd,
               "Log2AdjSD"=adj.sd[[t]]$log2.adj.sd,
               "Rank_Log2AdjSD"=compute_local_ev_rank(adj.sd[[t]]$mean,
                                                      adj.sd[[t]]$log2.adj.sd,
                                                      win.size=100),
               row.names=rownames(filtered.samples$log2.tmm.cpm.gf[[t]]))
  })
}
```

### Residual log2 (SD)
```{r run-resid-log2sd}
if ("Sex" %in% colnames(filtered.samples$metadata)) {
  resid.sd <- lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
    lapply(setNames(levels(Sex), levels(Sex)), function(s) {
      # Check if tissue-sex condition is included
      if (ncol(filtered.samples$log2.tmm.cpm.gf[[t]][[s]])==0) { NULL } 
      else {
        residual_log2sd(filtered.samples$log2.tmm.cpm.gf[[t]][[s]])
      }
    })
  })
} else {
  resid.sd <- lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
    residual_log2sd(filtered.samples$log2.tmm.cpm.gf[[t]])
  })
}

# Bind into data frame
if ("Sex" %in% colnames(filtered.samples$metadata)) {
  resid.sd.summary <- lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
    lapply(setNames(levels(Sex), levels(Sex)), function(s) {
      # Check if tissue-sex condition is included
      if (ncol(filtered.samples$log2.tmm.cpm.gf[[t]][[s]])==0) { NULL }
      else {
        data.frame("Mean"=resid.sd[[t]][[s]]$mean,
                   "Log2SD"=resid.sd[[t]][[s]]$log2.sd,
                   "Resid_Log2SD"=resid.sd[[t]][[s]]$resid.log2.sd,
                   "Rank_Resid_Log2SD"=compute_local_ev_rank(resid.sd[[t]][[s]]$mean,
                                                             resid.sd[[t]][[s]]$resid.log2.sd,
                                                             win.size=100),
                   row.names=rownames(filtered.samples$log2.tmm.cpm.gf[[t]][[s]]))
      }
    })
  })
} else {
  resid.sd.summary <- lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
    data.frame("Mean"=resid.sd[[t]]$mean,
               "Log2SD"=resid.sd[[t]]$log2.sd,
               "Resid_Log2SD"=resid.sd[[t]]$resid.log2.sd,
               "Rank_Resid_Log2SD"=compute_local_ev_rank(resid.sd[[t]]$mean,
                                                         resid.sd[[t]]$resid.log2.sd,
                                                         win.size=100),
               row.names=rownames(filtered.samples$log2.tmm.cpm.gf[[t]]))
  })
}
```


### Residual log2 (CV)^2
```{r run-resid-log2cv}
if ("Sex" %in% colnames(filtered.samples$metadata)) {
  resid.cv <- lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
    lapply(setNames(levels(Sex), levels(Sex)), function(s) {
      # Check if tissue-sex condition is included
      if (ncol(filtered.samples$log2.tmm.cpm.gf[[t]][[s]])==0) { NULL } 
      else {
        residual_log2cv(filtered.samples$log2.tmm.cpm.gf[[t]][[s]])
      }
    })
  })
} else {
  resid.cv <- lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
    residual_log2cv(filtered.samples$log2.tmm.cpm.gf[[t]])
  })
}

# Bind into data frame
if ("Sex" %in% colnames(filtered.samples$metadata)) {
  resid.cv.summary <- lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
    lapply(setNames(levels(Sex), levels(Sex)), function(s) {
      # Check if tissue-sex condition is included
      if (ncol(filtered.samples$log2.tmm.cpm.gf[[t]][[s]])==0) { NULL }
      else {
        data.frame("Mean"=resid.cv[[t]][[s]]$mean,
                   "Log2CV"=resid.cv[[t]][[s]]$log2.cv,
                   "Resid_Log2CV"=resid.cv[[t]][[s]]$resid.log2.cv,
                   "Rank_Resid_Log2CV"=compute_local_ev_rank(resid.cv[[t]][[s]]$mean,
                                                             resid.cv[[t]][[s]]$resid.log2.cv,
                                                             win.size=100),
                   row.names=rownames(filtered.samples$log2.tmm.cpm.gf[[t]][[s]]))
      }
    })
  })
} else {
  resid.cv.summary <- lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
    data.frame("Mean"=resid.cv[[t]]$mean,
               "Log2CV"=resid.cv[[t]]$log2.cv,
               "Resid_Log2CV"=resid.cv[[t]]$resid.log2.cv,
               "Rank_Resid_Log2CV"=compute_local_ev_rank(resid.cv[[t]]$mean,
                                                         resid.cv[[t]]$resid.log2.cv,
                                                         win.size=100),
               row.names=rownames(filtered.samples$log2.tmm.cpm.gf[[t]]))
  })
}
```

### Residual MAD
```{r run-resid-mad}
if ("Sex" %in% colnames(filtered.samples$metadata)) {
  resid.mad <- lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
    lapply(setNames(levels(Sex), levels(Sex)), function(s) {
      # Check if tissue-sex condition is included
      if (ncol(filtered.samples$log2.tmm.cpm.gf[[t]][[s]])==0) { NULL } 
      else {
        residual_mad(filtered.samples$log2.tmm.cpm.gf[[t]][[s]])
      }
    })
  })
} else {
  resid.mad <- lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
    residual_mad(filtered.samples$log2.tmm.cpm.gf[[t]])
  })
}

# Bind into data frame
if ("Sex" %in% colnames(filtered.samples$metadata)) {
  resid.mad.summary <- lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
    lapply(setNames(levels(Sex), levels(Sex)), function(s) {
      # Check if tissue-sex condition is included
      if (ncol(filtered.samples$log2.tmm.cpm.gf[[t]][[s]])==0) { NULL }
      else {
        data.frame("Mean"=resid.mad[[t]][[s]]$mean,
                   "MAD"=resid.mad[[t]][[s]]$mad,
                   "Resid_MAD"=resid.mad[[t]][[s]]$resid.mad,
                   "Rank_Resid_MAD"=compute_local_ev_rank(resid.mad[[t]][[s]]$mean,
                                                          resid.mad[[t]][[s]]$resid.mad,
                                                          win.size=100),
                   row.names=rownames(filtered.samples$log2.tmm.cpm.gf[[t]][[s]]))
      }
    })
  })
} else {
  resid.mad.summary <- lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
    data.frame("Mean"=resid.mad[[t]]$mean,
               "MAD"=resid.mad[[t]]$mad,
               "Resid_MAD"=resid.mad[[t]]$resid.mad,
               "Rank_Resid_MAD"=compute_local_ev_rank(resid.mad[[t]]$mean,
                                                      resid.mad[[t]]$resid.mad,
                                                      win.size=100),
               row.names=rownames(filtered.samples$log2.tmm.cpm.gf[[t]]))
  })
}
```


## Bind results across conditions

### Adjusted SD
```{r}
if ("Sex" %in% colnames(filtered.samples$metadata)) {
  adj.sd.summary.bind <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        if (!is.null(adj.sd.summary[[t]][[s]])) {
          adj.sd.summary[[t]][[s]] %>%
            tibble::rownames_to_column(var="GeneID")
          }
    }) %>%
        dplyr::bind_rows(.id="Sex")        
  }) %>%
  dplyr::bind_rows(.id="Tissue")
} else {
  adj.sd.summary.bind <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      adj.sd.summary[[t]] %>%
        tibble::rownames_to_column(var="GeneID")
    }) %>%
    dplyr::bind_rows(.id="Tissue")
}

## Add gene biotype information
adj.sd.summary.bind$Biotype <- 
  vapply(adj.sd.summary.bind$GeneID, FUN.VALUE=character(1), function(g) {
    biotype$Gene.type[biotype$Gene.stable.ID==g]
  })

adj.sd.summary.bind <- adj.sd.summary.bind %>%
  dplyr::filter(Biotype %in% c("protein_coding", "lncRNA", "lincRNA"))

```

### Residual log2 (SD)
```{r}
if ("Sex" %in% colnames(filtered.samples$metadata)) {
  resid.sd.summary.bind <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        if (!is.null(resid.sd.summary[[t]][[s]])) {
          resid.sd.summary[[t]][[s]] %>%
            tibble::rownames_to_column(var="GeneID")
          }
    }) %>%
        dplyr::bind_rows(.id="Sex")        
  }) %>%
  dplyr::bind_rows(.id="Tissue")
} else {
  resid.sd.summary.bind <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      resid.sd.summary[[t]] %>%
        tibble::rownames_to_column(var="GeneID")
    }) %>%
    dplyr::bind_rows(.id="Tissue")
}

## Add gene biotype information
resid.sd.summary.bind$Biotype <- 
  vapply(resid.sd.summary.bind$GeneID, FUN.VALUE=character(1), function(g) {
    biotype$Gene.type[biotype$Gene.stable.ID==g]
  })

resid.sd.summary.bind <- resid.sd.summary.bind %>%
  dplyr::filter(Biotype %in% c("protein_coding", "lncRNA", "lincRNA"))

```

### Residual log2 (CV)^2
```{r}
if ("Sex" %in% colnames(filtered.samples$metadata)) {
  resid.cv.summary.bind <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        if (!is.null(resid.cv.summary[[t]][[s]])) {
          resid.cv.summary[[t]][[s]] %>%
            tibble::rownames_to_column(var="GeneID")
          }
    }) %>%
        dplyr::bind_rows(.id="Sex")        
  }) %>%
  dplyr::bind_rows(.id="Tissue")
} else {
  resid.cv.summary.bind <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      resid.cv.summary[[t]] %>%
        tibble::rownames_to_column(var="GeneID")
    }) %>%
    dplyr::bind_rows(.id="Tissue")
}

## Add gene biotype information
resid.cv.summary.bind$Biotype <- 
  vapply(resid.cv.summary.bind$GeneID, FUN.VALUE=character(1), function(g) {
    biotype$Gene.type[biotype$Gene.stable.ID==g]
  })

resid.cv.summary.bind <- resid.cv.summary.bind %>%
  dplyr::filter(Biotype %in% c("protein_coding", "lncRNA", "lincRNA"))

```

### Residual MAD
```{r}
if ("Sex" %in% colnames(filtered.samples$metadata)) {
  resid.mad.summary.bind <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        if (!is.null(resid.mad.summary[[t]][[s]])) {
          resid.mad.summary[[t]][[s]] %>%
            tibble::rownames_to_column(var="GeneID")
          }
    }) %>%
        dplyr::bind_rows(.id="Sex")        
  }) %>%
  dplyr::bind_rows(.id="Tissue")
} else {
  resid.mad.summary.bind <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      resid.mad.summary[[t]] %>%
        tibble::rownames_to_column(var="GeneID")
    }) %>%
    dplyr::bind_rows(.id="Tissue")
}

## Add gene biotype information
resid.mad.summary.bind$Biotype <- 
  vapply(resid.mad.summary.bind$GeneID, FUN.VALUE=character(1), function(g) {
    biotype$Gene.type[biotype$Gene.stable.ID==g]
  })

resid.mad.summary.bind <- resid.mad.summary.bind %>%
  dplyr::filter(Biotype %in% c("protein_coding", "lncRNA", "lincRNA"))

```

## All
```{r}
if ("Sex" %in% colnames(filtered.samples$metadata)) {
  var.summary.bind <-
    dplyr::full_join(adj.sd.summary.bind[,1:(ncol(adj.sd.summary.bind)-1)], 
                     resid.sd.summary.bind[,1:(ncol(resid.sd.summary.bind)-1)], by=c("Tissue","Sex","GeneID","Mean")) %>%
    dplyr::full_join(resid.cv.summary.bind[,1:(ncol(resid.cv.summary.bind)-1)], by=c("Tissue","Sex","GeneID","Mean")) %>%
    dplyr::full_join(resid.mad.summary.bind, by=c("Tissue","Sex","GeneID","Mean"))
  
} else {
  var.summary.bind <-
    dplyr::full_join(adj.sd.summary.bind[,1:(ncol(adj.sd.summary.bind)-1)], 
                     resid.sd.summary.bind[,1:(ncol(resid.sd.summary.bind)-1)], by=c("Tissue","GeneID","Mean")) %>%
    dplyr::full_join(resid.cv.summary.bind[,1:(ncol(resid.cv.summary.bind)-1)], by=c("Tissue","GeneID","Mean")) %>%
    dplyr::full_join(resid.mad.summary.bind, by=c("Tissue","GeneID","Mean"))
  
}

```

# Save
```{r save-objects}
if (!params$combat) {
  if (!params$qc1.only) {
    output.path <- file.path(params$output.rds.path, "no_combat/full_qc")    
  } else {
    output.path <- file.path(params$output.rds.path, "no_combat/qc1_only")
  }
  
} else {
  if (!params$qc1.only) {
    output.path <- file.path(params$output.rds.path, "combat/full_qc")
  } else {
    output.path <- file.path(params$output.rds.path, "combat/qc1_only")    
  }
}

if (!dir.exists(output.path)) { dir.create(output.path, recursive=TRUE, showWarnings=FALSE) }

if (!params$all.nonzero.matrix) { 
    # Without jackknife
    saveRDS(adj.sd.summary,
            file.path(output.path, 
                      paste(params$species, "adj.sd.summary.rds", sep="_")))
  
    saveRDS(resid.sd.summary,
            file.path(output.path, 
                      paste(params$species, "resid.sd.summary.rds", sep="_")))
    
    saveRDS(resid.cv.summary,
            file.path(output.path, 
                      paste(params$species, "resid.cv.summary.rds", sep="_")))
    
    saveRDS(resid.mad.summary,
            file.path(output.path, 
                      paste(params$species, "resid.mad.summary.rds", sep="_")))
    
    saveRDS(var.summary.bind,
            file.path(output.path, 
                      paste(params$species, "var.summary.bind.rds", sep="_")))    
  
} else {
    # Without jackknife  
    saveRDS(adj.sd.summary,
            file.path(output.path, 
                      paste(params$species, "adj.sd.summary.nonzero.rds", sep="_")))
  
    saveRDS(resid.sd.summary,
            file.path(output.path, 
                      paste(params$species, "resid.sd.summary.nonzero.rds", sep="_")))
    
    saveRDS(resid.cv.summary,
            file.path(output.path, 
                      paste(params$species, "resid.cv.summary.nonzero.rds", sep="_")))
    
    saveRDS(resid.mad.summary,
            file.path(output.path, 
                      paste(params$species, "resid.mad.summary.nonzero.rds", sep="_")))
    
    saveRDS(var.summary.bind,
            file.path(output.path, 
                      paste(params$species, "var.summary.bind.nonzero.rds", sep="_")))    

}
```