---
title: "Variability Plots (*`r params$species.name`*)"
author: "Bucao, Christabel Floi"
output: html_document
date: "`r Sys.Date()`"
params:
  gene.metadata.path: "../../data/gene_metadata"
  input.samples.rds.path: "../../results/01_Filter_Samples"
  input.variability.rds.path: "../../results/02_Variability_Jackknife"
  output.rds.path: "../../results/03_Variability_Plots"
  species: "DRE"
  species.name: "Danio rerio"
  min.percentile: 0.00
  max.percentile: 0.95
  all.nonzero.matrix: FALSE
---

# Setup {.tabset}

## Parameters
```{r print-params}
print(params)
```

## Libraries
```{r import-libraries}
library(dplyr)
```

## Functions
```{r source-functions}
source("../functions/helper_functions.R")
source("../functions/variability_jackknife.R")
source("../functions/variability_plots.R")
```

## Session Info
```{r print-session-info}
sessionInfo()
```

## Data

### Metadata
```{r import-metadata}
biotype <- 
  read.delim(file.path(params$gene.metadata.path, 
                       paste(params$species, "biotype_Ensembl_105.txt", sep="_")), 
             sep="\t")
```

### Processed data
```{r import-filtered-samples}
# Load saved data from 01_Filter_Samples.Rmd
# Normalized by condition
if (!params$all.nonzero.matrix) {
  filtered.samples <- readRDS(
    file.path(params$input.samples.rds.path,
              paste(params$species, "4_reps_data.conditions.rds", sep="_")))
} else {
  filtered.samples <- readRDS(
    file.path(params$input.samples.rds.path,
              paste(params$species, "4_reps_data.conditions.nonzero.rds", sep="_")))  
}

Tissue <- as.factor(filtered.samples$metadata$Tissue)
if ("Sex" %in% colnames(filtered.samples$metadata)) { 
  Sex <- as.factor(filtered.samples$metadata$Sex) 
}

# Load jackknifed expression variability estimates from 02_Variability_Jackknife.Rmd
## Adjusted SD
if (!params$all.nonzero.matrix) {
  jack.adj.sd <- readRDS(
    file.path(params$input.variability.rds.path,
              paste(params$species, "jack.adj.sd.rds", sep="_")))

  ## Residual SD
  jack.resid.sd <- readRDS(
    file.path(params$input.variability.rds.path,
              paste(params$species, "jack.resid.sd.rds", sep="_")))

  ## Residual CV / Local (residual) CV
  jack.resid.cv <- readRDS(
    file.path(params$input.variability.rds.path,
              paste(params$species, "jack.resid.cv.rds", sep="_")))  
} else {
  jack.adj.sd <- readRDS(
    file.path(params$input.variability.rds.path,
              paste(params$species, "jack.adj.sd.nonzero.rds", sep="_")))

  ## Residual SD
  jack.resid.sd <- readRDS(
    file.path(params$input.variability.rds.path,
              paste(params$species, "jack.resid.sd.nonzero.rds", sep="_")))

  ## Residual CV / Local (residual) CV
  jack.resid.cv <- readRDS(
    file.path(params$input.variability.rds.path,
              paste(params$species, "jack.resid.cv.nonzero.rds", sep="_")))   
}
```

# Main {.tabset}

## Initial summary statistics
```{r compute-summary-statistics}
if ("Sex" %in% colnames(filtered.samples$metadata)) {
  summary.stats <- 
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        compute_gene_summary_stats(filtered.samples$log2.tmm.cpm.gf[[t]][[s]])
    })
  })
  
  # Remove top and bottom x% of genes by expression level
  summary.stats <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        summary.stats[[t]][[s]][
          summary.stats[[t]][[s]]$Rank_Mean>params$min.percentile &
            summary.stats[[t]][[s]]$Rank_Mean<params$max.percentile,]
      })
    })
  
  # Remove missing conditions
  summary.stats <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        summary.stats[[t]][[s]][complete.cases(summary.stats[[t]][[s]]),]
      })
    })
  
} else {
  summary.stats <- 
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      compute_gene_summary_stats(filtered.samples$log2.tmm.cpm.gf[[t]])
    })
  
  # Remove top and bottom x% of genes by expression level
  summary.stats <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
        summary.stats[[t]][
          summary.stats[[t]]$Rank_Mean>params$min.percentile &
            summary.stats[[t]]$Rank_Mean<params$max.percentile,]
    })
}
```

### Plots {.tabset}

#### SD x Mean
```{r plot-sd-vs-mean}
plot_sd_vs_mean(summary.stats, filtered.samples$metadata)
```

#### Log2 SD x Mean
```{r plot-log2sd-vs-mean}
plot_sd_vs_mean(summary.stats, filtered.samples$metadata, log2=TRUE)
```

#### CV vs Mean
```{r plot-cv2-vs-mean}
plot_cv_vs_mean(summary.stats, filtered.samples$metadata, log2=FALSE)
```

#### Log2 (CV^2) vs Mean
```{r plot-log2-cv2-vs-mean}
plot_cv_vs_mean(summary.stats, filtered.samples$metadata, log2=TRUE)
```

## Jackknife

### Summarized results
```{r retrieve-summarized-jackknife-results}
# Adjusted SD
if ("Sex" %in% colnames(filtered.samples$metadata)) { 
  jack.adj.sd.summary <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        if (!is.null(jack.adj.sd[[t]][[s]])) { 
          jack.adj.sd[[t]][[s]]$summary %>%
            tibble::rownames_to_column(., var="GeneID")
        }
      })
    })
} else { 
  jack.adj.sd.summary <- 
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      jack.adj.sd[[t]]$summary %>%
        tibble::rownames_to_column(., var="GeneID")
      })
}

# Residual SD
if ("Sex" %in% colnames(filtered.samples$metadata)) { 
  jack.resid.sd.summary <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        if (!is.null(jack.resid.sd[[t]][[s]])) { 
          jack.resid.sd[[t]][[s]]$summary %>%
            tibble::rownames_to_column(., var="GeneID")
        }
      })
    })
} else { 
  jack.resid.sd.summary <- 
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      jack.resid.sd[[t]]$summary %>%
        tibble::rownames_to_column(., var="GeneID")
      })
}

# Residual CV
if ("Sex" %in% colnames(filtered.samples$metadata)) { 
  jack.resid.cv.summary <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        if (!is.null(jack.resid.cv[[t]][[s]])) { 
          jack.resid.cv[[t]][[s]]$summary %>%
            tibble::rownames_to_column(., var="GeneID")
        }
      })
    })
} else { 
  jack.resid.cv.summary <- 
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      jack.resid.cv[[t]]$summary %>%
        tibble::rownames_to_column(., var="GeneID")
      })
}
```

### Plots {.tabset} 

#### Log2 SD x Mean
```{r plot-jackknife-log2sd-vs-mean}
plot_jackknife_log2sd_vs_mean(jack.resid.sd.summary, filtered.samples$metadata)
```

#### Log2 CV x Mean
```{r plot-jackknife-log2cv-vs-mean}
plot_jackknife_log2cv_vs_mean(jack.resid.cv.summary, filtered.samples$metadata)
```


#### Adjusted SD x Mean
```{r plot-jackknife-adj-sd-vs-mean}
plot_jackknife_adj_sd_vs_mean(jack.adj.sd.summary, filtered.samples$metadata)
```

#### Residual log2 SD x Mean
```{r plot-jackknife-residual-log2sd-vs-mean}
plot_jackknife_resid_var_vs_mean(jack.resid.sd.summary, filtered.samples$metadata, method="sd")
```

#### Residual log2 CV x Mean
```{r plot-jackknife-residual-log2cv-vs-mean}
plot_jackknife_resid_var_vs_mean(jack.resid.cv.summary, filtered.samples$metadata, method="cv")
```

#### Global variability rank - based on residual log2 CV
```{r plot-jackknife-global-rank-log2cv-vs-mean}
# Based on percentile rank of residual log2(CV^2) across the range of expression levels
plot_jackknife_global_var_rank_vs_mean(jack.resid.cv.summary, filtered.samples$metadata, method="cv")
```

#### Local variability rank - based on residual log2 CV
```{r plot-jackknife-local-rank-log2cv-vs-mean}
# Based on mean percentile rank of residual log2(CV^2) within sliding windows of genes with similar expression levels
plot_jackknife_local_var_rank_vs_mean(jack.resid.cv.summary, filtered.samples$metadata, method="cv")
```

## Protein-coding vs lncRNA

### Split by gene biotype
```{r split-by-biotype}
jack.resid.cv.biotype <- list()

if ("Sex" %in% colnames(filtered.samples$metadata)) {
  jack.resid.cv.biotype$coding <- 
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        jack.resid.cv[[t]][[s]]$summary[
          rownames(jack.resid.cv[[t]][[s]]$summary) %in%
            biotype$Gene.stable.ID[biotype$Gene.type=="protein_coding"],]
    })
  })
  
  jack.resid.cv.biotype$lncrna <- 
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        jack.resid.cv[[t]][[s]]$summary[
          rownames(jack.resid.cv[[t]][[s]]$summary) %in% 
            biotype$Gene.stable.ID[biotype$Gene.type %in% c("lncRNA", "lincRNA")],]
    })
  })  
} else {
  jack.resid.cv.biotype$coding <- 
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      jack.resid.cv[[t]]$summary[
        rownames(jack.resid.cv[[t]]$summary) %in%
          biotype$Gene.stable.ID[biotype$Gene.type=="protein_coding"],]
  })    

  jack.resid.cv.biotype$lncrna <- 
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      jack.resid.cv[[t]]$summary[
        rownames(jack.resid.cv[[t]]$summary) %in%
          biotype$Gene.stable.ID[biotype$Gene.type %in% c("lncRNA", "lincRNA")],]
  })   
}
```

### Bind jackknife results across all conditions
```{r bind-summarized-jackknife-results}
if ("Sex" %in% colnames(filtered.samples$metadata)) { 
  jack.resid.cv.bind <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        if (!is.null(jack.resid.cv[[t]][[s]])) { 
          jack.resid.cv[[t]][[s]]$summary %>%
            tibble::rownames_to_column(., var="GeneID")
          }
        }) %>%
        dplyr::bind_rows(.id="Sex")
      }) %>%
    dplyr::bind_rows(.id="Tissue")
} else { 
  jack.resid.cv.bind <- 
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      jack.resid.cv[[t]]$summary %>%
        tibble::rownames_to_column(., var="GeneID")
      }) %>%
    dplyr::bind_rows(.id="Tissue")
} 

jack.resid.cv.bind <- jack.resid.cv.bind %>%
  dplyr::inner_join(biotype[,-1], by=c("GeneID"="Gene.stable.ID"))
colnames(jack.resid.cv.bind)[ncol(jack.resid.cv.bind)] <- "Biotype"

jack.resid.cv.bind <- 
  jack.resid.cv.bind[jack.resid.cv.bind$Biotype %in% c("protein_coding", "lncRNA", "lincRNA"),]
```

### Compare variability rank distribution of protein-coding genes and lncRNA
```{r plot-ev-boxplot-lncrna-vs-coding, fig.width = 10, fig.height = 8}
if ("Sex" %in% colnames(filtered.samples$metadata)) {
  boxplot_var_rank_lncrna_vs_coding(jack.resid.cv.bind, sex=TRUE, method="cv")  
} else {
  boxplot_var_rank_lncrna_vs_coding(jack.resid.cv.bind, sex=FALSE, method="cv")
}
```

# Save
```{r save-objects}
if (!dir.exists(params$output.rds.path)) { dir.create(params$output.rds.path, showWarnings=FALSE) }

if (!params$all.nonzero.matrix) {
  saveRDS(jack.resid.cv.bind,
          file.path(params$output.rds.path, 
                    paste(params$species, "jack.resid.cv.bind.rds", sep="_")))
  
  saveRDS(jack.resid.cv.summary,
          file.path(params$output.rds.path, 
                    paste(params$species, "jack.resid.cv.summary.rds", sep="_")))
  
  saveRDS(jack.resid.cv.biotype,
          file.path(params$output.rds.path, 
                    paste(params$species, "jack.resid.cv.biotype.rds", sep="_")))  
} else {
  saveRDS(jack.resid.cv.bind,
          file.path(params$output.rds.path, 
                    paste(params$species, "jack.resid.cv.bind.nonzero.rds", sep="_")))
  
  saveRDS(jack.resid.cv.summary,
          file.path(params$output.rds.path, 
                    paste(params$species, "jack.resid.cv.summary.nonzero.rds", sep="_")))
  
  saveRDS(jack.resid.cv.biotype,
          file.path(params$output.rds.path, 
                    paste(params$species, "jack.resid.cv.biotype.nonzero.rds", sep="_")))   
}
```
