---
title: "Variability rank correlation across conditions (*`r params$species.name`*)"
author: "Bucao, Christabel Floi"
output: html_document
date: "`r Sys.Date()`"
params:
  input.variability.rds.path: "../../results/03_Variability_Plots"
  output.rds.path: "../../results/04_Correlation_Across_Conditions"
  species: "DRE"
  species.name: "Danio rerio"
  all.nonzero.matrix: FALSE
---

# Setup {.tabset}

## Parameters
```{r print-params}
print(params)
```

## Libraries
```{r import-libraries}
# Install reshape2
packages <- c("tidyr","dplyr","tibble","reshape2","ggplot2","viridis")

for (p in packages) {
  if (!require(p, character.only=TRUE)) {
    print(paste("Installing", p))
    install.packages(p, character.only=TRUE)
    renv::snapshot()
  }
  print(paste("Loading", p))
  library(p, character.only=TRUE)
}
```

## Functions
```{r source-functions}
source("../functions/correlation_across_conditions_plots.R")
```

## Session Info
```{r print-session-info}
sessionInfo()
```

## Data

### Expression variability
```{r import--variability}
# Load processed expression variability estimates from 03_Variability_Plots.Rmd
if (!params$all.nonzero.matrix) {
  jack.bind<- readRDS(
    file.path(params$input.variability.rds.path,
              paste(params$species, "jack.resid.cv.bind.rds", sep="_")))
} else {
  jack.bind<- readRDS(
    file.path(params$input.variability.rds.path,
              paste(params$species, "jack.resid.cv.bind.nonzero.rds", sep="_"))) 
}
```

# Main {.tabset}

## Mean expression level

### Expression matrix without missing values
```{r generate-expr-matrix}
# All values are estimates generated from jackknife (n-1) resampling
if ("Sex" %in% colnames(jack.bind)) {
  expr.matrix <- jack.bind[,c("Tissue","Sex","GeneID","Mean_Mean","Biotype")] %>%
    dplyr::filter(Biotype=="protein_coding")
  
  expr.matrix$Condition <- paste(expr.matrix$Tissue, expr.matrix$Sex, sep="_")
  
  expr.matrix <- expr.matrix[,c("Condition","GeneID","Mean_Mean")] %>%
    tidyr::pivot_wider(names_from=Condition, values_from="Mean_Mean") %>%
    na.omit(.) %>%
    tibble::column_to_rownames(var="GeneID")
  
} else {
  expr.matrix <- jack.bind[,c("Tissue","GeneID","Mean_Mean","Biotype")] %>%
    dplyr::filter(Biotype=="protein_coding")
  
  expr.matrix$Condition <- expr.matrix$Tissue
  
  expr.matrix <- expr.matrix[,c("Condition","GeneID","Mean_Mean")] %>%
    tidyr::pivot_wider(names_from=Condition, values_from="Mean_Mean") %>%
    na.omit(.) %>%
    tibble::column_to_rownames(var="GeneID")  
}
```

```{r}
# Number of protein-coding genes included
print(nrow(expr.matrix))
```

### Correlation matrix
```{r compute-expr-corr-matrix}
expr.corr.matrix <- cor(expr.matrix, method="spearman")

expr.corr.matrix.diag <- expr.corr.matrix
expr.corr.matrix.diag[upper.tri(expr.corr.matrix.diag, diag=TRUE)] <- NA

expr.corr.matrix.melt <- melt(expr.corr.matrix.diag, na.rm=TRUE)
colnames(expr.corr.matrix.melt) <- c("Condition1", "Condition2", "Spearman")
```

### Plot - Heatmap
```{r plot-expr-corr-heatmap, fig.height=10, fig.width=14}
plot_triangle_correlation_matrix(expr.corr.matrix.melt, title="Expression level")
```


## Log2 CV

### Log2 CV matrix without missing values
```{r generate-log2cv-matrix}
# All values are estimates generated from jackknife (n-1) resampling
if ("Sex" %in% colnames(jack.bind)) {
  log2.cv.matrix <- jack.bind[,c("Tissue","Sex","GeneID","Mean_Log2CV","Biotype")] %>%
    dplyr::filter(Biotype=="protein_coding")
  
  log2.cv.matrix$Condition <- paste(log2.cv.matrix$Tissue, log2.cv.matrix$Sex, sep="_")
  
  log2.cv.matrix <- log2.cv.matrix[,c("Condition","GeneID","Mean_Log2CV")] %>%
    tidyr::pivot_wider(names_from=Condition, values_from="Mean_Log2CV") %>%
    na.omit(.) %>%
    tibble::column_to_rownames(var="GeneID")
  
} else {
  log2.cv.matrix <- jack.bind[,c("Tissue","GeneID","Mean_Log2CV","Biotype")] %>%
    dplyr::filter(Biotype=="protein_coding")
  
  log2.cv.matrix$Condition <- log2.cv.matrix$Tissue
  
  log2.cv.matrix <- log2.cv.matrix[,c("Condition","GeneID","Mean_Log2CV")] %>%
    tidyr::pivot_wider(names_from=Condition, values_from="Mean_Log2CV") %>%
    na.omit(.) %>%
    tibble::column_to_rownames(var="GeneID")  
}
```

### Correlation matrix
```{r compute-log2cv-corr-matrix}
log2.cv.corr.matrix <- cor(log2.cv.matrix, method="spearman")

log2.cv.corr.matrix.diag <- log2.cv.corr.matrix
log2.cv.corr.matrix.diag[upper.tri(log2.cv.corr.matrix.diag, diag=TRUE)] <- NA

log2.cv.corr.matrix.melt <- melt(log2.cv.corr.matrix.diag, na.rm=TRUE)
colnames(log2.cv.corr.matrix.melt) <- c("Condition1", "Condition2", "Spearman")
```

### Plot - Heatmap
```{r plot-log2cv-corr-heatmap, fig.height=10, fig.width=14}
plot_triangle_correlation_matrix(log2.cv.corr.matrix.melt, title="Log2 CV")
```

## Residual log2 CV

### Residual log2 CV matrix without missing values
```{r generate-resid-log2cv-matrix}
# All values are estimates generated from jackknife (n-1) resampling
if ("Sex" %in% colnames(jack.bind)) {
  resid.log2.cv.matrix <- jack.bind[,c("Tissue","Sex","GeneID","Mean_Resid_Log2CV","Biotype")] %>%
    dplyr::filter(Biotype=="protein_coding")
  
  resid.log2.cv.matrix$Condition <- paste(resid.log2.cv.matrix$Tissue, resid.log2.cv.matrix$Sex, sep="_")
  
  resid.log2.cv.matrix <- resid.log2.cv.matrix[,c("Condition","GeneID","Mean_Resid_Log2CV")] %>%
    tidyr::pivot_wider(names_from=Condition, values_from="Mean_Resid_Log2CV") %>%
    na.omit(.) %>%
    tibble::column_to_rownames(var="GeneID")
} else {
  resid.log2.cv.matrix <- jack.bind[,c("Tissue","GeneID","Mean_Resid_Log2CV","Biotype")] %>%
    dplyr::filter(Biotype=="protein_coding")
  
  resid.log2.cv.matrix$Condition <- resid.log2.cv.matrix$Tissue
  
  resid.log2.cv.matrix <- resid.log2.cv.matrix[,c("Condition","GeneID","Mean_Resid_Log2CV")] %>%
    tidyr::pivot_wider(names_from=Condition, values_from="Mean_Resid_Log2CV") %>%
    na.omit(.) %>%
    tibble::column_to_rownames(var="GeneID")  
}
```

### Correlation matrix
```{r compute-resid-log2cv-corr-matrix}
resid.log2.cv.corr.matrix <- cor(resid.log2.cv.matrix, method="spearman")

resid.log2.cv.corr.matrix.diag <- resid.log2.cv.corr.matrix
resid.log2.cv.corr.matrix.diag[upper.tri(resid.log2.cv.corr.matrix.diag, diag=TRUE)] <- NA

resid.log2.cv.corr.matrix.melt <- melt(resid.log2.cv.corr.matrix.diag, na.rm=TRUE)
colnames(resid.log2.cv.corr.matrix.melt) <- c("Condition1", "Condition2", "Spearman")
```

### Plot - Heatmap
```{r plot-resid-log2cv-corr-heatmap, fig.height=10, fig.width=14}
plot_triangle_correlation_matrix(resid.log2.cv.corr.matrix.melt, title="Residual log2 CV")
```

## Variability rank

### Variability rank matrix without missing values
```{r generate-var-rank-matrix}
# All values are estimates generated from jackknife (n-1) resampling
if ("Sex" %in% colnames(jack.bind)) {
  var.rank.matrix <- jack.bind[,c("Tissue","Sex","GeneID","Mean_Local_Rank_Log2CV","Biotype")] %>%
    dplyr::filter(Biotype=="protein_coding")
  
  var.rank.matrix$Condition <- paste(var.rank.matrix$Tissue, var.rank.matrix$Sex, sep="_")
  
  var.rank.matrix <- var.rank.matrix[,c("Condition","GeneID","Mean_Local_Rank_Log2CV")] %>%
    tidyr::pivot_wider(names_from=Condition, values_from="Mean_Local_Rank_Log2CV") %>%
    na.omit(.) %>%
    tibble::column_to_rownames(var="GeneID")
} else {
  var.rank.matrix <- jack.bind[,c("Tissue","GeneID","Mean_Local_Rank_Log2CV","Biotype")] %>%
    dplyr::filter(Biotype=="protein_coding")
  
  var.rank.matrix$Condition <- var.rank.matrix$Tissue
  
  var.rank.matrix <- var.rank.matrix[,c("Condition","GeneID","Mean_Local_Rank_Log2CV")] %>%
    tidyr::pivot_wider(names_from=Condition, values_from="Mean_Local_Rank_Log2CV") %>%
    na.omit(.) %>%
    tibble::column_to_rownames(var="GeneID")  
}
```

### Correlation matrix
```{r compute-var-rank-corr-matrix}
var.rank.corr.matrix <- cor(var.rank.matrix, method="spearman")

var.rank.corr.matrix.diag <- var.rank.corr.matrix
var.rank.corr.matrix.diag[upper.tri(var.rank.corr.matrix.diag, diag=TRUE)] <- NA

var.rank.corr.matrix.melt <- melt(var.rank.corr.matrix.diag, na.rm=TRUE)
colnames(var.rank.corr.matrix.melt) <- c("Condition1", "Condition2", "Spearman")
```

### Plot - Heatmap
```{r plot-var-rank-corr-heatmap, fig.height=10, fig.width=14}
plot_triangle_correlation_matrix(var.rank.corr.matrix.melt, title="Variability rank")
```

# Save
```{r save-objects}
if (!dir.exists(params$output.rds.path)) { dir.create(params$output.rds.path, showWarnings=FALSE) }

if (!params$all.nonzero.matrix) {
  saveRDS(expr.corr.matrix.melt,
          file.path(params$output.rds.path, 
                    paste(params$species, "expr.corr.matrix.melt.rds", sep="_")))
  
  saveRDS(log2.cv.corr.matrix.melt,
          file.path(params$output.rds.path, 
                    paste(params$species, "log2.cv.corr.matrix.melt.rds", sep="_")))
  
  saveRDS(resid.log2.cv.corr.matrix.melt,
          file.path(params$output.rds.path, 
                    paste(params$species, "resid.log2.cv.corr.matrix.melt.rds", sep="_")))

  saveRDS(var.rank.corr.matrix.melt,
          file.path(params$output.rds.path, 
                    paste(params$species, "var.rank.corr.matrix.melt.rds", sep="_")))

} else {
  saveRDS(expr.corr.matrix.melt,
          file.path(params$output.rds.path, 
                    paste(params$species, "expr.corr.matrix.melt.nonzero.rds", sep="_")))
  
  saveRDS(log2.cv.corr.matrix.melt,
          file.path(params$output.rds.path, 
                    paste(params$species, "log2.cv.corr.matrix.melt.nonzero.rds", sep="_")))
  
  saveRDS(resid.log2.cv.corr.matrix.melt,
          file.path(params$output.rds.path, 
                    paste(params$species, "resid.log2.cv.corr.matrix.melt.nonzero.rds", sep="_")))

  saveRDS(var.rank.corr.matrix.melt,
          file.path(params$output.rds.path, 
                    paste(params$species, "var.rank.corr.matrix.melt.nonzero.rds", sep="_")))
}
```