---
title: "Bimodality Test (*`r params$species.name`*)"
author: "Bucao, Christabel Floi"
output: html_document
date: "`r Sys.Date()`"
params:
  input.samples.rds.path: "../../results/01_Filter_Samples"
  input.variability.rds.path: "../../results/03_Variability_Plots"
  output.rds.path: "../../results/05_Bimodality_Test"
  species: "DRE"
  species.name: "Danio rerio"
  min.replicates: 10
---

This notebook runs the bimodality test for organ-sex conditions with at least (`r params$min.replicates`) replicates and summarizes the results.

# Setup {.tabset}

## Parameters
```{r print-params}
print(params)
```

## Libraries
```{r import-libraries}
# Install BimodalIndex
packages <- c("dplyr","tibble","matrixStats","BimodalIndex","ggplot2","ggpubr")

for (p in packages) {
  if (!require(p, character.only=TRUE)) {
    print(paste("Installing", p))
    install.packages(p, character.only=TRUE)
    renv::snapshot()
  }
  print(paste("Loading", p))
  library(p, character.only=TRUE)
}
```

## Functions
```{r source-functions}
source("../functions/bimodality_test.R")
source("../functions/bimodality_test_plots.R")
```

## Session Info
```{r print-session-info}
sessionInfo()
```

## Data

### Expression levels
```{r import-expression-levels}
# Load saved data from 01_Filter_Samples.Rmd
# Normalized within-condition
filtered.samples <- readRDS(
  file.path(params$input.samples.rds.path,
            paste(params$species, "4_reps_data.conditions.rds", sep="_")))
```

### Expression variability
```{r import-variability}
# Load processed expression variability estimates from 03_Variability_Plots.Rmd
jack.bind <- readRDS(
  file.path(params$input.variability.rds.path,
            paste(params$species, "jack.resid.cv.bind.rds", sep="_")))
```

# Main {.tabset}

## Data processing

### Select conditions with minimum replicates
```{r select-conditions}
filtered.samples.retained <- list()

filtered.samples.retained$conditions <- filtered.samples$conditions %>%
  dplyr::filter(n>=params$min.replicates)

filtered.samples.retained$metadata <- filtered.samples$metadata %>%
  dplyr::filter(grepl(paste(filtered.samples.retained$conditions$regex,collapse="|"),.$SampleName))

if ("Sex" %in% colnames(filtered.samples.retained$conditions)) {
  conditions <- 
    paste(filtered.samples.retained$conditions$Tissue, 
          filtered.samples.retained$conditions$Sex, sep="_")
} else {
  conditions <- filtered.samples.retained$conditions$Tissue
}

filtered.samples.retained$log2.tmm.cpm.gf <- vector(mode="list", length=length(conditions))
names(filtered.samples.retained$log2.tmm.cpm.gf) <- conditions

jack.condition.retained <- vector(mode="list", length=length(conditions))
names(jack.condition.retained) <- conditions

for (c in 1:length(conditions)) {
  if ("Sex" %in% colnames(filtered.samples.retained$conditions)) {
    t <- filtered.samples.retained$conditions$Tissue[[c]]
    s <- filtered.samples.retained$conditions$Sex[[c]]
    
    ts <- paste(t,s,sep="_")
    
    filtered.samples.retained$log2.tmm.cpm.gf[[ts]] <- filtered.samples$log2.tmm.cpm.gf[[t]][[s]]
    
    jack.condition.retained[[ts]] <- jack.bind %>%
      dplyr::filter(Biotype=="protein_coding") %>%
      dplyr::filter(paste(.$Tissue, .$Sex, sep="_")==ts)
    
  } else {
    ts <- filtered.samples.retained$conditions$Tissue[[c]]
    
    filtered.samples.retained$log2.tmm.cpm.gf[[ts]] <- filtered.samples$log2.tmm.cpm.gf[[ts]]
    
    jack.condition.retained[[ts]] <- jack.bind %>%
      dplyr::filter(Biotype=="protein_coding") %>%
      dplyr::filter(Tissue==ts)
  }
  
  # Filter gene list
  filtered.samples.retained$log2.tmm.cpm.gf[[ts]] <- filtered.samples.retained$log2.tmm.cpm.gf[[ts]] %>%
    .[rownames(.) %in% jack.condition.retained[[ts]]$GeneID,]
}
```

### Compute z-score per condition
```{r compute-zscore}
zscore.matrix <- lapply(setNames(conditions, conditions), function(ts) {
  zscore(filtered.samples.retained$log2.tmm.cpm.gf[[ts]])
})
```

### Compute bimodality test per condition
```{r compute-bimodality-index}
bimodality.test <- lapply(setNames(conditions, conditions), function(ts) {
  print(ts)
  bimodalIndex(zscore.matrix[[ts]]) %>%
    tibble::rownames_to_column(var="GeneID")
})
```

### Bin variability ranks
```{r bin-var-ranks}
for (ts in conditions) {
  bimodality.test[[ts]]$VarRank <-
    vapply(bimodality.test[[ts]]$GeneID, FUN.VALUE=numeric(1), FUN=function(g) {
      jack.condition.retained[[ts]]$Mean_Local_Rank_Log2CV[jack.condition.retained[[ts]]$GeneID==g]
    })
  
  # Group variability ranks into bins of size 0.1
  bimodality.test[[ts]]$VarRankBin <- round(bimodality.test[[ts]]$VarRank, 1)
}
```

## Plots

### Boxplots - Bimodality index (BI) by variability rank

```{r bind-df}
# Bind data frames for plotting
bimodality.test.bind <- dplyr::bind_rows(bimodality.test, .id="Condition")
```

#### Landscape
```{r boxplot-bimodality-index-landscape, fig.width=28, fig.height=18}
boxplot_bimodality_by_variability(bimodality.test.bind, title=params$species.name, facet=TRUE)
```

#### Portrait
```{r boxplot-bimodality-index-portrait, fig.width=18, fig.height=22}
boxplot_bimodality_by_variability(bimodality.test.bind, title=params$species.name, facet=TRUE)
```

### Expression level z-scores by variability rank bin

**Question:** At which variability rank does the z-score distribution visually appear bimodal?

#### All
```{r fig.width=8, fig.height=5}
# Will only print last plot to HTML, but works on RStudio
for (ts in conditions) {
  plot_density_zscore(zscore.matrix[[ts]], bimodality.test[[ts]], title=ts)
}
```

#### Example: Brain
```{r plot-density-zscore-var-rank-brain, fig.width=8, fig.height=5}
if ("brain" %in% filtered.samples.retained$conditions$Tissue) {
  if ("Sex" %in% colnames(filtered.samples.retained$conditions)) {
    ts <- "brain_F"
  } else { ts <- "brain" }

  plot_density_zscore(zscore.matrix[[ts]], bimodality.test[[ts]], title=ts)  
}

```

#### Example: Gonads
```{r plot-density-zscore-var-rank-gonads, fig.width=8, fig.height=5}
if ("gonads" %in% filtered.samples.retained$condtions$Tissue) {
  if ("Sex" %in% colnames(filtered.samples.retained$conditions)) {
    ts <- "gonads_F"
  } else { ts <- "gonads" }

  plot_density_zscore(zscore.matrix[[ts]], bimodality.test[[ts]], title=ts)  
}
```


# Save
```{r save-objects}
if (!dir.exists(params$output.rds.path)) { dir.create(params$output.rds.path, showWarnings=FALSE) }

saveRDS(filtered.samples.retained,
        file.path(params$output.rds.path, 
                  paste(params$species, "filtered.samples.retained.rds", sep="_")))

saveRDS(jack.condition.retained,
        file.path(params$output.rds.path, 
                  paste(params$species, "jack.condition.retained.rds", sep="_")))

saveRDS(zscore.matrix,
        file.path(params$output.rds.path, 
                  paste(params$species, "zscore.matrix.rds", sep="_")))

saveRDS(bimodality.test,
        file.path(params$output.rds.path, 
                  paste(params$species, "bimodality.test.rds", sep="_")))

saveRDS(bimodality.test.bind,
        file.path(params$output.rds.path, 
                  paste(params$species, "bimodality.test.bind.rds", sep="_")))
```

