---
title: "Simulated Bimodality (*`r params$species.name`*)"
author: "Bucao, Christabel Floi"
output: html_document
date: "`r Sys.Date()`"
params:
  gene.metadata.path: "../../data/gene_metadata"
  input.samples.rds.path: "../../results/01_Filter_Samples"
  input.variability.rds.path: "../../results/03_Variability_Plots"
  output.rds.path: "../../results/06_Simulated_Bimodality"
  species: "ELU"
  species.name: "Esox lucius"
  sex: "F"
  total.replicates: 10
  tissue1: "brain"
  tissue2: "gonads"
  set.seed: 12345
  min.cpm: 1.0
  min.percentile: 0.00
  max.percentile: 0.95
  win.size: 100
---

# Setup {.tabset}

## Parameters
```{r print-params}
print(params)
```

## Libraries
```{r import-libraries}
library(dplyr)
library(tibble)
library(ggplot2)
library(ggpubr)
library(BimodalIndex)
```

## Functions
```{r source-functions}
source("../functions/filter_samples.R")
source("../functions/variability_jackknife.R")
source("../functions/bimodality_test.R")
source("../functions/bimodality_test_plots.R")
#source("../functions/annotate_boxplots.R")
```

## Session Info
```{r print-session-info}
sessionInfo()
```

## Data

### Gene biotype
```{r import-biotype}
biotype <- 
  read.delim(file.path(params$gene.metadata.path, 
                       paste(params$species, "biotype_Ensembl_105.txt", sep="_")), 
             sep="\t")
```

### Expression levels
```{r import-expression-levels}
# Load saved data from 01_Filter_Samples.Rmd
## Normalized separately per condition
samples.norm.within <- readRDS(
  file.path(params$input.samples.rds.path,
            paste(params$species, "4_reps_data.conditions.rds", sep="_")))

## Normalized across conditions
samples.norm.across <- readRDS(
  file.path(params$input.samples.rds.path,
            paste(params$species, "4_reps_data.qc2.cor.rds", sep="_")))
```

### Expression variability
```{r import-variability}
# Load processed expression variability estimates from 03_Variability_Plots.Rmd
jack.bind <- readRDS(
  file.path(params$input.variability.rds.path,
            paste(params$species, "jack.resid.cv.bind.rds", sep="_")))
```

# Main {.tabset}

## Observed bimodality

### Filter samples and genes
```{r filter-samples-tissue1}
jack.tissue1 <- jack.bind %>%
  dplyr::filter(Biotype=="protein_coding") %>%
  dplyr::filter(Tissue==params$tissue1 & Sex==params$sex)

expr.matrix.tissue1 <- samples.norm.within$log2.tmm.cpm.gf[[params$tissue1]][[params$sex]] %>%
  .[rownames(.) %in% jack.tissue1$GeneID,]
```

### Compute z-score
```{r compute-zscore-tissue1}
zscore.matrix.tissue1 <- zscore(expr.matrix.tissue1)
```

### Bimodality test
```{r compute-bimodality-index-tissue1}
bimodality.test.tissue1 <- 
  bimodalIndex(zscore.matrix.tissue1) %>%
  tibble::rownames_to_column(var="GeneID")
```

```{r}
summary(bimodality.test.tissue1$BI)
```

### Variability rank - Bimodality index

#### Bin variability ranks
```{r bin-var-ranks-tissue1}
bimodality.test.tissue1$VarRank <-
  vapply(bimodality.test.tissue1$GeneID, FUN.VALUE=numeric(1), FUN=function(g) {
    jack.tissue1$Mean_Local_Rank_Log2CV[jack.tissue1$GeneID==g]
  })

# Group variability ranks into bins of size 0.1
bimodality.test.tissue1$VarRankBin <- round(bimodality.test.tissue1$VarRank,1)
```

```{r}
dplyr::count(bimodality.test.tissue1, VarRankBin)
```

#### Boxplot - By variability rank bin
```{r boxplot-bimodality-by-var-rank-tissue1, fig.width=12, fig.height=12}
boxplot_bimodality_by_variability(
  bimodality.test.tissue1, 
  title=params$species.name, subtitle=paste(params$tissue1, params$sex),
  facet=FALSE)
```

### Expression level z-scores by BI value

**Question:** At which value of BI does the z-score distribution visually appear bimodal?

#### Bin BI values
```{r bin-bi-values-tissue1}
# Group BI values into bins of size 0.1
bimodality.test.tissue1$BIBin <- round(bimodality.test.tissue1$BI,1)

dplyr::count(bimodality.test.tissue1, BIBin)
```

#### Density plot
```{r plot-density-zscore-bi-tissue1, fig.width=8, fig.height=5}
plot_density_zscore_by_bi(zscore.matrix.tissue1, bimodality.test.tissue1, title="")
```

### Expression level z-scores by variability rank bin

**Question:** At which variability rank does the z-score distribution visually appear bimodal?

```{r plot-density-zscore-tissue1, fig.width=8, fig.height=5}
plot_density_zscore(zscore.matrix.tissue1, bimodality.test.tissue1, title="")
```

## Simulated strong bimodality

### Pre-processing

#### Generate expression matrix from two conditions
```{r generate-count-matrix-tissue1-2}
set.seed(params$set.seed)
replicates.tissue1 <- 
  sample(colnames(samples.norm.within$counts[[params$tissue1]][[params$sex]]), 
         size=params$total.replicates/2, 
         replace=FALSE)
replicates.tissue2 <- 
  sample(colnames(samples.norm.within$counts[[params$tissue2]][[params$sex]]), 
         size=params$total.replicates/2, 
         replace=FALSE)

metadata.tissue1.2 <- samples.norm.across$metadata %>%
  dplyr::filter(SampleName %in% c(replicates.tissue1, replicates.tissue2))

counts.tissue1.2 <- samples.norm.across$counts[,c(replicates.tissue1, replicates.tissue2)]
```

#### Gene filtering and normalization
```{r filter-genes-tissue1-2}
counts.gf.tissue1.2 <- 
  filter_genes_by_cpm(counts.matrix=counts.tissue1.2, 
                      metadata=metadata.tissue.1.2, 
                      min.cpm=params$min.cpm)
log2.tmm.cpm.gf.tissue1.2 <-
  transform_counts_to_tmm_cpm(counts.matrix=counts.gf.tissue1.2, 
                              metadata=metadata.tissue1.2,
                              log2=TRUE, shift.min=TRUE)
```

#### Jackknifed expression variability estimates
```{r run-variability-jackknife-tissue1-2}
jack.tissue1.2 <- 
  jackknife_resid_log2cv(expr.matrix=log2.tmm.cpm.gf.tissue1.2,
                         min.percentile=params$min.percentile,
                         max.percentile=params$max.percentile,
                         win.size=params$win.size)
jack.tissue1.2 <- jack.tissue1.2[["summary"]] %>%
  tibble::rownames_to_column(var="GeneID")
```

##### Filter by biotype
```{r filter-biotype-tissue1-2}
jack.tissue1.2 <- jack.tissue1.2 %>%
  dplyr::inner_join(biotype[,c("Gene.stable.ID","Gene.type")],
                    by=c("GeneID"="Gene.stable.ID"))

colnames(jack.tissue1.2)[length(jack.tissue1.2)] <- "Biotype"

jack.tissue1.2 <- jack.tissue1.2 %>%
   dplyr::filter(Biotype=="protein_coding")
```

#### Filter expression matrix
```{r filter-expression-matrix-tissue1-2}
expr.matrix.tissue1.2 <- 
  log2.tmm.cpm.gf.tissue1.2[rownames(log2.tmm.cpm.gf.tissue1.2) %in% jack.tissue1.2$GeneID,]
```

### Compute z-score
```{r compute-zscore-tissue1-2}
zscore.matrix.tissue1.2 <- zscore(expr.matrix.tissue1.2)
```

### Bimodality test
```{r compute-bimodality-index-tissue1-2}
bimodality.test.tissue1.2 <- 
  bimodalIndex(zscore.matrix.tissue1.2) %>%
  tibble::rownames_to_column(var="GeneID")
```

```{r}
summary(bimodality.test.tissue1.2$BI)
```

### Variability rank - Bimodality index

#### Bin variability ranks
```{r bin-var-ranks-tissue1-2}
bimodality.test.tissue1.2$VarRank <-
  vapply(bimodality.test.tissue1.2$GeneID, FUN.VALUE=numeric(1), FUN=function(g) {
    jack.tissue1.2$Mean_Local_Rank_Log2CV[jack.tissue1.2$GeneID==g]
  })

# Group variability ranks into bins of size 0.1
bimodality.test.tissue1.2$VarRankBin <- round(bimodality.test.tissue1.2$VarRank,1)
```

```{r}
dplyr::count(bimodality.test.tissue1.2, VarRankBin)
```

#### Boxplot - By variability rank bin
```{r boxplot-bimodality-by-var-rank-tissue1-2, fig.width=12, fig.height=12}
boxplot_bimodality_by_variability(
  bimodality.test.tissue1.2, 
  title=params$species.name, subtitle=paste(params$tissue1, params$tissue2, params$sex),
  facet=FALSE)
```

```{r boxplot-bimodality-by-var-rank-tissue1-2-v2, fig.width=12, fig.height=12}
# Single color, set ylim
boxplot_bimodality_by_variability_v2(
  bimodality.test.tissue1.2, 
  title=params$species.name, subtitle=paste(params$tissue1, params$tissue2, params$sex))
```

### Expression level z-scores by BI value

**Question:** At which value of BI does the z-score distribution visually appear bimodal?

#### Bin BI values
```{r bin-bi-values-tissue1-2}
# Group BI values into bins of size 0.1
bimodality.test.tissue1.2$BIBin <- round(bimodality.test.tissue1.2$BI,1)

dplyr::count(bimodality.test.tissue1.2, BIBin)
```

#### Density plot
```{r plot-density-zscore-bi-tissue1-2, fig.width=8, fig.height=5}
plot_density_zscore_by_bi(zscore.matrix.tissue1.2, bimodality.test.tissue1.2, title="")
```

### Expression level z-scores by variability rank bin

**Question:** At which variability rank does the z-score distribution visually appear bimodal?

```{r plot-density-zscore-tissue1-2, fig.width=8, fig.height=5}
plot_density_zscore(zscore.matrix.tissue1.2, bimodality.test.tissue1.2, title="")
```

## Comparison

### Combine observed and mixed dataset
```{r combine-observed-mixed-datasets}
bimodality.test.combined <- 
  dplyr::bind_rows(list("observed"=bimodality.test.tissue1, "mixed"=bimodality.test.tissue1.2), .id="Dataset")
bimodality.test.combined$Dataset <- factor(bimodality.test.combined$Dataset, levels=c("observed","mixed"))
```

### Boxplot - By variability rank bin
```{r boxplot-bimodality-by-var-rank-combined, fig.width=16, fig.height=12}
boxplot_bimodality_by_variability_comparison(
  bimodality.test.combined, title=params$species.name)
```

# Save
```{r save-objects}
if (!dir.exists(params$output.rds.path)) { dir.create(params$output.rds.path, showWarnings=FALSE) }

# Observed
saveRDS(zscore.matrix.tissue1,
        file.path(params$output.rds.path, 
                  paste(params$species, "zscore.matrix.tissue1.rds", sep="_")))

saveRDS(bimodality.test.tissue1,
        file.path(params$output.rds.path, 
                  paste(params$species, "bimodality.test.tissue1.rds", sep="_")))

# Simulated
saveRDS(jack.tissue1.2,
        file.path(params$output.rds.path, 
                  paste(params$species, "jack.tissue1.2.rds", sep="_")))

saveRDS(expr.matrix.tissue1.2,
        file.path(params$output.rds.path, 
                  paste(params$species, "expr.matrix.tissue1.2.rds", sep="_")))

saveRDS(zscore.matrix.tissue1.2,
        file.path(params$output.rds.path, 
                  paste(params$species, "zscore.matrix.tissue1.2.rds", sep="_")))

saveRDS(bimodality.test.tissue1.2,
        file.path(params$output.rds.path, 
                  paste(params$species, "bimodality.test.tissue1.2.rds", sep="_")))

# Combined
saveRDS(bimodality.test.combined,
        file.path(params$output.rds.path, 
                  paste(params$species, "bimodality.test.combined.rds", sep="_")))
```

