---
title: "GO Enrichment (*`r params$species.name`*)"
author: "Bucao, Christabel Floi"
output: html_document
date: "`r Sys.Date()`"
params:
  input.samples.rds.path: "../../results/01_Filter_Samples"
  input.variability.rds.path: "../../results/03_Variability_Plots"
  output.path: "../../results/07_GO_Enrichment"
  species: "DRE"
  species.name: "Danio rerio"
  ev.percentile: 0.20
  pvalue.cutoff: 0.01
  condition.cutoff: 3
---

# Setup {.tabset}

## Parameters
```{r print-params}
print(params)
```

## Libraries
```{r import-libraries}
library(dplyr)
library(parallel)
library(GOstats)
library(org.Dr.eg.db)
```

## Functions
```{r source-functions}
source("../functions/go_enrichment.R")
```

## Session Info
```{r print-session-info}
sessionInfo()
```

## Data

### Expression levels
```{r import-expression-levels}
# Load saved data from 01_Filter_Samples.Rmd
## Normalized separately per condition
filtered.samples <- readRDS(
  file.path(params$input.samples.rds.path,
            paste(params$species, "4_reps_data.conditions.rds", sep="_")))
```

### Expression variability
```{r import-variability}
# Load processed expression variability estimates from 03_Variability_Plots.Rmd
jack.summary <- readRDS(
  file.path(params$input.variability.rds.path,
            paste(params$species, "jack.resid.cv.summary.rds", sep="_")))
```

# Main {.tabset}

## Lowly and highly variable genes genes

### Retrieve top and bottom x% of genes by variability rank
```{r retrieve-lv-hv-genes}
Tissue <- as.factor(filtered.samples$metadata$Tissue)
if ("Sex" %in% colnames(filtered.samples$metadata)) { 
  Sex <- as.factor(filtered.samples$metadata$Sex) 
}

if ("Sex" %in% colnames(filtered.samples$metadata)) { 
  ev.genes <- 
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        retrieve_variability_gene_lists(jack.summary[[t]][[s]], 
                                        ev.percentile=params$ev.percentile)
    })
  })
} else {
  ev.genes <- 
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
        retrieve_variability_gene_lists(jack.summary[[t]], 
                                        ev.percentile=params$ev.percentile)
  })    
}
```

### Convert Ensembl to Entrez IDs for each gene list
```{r convert-ensembl-to-entrez}
keys <- keys(org.Dr.eg.db, keytype="ENSEMBL")
entrez <- 
  mapIds(org.Dr.eg.db, keys=keys, column="ENTREZID", keytype="ENSEMBL", multiVals="first")

if ("Sex" %in% colnames(filtered.samples$metadata)) {
  entrez.ids <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        lapply(c("LV"="LV","MV"="MV","HV"="HV","Bgd"="Bgd"), function(g) {
        entrez[ev.genes[[t]][[s]][[g]]]
        })
      })
    })  
} else {
  entrez.ids <-
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(c("LV"="LV","MV"="MV","HV"="HV","Bgd"="Bgd"), function(g) {
        entrez[ev.genes[[t]][[g]]]
      })
    })   
}
```

## GO Stats Hypergeometric Test

### Set parameters
```{r set-go-stats-params}
if ("Sex" %in% colnames(filtered.samples$metadata)) {
  go.params <- 
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        lapply(c("LV"="LV","MV"="MV","HV"="HV"), function(ev) {
          lapply(c("over"="over","under"="under"), function(test) {
            new("GOHyperGParams", 
                geneIds=entrez.ids[[t]][[s]][[ev]],
                universeGeneIds=entrez.ids[[t]][[s]][["Bgd"]],
                annotation=org.Dr.eg.db,
                ontology="BP",
                pvalueCutoff=params$pvalue.cutoff,
                conditional=TRUE,
                testDirection=test)
          })
        })
      })
    })  
} else {
  go.params <- 
    lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(c("LV"="LV","MV"="MV","HV"="HV"), function(ev) {
        lapply(c("over"="over","under"="under"), function(test) {
          new("GOHyperGParams", 
              geneIds=entrez.ids[[t]][[ev]],
              universeGeneIds=entrez.ids[[t]][["Bgd"]],
              annotation=org.Dr.eg.db,
              ontology="BP",
              pvalueCutoff=0.01,
              conditional=TRUE,
              testDirection=test)
        })
      })
    })    
}
```

### Run conditional hypergeometric test
```{r run-go-stats}
if ("Sex" %in% colnames(filtered.samples$metadata)) {
  go.results <- 
    mclapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        lapply(c("LV"="LV","MV"="MV","HV"="HV"), function(ev) {
          lapply(c("over"="over","under"="under"), function(test) {
            hyperGTest(go.params[[t]][[s]][[ev]][[test]])
          })
        })
      })
    }, mc.cores=4)  
} else {
  go.results <- 
    mclapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(c("LV"="LV","MV"="MV","HV"="HV"), function(ev) {
        lapply(c("over"="over","under"="under"), function(test) {
          hyperGTest(go.params[[t]][[ev]][[test]])
          })
        })
    }, mc.cores=4)  
}
```

### Summarize results
```{r summarize-go-stats}
if ("Sex" %in% colnames(filtered.samples$metadata)) { 
  go.summary <- 
    mclapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(setNames(levels(Sex), levels(Sex)), function(s) {
        lapply(c("LV"="LV","MV"="MV","HV"="HV"), function(ev) {
          lapply(c("over"="over","under"="under"), function(test) {
            summary(go.results[[t]][[s]][[ev]][[test]])
          })
        })
    })
  }, mc.cores=4)
} else {
  go.summary <- 
    mclapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
      lapply(c("LV"="LV","MV"="MV","HV"="HV"), function(ev) {
        lapply(c("over"="over","under"="under"), function(test) {
          summary(go.results[[t]][[ev]][[test]])
        })
      })
  }, mc.cores=4)  
}
```

### Aggregate results from multiple GO enrichment tests
```{r aggregate-go-results}
go.aggregate <- combine_go_results(go.summary)
```

### Export results for external plotting
```{r export-results}
export_go_results(go.aggregate, params$condition.cutoff, 
                  output.dir=params$output.path, prefix=params$species)
```

# Save
```{r save-objects}
if (!dir.exists(params$output.path)) { dir.create(params$output.path, showWarnings=FALSE) }

saveRDS(entrez.ids,
        file.path(params$output.path, 
                  paste(params$species, "entrez.ids.rds", sep="_")))

saveRDS(ev.genes,
        file.path(params$output.path, 
                  paste(params$species, "ev.genes.rds", sep="_")))

saveRDS(go.params,
        file.path(params$output.path, 
                  paste(params$species, "go.params.rds", sep="_")))

saveRDS(go.results,
        file.path(params$output.path, 
                  paste(params$species, "go.results.rds", sep="_")))

saveRDS(go.summary,
        file.path(params$output.path, 
                  paste(params$species, "go.summary.rds", sep="_")))

saveRDS(go.aggregate,
        file.path(params$output.path, 
                  paste(params$species, "go.aggregate.rds", sep="_")))
```

