---
title: "Gene Set Enrichment (*`r params$species.name`*)"
author: "Bucao, Christabel Floi"
output: html_document
date: "`r Sys.Date()`"
params:
  input.samples.path: "../../results/01_Filter_Samples"
  input.variability.path: "../../results/03_Variability_Plots"
  output.path: "../../results/07b_GO_cameraPR"
  species: "DRE"
  species.name: "Danio rerio"
  combat: FALSE
  qc1.only: FALSE
  min.genes: 20
  max.genes: 1000
  condition.cutoff: 3

---

# Setup {.tabset}

## Parameters
```{r print-params}
print(params)
```

## Libraries
```{r import-libraries}
library(dplyr)
library(org.Dr.eg.db)
library(GO.db)
library(limma)

#renv::snapshot()
```

## Functions
```{r source-functions}

```

## Session Info
```{r print-session-info}
sessionInfo()
```


## Data

### Expression levels
```{r import-expression-levels}
# Load saved data from 01_Filter_Samples.Rmd
if (!params$combat) {
  if (!params$qc1.only) {
    input.samples.path <- file.path(params$input.samples.path, "no_combat", "full_qc")
    
    ## Normalized separately per condition
    filtered.samples <- readRDS(
      file.path(input.samples.path,
                paste(params$species, "4_reps_data.conditions.rds", sep="_")))

    ## Normalized across all conditions
    filtered.samples.across <- readRDS(
      file.path(input.samples.path,
              paste(params$species, "4_reps_data.qc2.cor.rds", sep="_")))    
  } else {
    input.samples.path <- file.path(params$input.samples.path, "no_combat", "qc1_only")
    
    ## Normalized separately per condition
    filtered.samples <- readRDS(
      file.path(input.samples.path,
                paste(params$species, "4_reps_data.qc1.conditions.rds", sep="_")))
    
    ## Normalized across all conditions
    filtered.samples.across <- readRDS(
      file.path(input.samples.path,
                paste(params$species, "4_reps_data.qc1.seq.rds", sep="_")))
    
  }

} else {
  if (!params$qc1.only) {
    input.samples.path <- file.path(params$input.samples.path, "combat", "full_qc")
    
    ## Normalized separately per condition
    filtered.samples <- readRDS(
      file.path(input.samples.path,
                paste(params$species, "4_reps_data.conditions.rds", sep="_")))

    ## Normalized across all conditions
    filtered.samples.across <- readRDS(
      file.path(input.samples.path,
              paste(params$species, "4_reps_data.qc2.cor.rds", sep="_"))) 
    
  } else {
    
    input.samples.path <- file.path(params$input.samples.path, "combat", "qc1_only")
    
    ## Normalized separately per condition
    filtered.samples <- readRDS(
      file.path(input.samples.path,
                paste(params$species, "4_reps_data.qc1.conditions.rds", sep="_")))
    
    ## Normalized across all conditions
    filtered.samples.across <- readRDS(
      file.path(input.samples.path,
                paste(params$species, "4_reps_data.qc1.seq.rds", sep="_")))    
  }
}

```


### Expression variability
```{r import-variability}
if (!params$combat) {
  if (!params$qc1.only) {
    input.variability.path <- file.path(params$input.variability.path, "no_combat", "full_qc")    
  } else {
    input.variability.path <- file.path(params$input.variability.path, "no_combat", "qc1_only")        
  }

} else {
  if (!params$qc1.only) {
    input.variability.path <- file.path(params$input.variability.path, "combat", "full_qc")    
  } else {
    input.variability.path <- file.path(params$input.variability.path, "combat", "qc1_only")        
  }

}

# Load processed expression variability estimates from 03_Variability_Plots.Rmd
#jack.summary <- readRDS(
#  file.path(input.variability.path,
#            paste(params$species, "jack.resid.cv.summary.rds", sep="_")))

jack.bind <- readRDS(
  file.path(input.variability.path,
            paste(params$species, "jack.resid.cv.bind.rds", sep="_")))

```


# Main {.tabset}

## Retrieve genes per GO term
```{r retrieve-genes-per-go}
min.genes <- params$min.genes
max.genes <- params$max.genes

# Map GO terms to genes
go.sets <- as.list(org.Dr.egGO2ALLEGS)
go.sets.filt <- go.sets[lengths(go.sets)>=min.genes & lengths(go.sets)<=max.genes]

go.sets.bind <- data.frame("GOID"=names(unlist(go.sets.filt)), "Entrez"=unlist(go.sets.filt))
go.sets.bind$EvidenceCode <- limma::strsplit2(go.sets.bind$GOID, split="[.]")[,2]
go.sets.bind$GOID <- limma::strsplit2(go.sets.bind$GOID, split="[.]")[,1]

go2term <- AnnotationDbi::select(GO.db, 
                                 keys=unique(go.sets.bind$GOID),
                                 columns=c("GOID","TERM","ONTOLOGY"),
                                 keytype="GOID")

go.sets.bind <- go.sets.bind %>%
  dplyr::left_join(go2term,by="GOID")
```

## Run cameraPR for GO terms sets with variability metric as statistic
```{r run-camera}
# Convert Ensembl to Entrez IDs for each gene list 
keys <- keys(org.Dr.eg.db, keytype="ENSEMBL")
entrez <- 
  mapIds(org.Dr.eg.db, keys=keys, column="ENTREZID", keytype="ENSEMBL", multiVals="first")

jack.bind$Entrez <- vapply(jack.bind$GeneID, FUN.VALUE=character(1), FUN=function(g) {
  entrez[g]
})

# Run cameraPR
Tissue <- as.factor(filtered.samples$metadata$Tissue)
Sex <- as.factor(filtered.samples$metadata$Sex)

camera.test <-
  lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
    lapply(setNames(levels(Sex), levels(Sex)), function(s) {
      print(paste(t,s))
      
      var.metric <- jack.bind %>%
        dplyr::filter(Tissue==t & Sex==s) %>%
        dplyr::filter(Entrez %in% go.sets.bind$Entrez) %>%
        dplyr::pull(Mean_Local_Rank_Log2CV, Entrez)
      
      cameraPR(statistic=var.metric, index=go.sets.filt, use.ranks=TRUE) %>%
        tibble::rownames_to_column(var="GOID") %>%
        dplyr::left_join(go2term, by="GOID")
    })
  })
```

```{r table-camera-signif}
# How many GO terms are significantly highly variable ("Up") or lowly variable ("Down")?
for (t in levels(Tissue)) {
  for (s in levels(Sex)) {
    print(paste(t,s, "Up"))
    print(table(camera.test[[t]][[s]]$Direction=="Up", camera.test[[t]][[s]]$FDR<0.05))
    
    print(paste(t,s, "Down"))
    print(table(camera.test[[t]][[s]]$Direction=="Down", camera.test[[t]][[s]]$FDR<0.05))    
  }
}
```


## Significantly highly or lowly variable GO terms
```{r retrieve-camera-signif}
camera.signif <-
  lapply(setNames(levels(Tissue), levels(Tissue)), function(t) {
    lapply(setNames(levels(Sex), levels(Sex)), function(s) {
      lapply(c("Up"="Up", "Down"="Down"), function(d) {
        camera.test[[t]][[s]] %>%
          dplyr::filter(Direction==d) %>%
          dplyr::filter(FDR<0.05)
      })
    })
  })
```


## Aggregate results from multiple cameraPR tests
```{r aggregate-camera-results}
camera.bind <- list("Up"=NULL, "Down"=NULL)

for (t in levels(Tissue)) {
  for (s in levels(Sex)) {
    for (d in c("Up","Down")) {
      camera.bind[[d]] <- rbind(camera.bind[[d]], camera.signif[[t]][[s]][[d]])
    }
  }
}

for (d in c("Up","Down")) {
  camera.bind[[d]] <- camera.bind[[d]] %>%
    dplyr::group_by(GOID,Direction,TERM,ONTOLOGY) %>%
    dplyr::summarise(MeanPvalue=mean(PValue),
                     MeanFDR=mean(FDR),
                     NConditions=n()) %>%
    dplyr::ungroup()
}
```

# Save {.tabset}

## Expert GO results for external plotting
```{r export-results}
if (!params$combat) {
  if (!params$qc1.only) {
    output.path <- file.path(params$output.path, "no_combat", "full_qc")    
  } else {
    output.path <- file.path(params$output.path, "no_combat", "qc1_only")        
  }

} else {
  if (!params$qc1.only) {
  output.path <- file.path(params$output.path, "combat", "full_qc")    
  } else {
  output.path <- file.path(params$output.path, "combat", "qc1_only")        
  }

}

terms.dir <- file.path(output.path, "terms")
if (!dir.exists(terms.dir)) { dir.create(terms.dir, recursive=TRUE) }

go.export <- lapply(c("Up"="Up","Down"="Down"), function(d) {
  camera.bind[[d]] %>%
    dplyr::filter(NConditions>=params$condition.cutoff) %>%
    dplyr::select(GOID, MeanFDR, NConditions)
})

tmp_filepath <- file.path(terms.dir, paste0(paste(params$species,"up",sep="_"),".txt"))
write.table(go.export[["Up"]],tmp_filepath, quote=FALSE, row.names=FALSE, col.names=FALSE, sep="\t")

tmp_filepath <- file.path(terms.dir, paste0(paste(params$species,"down",sep="_"),".txt"))
write.table(go.export[["Down"]],tmp_filepath, quote=FALSE, row.names=FALSE, col.names=FALSE, sep="\t")
```

## Save RDS
```{r save-objects}
saveRDS(camera.test,
        file.path(output.path, 
                  paste(params$species, "camera.test.rds", sep="_")))

saveRDS(camera.signif,
        file.path(output.path, 
                  paste(params$species, "camera.signif.rds", sep="_")))

saveRDS(camera.bind,
        file.path(output.path, 
                  paste(params$species, "camera.bind.rds", sep="_")))
```

