---
title: "Selection analysis (*`r params$species.name`*)"
author: "Bucao, Christabel Floi"
output: html_document
date: "`r Sys.Date()`"
params:
  selectome.path: "../../data/selectome"
  input.samples.rds.path: "../../results/01_Filter_Samples"
  input.variability.rds.path: "../../results/03_Variability_Plots"
  output.rds.path: "../../results/08_Selection"
  species: "DRE"
  species.name: "Danio rerio"
  ev.percentile: 0.20
  set.seed: 12345
  n.permutations: 2000
  qc1.only: FALSE
  combat: FALSE
---

# Setup {.tabset}

## Parameters
```{r print-params}
print(params)
```

## Libraries
```{r import-libraries}
library(dplyr)
library(matrixStats)
```

## Functions
```{r source-functions}
source("../functions/selection.R")
source("../functions/selection_plots.R")
```

## Session Info
```{r print-session-info}
sessionInfo()
```

## Data

### Expression levels
```{r import-expression-levels}
if (!params$combat) {
  if (!params$qc1.only) {
    input.samples.path <- file.path(params$input.samples.rds.path, "no_combat", "full_qc")    
  } else {
    input.samples.path <- file.path(params$input.samples.rds.path, "no_combat", "qc1_only")        
  }

} else {
  if (!params$qc1.only) {
    input.samples.path <- file.path(params$input.samples.rds.path, "combat", "full_qc")    
  } else {
    input.samples.path <- file.path(params$input.samples.rds.path, "combat", "qc1_only")        
  }

}

# Load saved data from 01_Filter_Samples.Rmd
# Normalized within-condition
if (!params$qc1.only) {
  filtered.samples <- readRDS(
    file.path(input.samples.path,
              paste(params$species, "4_reps_data.conditions.rds", sep="_")))    
} else {
  filtered.samples <- readRDS(
    file.path(input.samples.path,
              paste(params$species, "4_reps_data.qc1.conditions.rds", sep="_")))      
}

```

### Expression variability
```{r import-variability}
if (!params$combat) {
  if (!params$qc1.only) {
    input.variability.path <- file.path(params$input.variability.rds.path, "no_combat", "full_qc")    
  } else {
    input.variability.path <- file.path(params$input.variability.rds.path, "no_combat", "qc1_only")        
  }

} else {
  if (!params$qc1.only) {
    input.variability.path <- file.path(params$input.variability.rds.path, "combat", "full_qc")    
  } else {
    input.variability.path <- file.path(params$input.variability.rds.path, "combat", "qc1_only")        
  }

}

# Load processed expression variability estimates from 03_Variability_Plots.Rmd
jack.summary <- readRDS(
  file.path(input.variability.path,
            paste(params$species, "jack.resid.cv.summary.rds", sep="_")))

```

### Selectome
```{r import-selectome-dnds}
# Load dN/dS results from Selectome
selectome <- 
  read.delim(file.path(
    params$selectome.path, 
    paste(params$species, "protein_coding_Ensembl_105.res.txt", sep="_")),
    sep="\t")
colnames(selectome) <- c("GeneID", "TaxID", "Taxon", "qvalue", "LRT", "omega0", "omega2")
```

# Main {.tabset}

## Gene expression variability across conditions

### Variability bins

#### Merge variability rank
```{r merge-ev-rank}
ev.rank <- 
  merge_ev_rank(jack.summary, filtered.samples$metadata, by="local", join="full", method="cv")
```

#### Classify genes as low, high, or mid variability (LV, HV, MV) based on percentile cutoff
```{r classify-ev}
if ("Sex" %in% colnames(filtered.samples$metadata)) { 
  ev.tally <- vector(mode="list",3)
  names(ev.tally) <- c("all","F","M")

  ev.tally$all <- 
    tally_ev(ev.rank, 
             ev.percentile=params$ev.percentile, output="all")
  ev.tally$F <- 
    tally_ev(ev.rank[,grepl("_F",colnames(ev.rank))], 
             ev.percentile=params$ev.percentile, output="all")
  ev.tally$M <- 
    tally_ev(ev.rank[,grepl("_M",colnames(ev.rank))], 
             ev.percentile=params$ev.percentile, output="all")
} else {
  ev.tally <- 
    tally_ev(ev.rank,
             ev.percentile=params$ev.percentile, output="all")
}
```

Criteria:
(1) Consider only genes with variability data in at least n.conditions
(2) Classify genes as pan-LV (or HV, MV) if it is LV (HV, MV) in more than half of conditions
```{r classify-pan-ev}
ev.cat <- c("LV"="LV","MV"="MV","HV"="HV","Bgd"="Bgd")
n.conditions <- floor(nrow(filtered.samples$conditions)/2 + 1)

# The background gene set includes genes which satisfy (1) but not necessarily (2)

if ("Sex" %in% colnames(filtered.samples$metadata)) {
  ev.tally.sort <- lapply(ev.cat, function(ev) {
    if (ev=="Bgd") { 
      ev.tally$all$values[ev.tally$all$values$N_Conditions>=n.conditions,] 
    }
    else { 
      ev.tally$all$values[
        ev.tally$all$values$N_Conditions>=n.conditions &
          ev.tally$all$values[[paste(ev, "Ratio", sep="_")]]>0.5,] }
  })  
} else {
  ev.tally.sort <- lapply(ev.cat, function(ev) {
    if (ev=="Bgd") { 
      ev.tally$values[ev.tally$values$N_Conditions>=n.conditions,] 
    }
    else { 
      ev.tally$values[
        ev.tally$values$N_Conditions>=n.conditions &
          ev.tally$values[[paste(ev, "Ratio", sep="_")]]>0.5,] }
  })
}
```

### Median variability across conditions

#### Compute median variability across conditions
```{r compute-median-ev}
# Without gene filtering
ev.rank.median <-
  matrixStats::rowMedians(as.matrix(ev.rank), useNames=TRUE, na.rm=TRUE)

# With gene filtering
## Consider only genes with variability data in at least n.conditions
ev.rank.filtered <- ev.rank[rowSums(!is.na(ev.rank))>=n.conditions,]
ev.rank.filtered.median <- 
  matrixStats::rowMedians(as.matrix(ev.rank.filtered), useNames=TRUE, na.rm=TRUE)
```


## Selection analysis

### Filter Selectome results

#### Remove GeneID - Taxon duplicates from Selectome results

GeneID - Taxon duplicated rows may be due to uncertainties in the gene tree
```{r remove-selectome-duplicates}
# No duplicated rows
selectome.uniq <-
  dplyr::distinct(selectome, GeneID, TaxID, Taxon, .keep_all=TRUE)
```

#### Background gene set
```{r subset-selectome-results-bgd}
# Larger gene list without pan-LV, MV HV classifications
# The background gene set includes genes which have variability data in at least n.conditions
selectome.bgd <- selectome.uniq %>%
  dplyr::filter(GeneID %in% rownames(ev.tally.sort[["Bgd"]]))

# Group by GeneID
## The estimated omega0 is the same across all branches of each gene tree
selectome.bgd.groupby.gene <- selectome.bgd %>%
  dplyr::group_by(GeneID) %>%
  dplyr::summarise(
    omega0 = mean(omega0)
  ) %>%
dplyr::ungroup()

# Number of unique gene IDs
dplyr::n_distinct(selectome.bgd.groupby.gene$GeneID)
```

##### Group median variability rank into bins
```{r group-selectome-bgd-varbins}
# Add median variability across conditions
selectome.bgd.groupby.gene.ev.median <- selectome.bgd.groupby.gene %>%
  dplyr::inner_join(data.frame("GeneID"=names(ev.rank.median),"MedianVar"=ev.rank.median),by="GeneID")

# Option 1: Group median variability rank into bins by rounding to the first decimal place
#selectome.bgd.groupby.gene.ev.median$VarRankBin <-
#  round(selectome.bgd.groupby.gene.ev.median$MedianVar, 1)

# Option 2: Group median variability rank into 10 bins of size 0.1
selectome.bgd.groupby.gene.ev.median$VarRankBin <-
  ceiling(selectome.bgd.groupby.gene.ev.median$MedianVar*10)/10

# Define lowly and highly variable genes based on ev.percentile cutoff
# e.g., for ev.percentile=0.20
# Given 10 equally sized bins (0-0.1, 0.1-0.2, ..., 0.8-0.9, 0.9-1.0)
# bins 0.1 and 0.2 are considered lowly variable
# bins 0.9 and 1.0 are considered highly variable
selectome.bgd.groupby.gene.ev.median$VarRankCategory <-
  apply(selectome.bgd.groupby.gene.ev.median, MARGIN=1, function(r) {
    if (r[["VarRankBin"]] <= params$ev.percentile) { "low" }
    else if (r[["VarRankBin"]] >= (1.0-params$ev.percentile+0.1)) { "high" }
    else { "mid" }
  })

# Add any missing bins for visualization purposes
selectome.bgd.groupby.gene.ev.median <- selectome.bgd.groupby.gene.ev.median %>%
  check_missing_bins(all.bins=seq(0.1,1,0.1))
```

##### Revision: Boxplot - Purifying selection with variability rank bins
```{r boxplot-omega0-varbins, fig.width=10, fig.height=8}
boxplot_omega0_varbins(selectome.bgd.groupby.gene.ev.median, title=params$species.name)
boxplot_omega0_varbins(selectome.bgd.groupby.gene.ev.median, compare.extremes=TRUE, title=params$species.name)
```

##### Revision: Scatterplot
```{r scatterplot-omega0-bgd, fig.width=6, fig.height=6}
scatterplot_omega0_variability(selectome.bgd.groupby.gene.ev.median, title=params$species.name)
```

#### Version 1.0: Classified by low, mid, high variability
```{r subset-selectome-results}
# Criteria:
# (1) Consider only genes with variability data in at least n.conditions
# (2) Classify genes as pan-LV (or HV, MV) if it is LV (HV, MV) in more than half of conditions

selectome.sub <- lapply(ev.cat, function(ev) {
  dplyr::filter(selectome.uniq, GeneID %in% rownames(ev.tally.sort[[ev]]))
})

selectome.sub.bind <-
  dplyr::bind_rows(selectome.sub[c("LV","MV","HV")], .id="EV")

selectome.sub.bind$EV <- 
  factor(selectome.sub.bind$EV, levels=c("LV","MV","HV"), labels=c("low","mid","high"))

# Group by GeneID
## The estimated omega0 is the same across all branches of each gene tree
selectome.sub.bind.groupby.gene <- selectome.sub.bind %>%
  dplyr::group_by(EV,GeneID) %>%
  dplyr::summarise(
    omega0 = mean(omega0)
  ) %>%
  dplyr::ungroup()

# Number of unique gene IDs
dplyr::n_distinct(selectome.sub.bind.groupby.gene$GeneID)

# Branches
print(levels(as.factor(selectome.sub$Bgd$Taxon)))
```

###### Version 1.0: Boxplot - Purifying selection with variability categories
```{r boxplot-omega0, fig.width=8, fig.height=10}
# With pan-LV, MV, HV classifications
boxplot_omega0(selectome.sub.bind.groupby.gene, title=params$species.name)
boxplot_omega0(selectome.sub.bind.groupby.gene, compare.means=TRUE, title=params$species.name)
```

### Positive selection

#### Observed proportion of branches under positive selection (q-value < 0.05)
```{r positive-selection-enrichment}
# Based on version 1.0: Variability categories (low, mid, high variability)

selectome.enrichment <- data.frame(
  "Category"=ev.cat,
  "Selection_Yes"=rep(NA,length(ev.cat)),
  "Selection_No"=rep(NA,length(ev.cat)),
  "Proportion"=rep(NA,length(ev.cat)),
  "Enrichment"=rep(NA,length(ev.cat))
)

for (ev in ev.cat[length(ev.cat):1]) {
  selectome.enrichment[ev,"Selection_Yes"] <- sum(selectome.sub[[ev]]$qvalue<0.05)
  selectome.enrichment[ev,"Selection_No"] <- sum(selectome.sub[[ev]]$qvalue>=0.05)
  selectome.enrichment[ev,"Proportion"] <-
    sum(selectome.sub[[ev]]$qvalue<0.05) / nrow(selectome.sub[[ev]])
  selectome.enrichment[ev,"Enrichment"] <-
    selectome.enrichment[ev,"Proportion"] /
    (sum(selectome.sub[["Bgd"]]$qvalue<0.05) / nrow(selectome.sub[["Bgd"]]))
}
```

```{r}
print(selectome.enrichment)
```

#### Chi-squared contingency table
```{r positive-selection-chisq}
selectome.chisq.table <- selectome.enrichment[c("LV","MV","HV"),c("Selection_Yes","Selection_No")]

selectome.chisq.test <- chisq.test(selectome.chisq.table)
selectome.chisq.test
```

```{r}
selectome.chisq.test$observed
selectome.chisq.test$expected
```

#### Permutation test

##### Permute q-value
```{r permute-qvalue}
selectome.perm <- vector(mode="list", length=params$n.permutations)

set.seed(params$set.seed)

for (i in 1:params$n.permutations) {
  selectome.perm[[i]] <- selectome.sub$Bgd
  selectome.perm[[i]]$qvalue <- sample(selectome.perm[[i]]$qvalue)
}
```

##### Subset by gene lists
```{r subset-gene-lists-permutation}
selectome.perm.sub <- vector(mode="list", length=params$n.permutations)

for (i in 1:params$n.permutations) {
  selectome.perm.sub[[i]] <- lapply(ev.cat, function(ev) {
    dplyr::filter(selectome.perm[[i]], GeneID %in% rownames(ev.tally.sort[[ev]]))
  })
}
```

##### Proportion of branches under positive selection
```{r positive-selection-enrichment-permutation}
selectome.perm.enrichment <- vector(mode="list", length=3)
names(selectome.perm.enrichment) <- c("LV","MV","HV")

for (ev in c("LV","MV","HV")) {
  selectome.perm.enrichment[[ev]] <- data.frame(
    "Selection_Yes"=rep(NA, params$n.permutations),
    "Selection_No"=rep(NA, params$n.permutations),
    "Proportion"=rep(NA, params$n.permutations),
    "Enrichment"=rep(NA, params$n.permutations),
    "PercentRank"=rep(NA, params$n.permutations)
  )
  
  for (i in 1:params$n.permutations) {
    selectome.perm.enrichment[[ev]][i,"Selection_Yes"] <- sum(selectome.perm.sub[[i]][[ev]]$qvalue<0.05)
    selectome.perm.enrichment[[ev]][i,"Selection_No"] <- sum(selectome.perm.sub[[i]][[ev]]$qvalue>=0.05)
    selectome.perm.enrichment[[ev]][i,"Proportion"] <- 
      sum(selectome.perm.sub[[i]][[ev]]$qvalue<0.05) / nrow(selectome.perm.sub[[i]][[ev]])
    
    selectome.perm.enrichment[[ev]][i,"Enrichment"] <-
      selectome.perm.enrichment[[ev]][i,"Proportion"] / 
      (sum(selectome.perm.sub[[i]][["Bgd"]]$qvalue<0.05) / nrow(selectome.perm.sub[[i]][["Bgd"]]))
  }
  
  selectome.perm.enrichment[[ev]][,"PercentRank"] <- 
    dplyr::percent_rank(selectome.perm.enrichment[[ev]]$Enrichment)
}
```

##### Compute p-value

Null hypothesis: 
- Expression variability has no association with positive selection.
- For each expression variability category, the proportion of gene - branches under positive selection is the same as the background proportion. (Enrichment = 1)

Alternative hypothesis:
- Expression variability is associated with positive selection (in either direction)

###### Lowly variable genes
```{r compute-pvalue-lv}
# Observed enrichment of gene - branches under positive selection
print(paste("Observed =", round(selectome.enrichment["LV","Enrichment"],3)))

# From the permutation test, how many runs have enrichment equal to or lower than this value?
sum(selectome.perm.enrichment$LV$Enrichment<=selectome.enrichment["LV","Enrichment"])

# Two-tailed p-value
## If the proportion is the same as the background, we expect Enrichment = 1
## From the permutation test, how many runs have enrichment more extreme than the observed value?
diff.cutoff <- rep(NA,3)
names(diff.cutoff) <- c("LV","MV","HV")

diff.cutoff[["LV"]] <- abs(1.0 - selectome.enrichment["LV","Enrichment"])

## If p-value = 0, it is at least p < (1/n.permutations)
perm.pvalue <- rep(NA,3)
names(perm.pvalue) <- c("LV","MV","HV")

perm.pvalue[["LV"]] <- 
  sum(abs(1.0 - selectome.perm.enrichment$LV$Enrichment)>=diff.cutoff[["LV"]]) / params$n.permutations

print(paste("p =", perm.pvalue[["LV"]]))
```

###### Highly variable genes
```{r compute-pvalue-hv}
# Observed enrichment of gene - branches under positive selection
print(paste("Observed =", round(selectome.enrichment["HV","Enrichment"],3)))

# From the permutation test, how many runs have enrichment equal to or lower than this value?
sum(selectome.perm.enrichment$HV$Enrichment<=selectome.enrichment["HV","Enrichment"])

# Two-tailed p-value
## If the proportion is the same as the background, we expect Enrichment = 1
## From the permutation test, how many runs have enrichment more extreme than the observed value?
diff.cutoff[["HV"]] <- abs(1.0 - selectome.enrichment["HV","Enrichment"])

## If p-value = 0, it is at least p < (1/n.permutations)
perm.pvalue[["HV"]] <- 
  sum(abs(1.0 - selectome.perm.enrichment$HV$Enrichment)>=diff.cutoff[["HV"]]) / params$n.permutations

print(paste("p =", perm.pvalue[["HV"]]))
```

###### Moderately variable genes
```{r compute-pvalue-mv}
# Observed enrichment of gene - branches under positive selection
print(paste("Observed =", round(selectome.enrichment["MV","Enrichment"],3)))

# From the permutation test, how many runs have enrichment equal to or lower than this value?
sum(selectome.perm.enrichment$MV$Enrichment<=selectome.enrichment["MV","Enrichment"])

# Two-tailed p-value
## If the proportion is the same as the background, we expect Enrichment = 1
## From the permutation test, how many runs have enrichment more extreme than the observed value?
diff.cutoff[["MV"]] <- abs(1.0 - selectome.enrichment["MV","Enrichment"])

## If p-value = 0, it is at least p < (1/n.permutations)
perm.pvalue[["MV"]] <- 
  sum(abs(1.0 - selectome.perm.enrichment$MV$Enrichment)>=diff.cutoff[["MV"]]) / params$n.permutations

print(paste("p =", perm.pvalue[["MV"]]))
```

# Save
```{r save-objects}
if (!params$combat) {
  if (!params$qc1.only) {
    output.path <- file.path(params$output.rds.path, "no_combat", "full_qc")    
  } else {
    output.path <- file.path(params$output.rds.path, "no_combat", "qc1_only")        
  }

} else {
  if (!params$qc1.only) {
  output.path <- file.path(params$output.rds.path, "combat", "full_qc")    
  } else {
  output.path <- file.path(params$output.rds.path, "combat", "qc1_only")        
  }

}

if (!dir.exists(output.path)) { dir.create(output.path, recursive=TRUE, showWarnings=FALSE) }

# Variability rank categories
saveRDS(ev.tally,
        file.path(output.path, 
                  paste(params$species, "ev.tally.rds", sep="_")))

saveRDS(ev.tally.sort,
        file.path(output.path, 
                  paste(params$species, "ev.tally.sort.rds", sep="_")))

# Filtered Selectome results
saveRDS(selectome.sub,
        file.path(output.path, 
                  paste(params$species, "selectome.sub.rds", sep="_")))

## Background gene set
saveRDS(selectome.bgd,
        file.path(output.path,
                  paste(params$species, "selectome.bgd.rds", sep="_")))

### Background gene set with median variability across conditions
saveRDS(selectome.bgd.groupby.gene.ev.median,
        file.path(output.path,
                  paste(params$species, "selectome.bgd.groupby.gene.ev.median.rds", sep="_")))

## Subset by variability rank categories
saveRDS(selectome.sub.bind,
        file.path(output.path, 
                  paste(params$species, "selectome.sub.bind.rds", sep="_")))

# Positive selection
saveRDS(selectome.enrichment,
        file.path(output.path, 
                  paste(params$species, "selectome.enrichment.rds", sep="_")))

saveRDS(selectome.chisq.test,
        file.path(output.path, 
                  paste(params$species, "selectome.chisq.test.rds", sep="_")))

# Permutation test
saveRDS(selectome.perm.sub, 
        file.path(output.path, 
                  paste(params$species, "selectome.perm.sub.rds", sep="_")))

saveRDS(selectome.perm.enrichment,
        file.path(output.path, 
                  paste(params$species, "selectome.perm.enrichment.rds", sep="_")))

saveRDS(perm.pvalue,
        file.path(output.path, 
                  paste(params$species, "perm.pvalue.rds", sep="_")))  

```

