---
title: "Expression Specificity"
author: "Bucao, Christabel Floi"
output: html_document
date: "`r Sys.Date()`"
params:
  input.rds.path: "../../results/01_Filter_Samples"
  output.rds.path: "../../results/10_Expression_Specificity"
  min.cpm: 1.0
  tau.cutoff: 0.30
  all.nonzero.matrix: FALSE
  randomize.mean.expr: FALSE
  set.seed: 12345
---

# Setup {.tabset}

## Parameters
```{r print-params}
print(params)
```

## Libraries
```{r import-libraries}
library(dplyr)
library(tibble)
library(tidyr)
```

## Functions
```{r source-functions}
source("../functions/filter_samples.R")
source("../functions/expression_specificity.R")
source("../functions/expression_specificity_plots.R")
source("../functions/helper_functions.R")
```

## Session Info
```{r print-session-info}
sessionInfo()
```

## Data
```{r import-filtered-samples}
# Load saved data from 01_Filter_Samples.Rmd
# Normalized across conditions
# Minimum replicates
species <- c("LOC","ELU","DRE")

filtered.samples <- list()

if (!params$all.nonzero.matrix) {
  filtered.samples$LOC <- readRDS(
    file.path(params$input.rds.path, "LOC_2_reps_data.qc2.cor.rds"))
  filtered.samples$ELU <- readRDS(
    file.path(params$input.rds.path, "ELU_1_reps_data.qc2.cor.rds"))
  filtered.samples$DRE <- readRDS(
    file.path(params$input.rds.path, "DRE_1_reps_data.qc2.cor.rds"))
} else {
  filtered.samples$LOC <- readRDS(
    file.path(params$input.rds.path, "LOC_2_reps_data.qc2.cor.nonzero.rds"))
  filtered.samples$ELU <- readRDS(
    file.path(params$input.rds.path, "ELU_1_reps_data.qc2.cor.nonzero.rds"))
  filtered.samples$DRE <- readRDS(
    file.path(params$input.rds.path, "DRE_1_reps_data.qc2.cor.nonzero.rds"))
}
```

# Main {.tabset}

## Expression specificity given all available conditions per species, except skin
```{r compute-tau-all-conditions-except-skin}
filtered.samples.exc.skin <- vector("list",3)
names(filtered.samples.exc.skin) <- species

for (sp in species) {
  # Remove skin
  filtered.samples.exc.skin[[sp]]$conditions <-
    filtered.samples[[sp]]$conditions[filtered.samples[[sp]]$conditions$Tissue!="skin",]
  
  filtered.samples.exc.skin[[sp]]$metadata <-
    filtered.samples[[sp]]$metadata[filtered.samples[[sp]]$metadata$Tissue!="skin",]
  
  filtered.samples.exc.skin[[sp]]$counts <-
    filtered.samples[[sp]]$counts[,!grepl("skin",colnames(filtered.samples[[sp]]$counts))]
  
  # Rename "gonads" to "ovary" or "testis" (if known)
  # They are considered separate anatomical entities for tissue specificity computations
  if ("Sex" %in% colnames(filtered.samples.exc.skin[[sp]]$metadata)) {
    filtered.samples.exc.skin[[sp]]$metadata <-
      split_gonads_df(filtered.samples.exc.skin[[sp]]$metadata)
    filtered.samples.exc.skin[[sp]]$counts <- 
      split_gonads_matrix(filtered.samples.exc.skin[[sp]]$counts)
  }
  
  # Repeat gene filtering and normalization after removal of conditions
  ## Filter lowly or non-expressed genes from raw read counts
  if (!params$all.nonzero.matrix) {
    filtered.samples.exc.skin[[sp]]$counts.gf <-
      filter_genes_by_cpm_per_condition(
        filtered.samples.exc.skin[[sp]]$counts, filtered.samples.exc.skin[[sp]]$metadata, min.cpm=params$min.cpm)    
  } else {
    # For the expression matrix without any zero counts, there is no need to repeat gene filtering
    filtered.samples.exc.skin[[sp]]$counts.gf <-
      filtered.samples[[sp]]$counts.gf[,!grepl("skin",colnames(filtered.samples[[sp]]$counts.gf))]
    
    if ("Sex" %in% colnames(filtered.samples.exc.skin[[sp]]$metadata)) {
      filtered.samples.exc.skin[[sp]]$counts.gf <-
        split_gonads_matrix(filtered.samples.exc.skin[[sp]]$counts.gf)      
    }
  }

  ## Normalize across all conditions
  filtered.samples.exc.skin[[sp]]$log2.tmm.cpm.gf <-
    transform_counts_to_tmm_cpm(
      filtered.samples.exc.skin[[sp]]$counts.gf, filtered.samples.exc.skin[[sp]]$metadata,
      log2=TRUE, shift.min=TRUE)
  
  # Expression specificity computations
  ## Compute mean expression per tissue
  filtered.samples.exc.skin[[sp]]$mean.expr <-
    mean_expr_per_tissue(filtered.samples.exc.skin[[sp]]$log2.tmm.cpm.gf,
                         filtered.samples.exc.skin[[sp]]$metadata)
  
  if (params$randomize.mean.expr==TRUE) {
    # Shuffle mean expression vector per gene
    filtered.samples.exc.skin[[sp]]$mean.expr <-
      permute_by_rows(filtered.samples.exc.skin[[sp]]$mean.expr, seed=params$set.seed)
  }
  
  ## Scale by maximum expression
  filtered.samples.exc.skin[[sp]]$scaled.expr <-
    scale_by_max_expr(filtered.samples.exc.skin[[sp]]$mean.expr)
  
  ## Compute tau from mean expression
  filtered.samples.exc.skin[[sp]]$tau <-
    compute_tau(filtered.samples.exc.skin[[sp]]$mean.expr)
}
```

```{r tau-summary-all-conditions-except-skin}
for (sp in species) {
  print(sp)
  print(summary(filtered.samples.exc.skin[[sp]]$tau))
}
```

### Plot - Tau distribution
```{r plot-tau-distribution-all-conditions-except-skin}
for (sp in species) {
  print(sp)
  print(levels(as.factor(filtered.samples.exc.skin[[sp]]$metadata$Tissue)))
  
  summary(filtered.samples.exc.skin[[sp]]$tau)
  
  if (!params$all.nonzero.matrix) {
      plot_tau_distribution(filtered.samples.exc.skin[[sp]]$tau, prob=seq(0,1,0.25), ylim=c(0.0,2.4))
  } else {
    plot_tau_distribution(filtered.samples.exc.skin[[sp]]$tau, prob=seq(0,1,0.25), ylim=c(0.0,3.6))    
  }
}
```


### Rank organs for each gene
```{r rank-organs-all-conditions-except-skin}
## For each gene, rank conditions by expression level
for (sp in species) {
  filtered.samples.exc.skin[[sp]]$expr.rank <-
    t(apply(-filtered.samples.exc.skin[[sp]]$mean.expr, MARGIN=1, rank, ties.method="max")) %>%
    as.data.frame() %>%
    tibble::rownames_to_column(var="GeneID")
  
  filtered.samples.exc.skin[[sp]]$expr.rank.pivot <-
    tidyr::pivot_longer(as.data.frame(filtered.samples.exc.skin[[sp]]$expr.rank),
                        cols=any_of(colnames(filtered.samples.exc.skin[[sp]]$mean.expr)), 
                        names_to="Tissue", values_to="ExpRank")
  
  filtered.samples.exc.skin[[sp]]$expr.rank.first <- filtered.samples.exc.skin[[sp]]$expr.rank.pivot %>%
    dplyr::filter(ExpRank==1) %>%
    dplyr::inner_join(data.frame("GeneID"=names(filtered.samples.exc.skin[[sp]]$tau),
                                 "tau"=filtered.samples.exc.skin[[sp]]$tau),
                      by="GeneID")
  colnames(filtered.samples.exc.skin[[sp]]$expr.rank.first) <- c("GeneID","PrimaryTissue","ExpRank","tau")
  filtered.samples.exc.skin[[sp]]$expr.rank.first <- 
    filtered.samples.exc.skin[[sp]]$expr.rank.first[,c("GeneID","PrimaryTissue","tau")]
}
```

#### Barplots - Number of genes with top expression in organ

##### No tau cutoff
```{r distribution-of-top-tissues-all-conditions-except-skin}
for (sp in species) {
  print(barplot_top_tissue(
    expr.rank.first=filtered.samples.exc.skin[[sp]]$expr.rank.first,
    apply.cutoff=FALSE,
    title=sp))
}
```

##### With tau cutoff
```{r distribution-of-top-tissues-tau-cutoff-all-conditions-except-skin}
for (sp in species) {
  print(barplot_top_tissue(
    expr.rank.first=filtered.samples.exc.skin[[sp]]$expr.rank.first,
    apply.cutoff=TRUE, min.tau=params$tau.cutoff,
    title=sp))
}
```


## Expression specificity given set of common conditions only
```{r compute-tau-common-conditions}
common.tissues <-
  intersect(filtered.samples$LOC$conditions$Tissue, filtered.samples$ELU$conditions$Tissue) %>%
  intersect(filtered.samples$DRE$conditions$Tissue)

filtered.samples.common <- vector("list",3)
names(filtered.samples.common) <- species

for (sp in species) {
  # Remove skin
  filtered.samples.common[[sp]]$conditions <-
    filtered.samples[[sp]]$conditions[filtered.samples[[sp]]$conditions$Tissue %in% common.tissues,]
  
  filtered.samples.common[[sp]]$metadata <-
    filtered.samples[[sp]]$metadata[filtered.samples[[sp]]$metadata$Tissue %in% common.tissues,]
  
  filtered.samples.common[[sp]]$counts <-
    filtered.samples[[sp]]$counts[,filtered.samples[[sp]]$metadata$Tissue %in% common.tissues]
  
  # Rename "gonads" to "ovary" or "testis" (if known)
  # They are considered separate anatomical entities for tissue specificity computations
  if ("Sex" %in% colnames(filtered.samples.common[[sp]]$metadata)) {
    filtered.samples.common[[sp]]$metadata <-
      split_gonads_df(filtered.samples.common[[sp]]$metadata)
    filtered.samples.common[[sp]]$counts <- 
      split_gonads_matrix(filtered.samples.common[[sp]]$counts)
  }
  
  # Repeat gene filtering and normalization after removal of conditions
  ## Filter lowly or non-expressed genes from raw read counts
  if (!params$all.nonzero.matrix) {
    filtered.samples.common[[sp]]$counts.gf <-
      filter_genes_by_cpm_per_condition(
        filtered.samples.common[[sp]]$counts, filtered.samples.common[[sp]]$metadata, min.cpm=params$min.cpm)    
  } else {
    # For the expression matrix without any zero counts, there is no need to repeat gene filtering
    filtered.samples.common[[sp]]$counts.gf <-
      filtered.samples[[sp]]$counts.gf[,filtered.samples[[sp]]$metadata$Tissue %in% common.tissues]
    
    if ("Sex" %in% colnames(filtered.samples.common[[sp]]$metadata)) {
      filtered.samples.common[[sp]]$counts.gf <-
        split_gonads_matrix(filtered.samples.common[[sp]]$counts.gf)      
    } 
  }

  ## Normalize across all conditions
  filtered.samples.common[[sp]]$log2.tmm.cpm.gf <-
    transform_counts_to_tmm_cpm(
      filtered.samples.common[[sp]]$counts.gf, filtered.samples.common[[sp]]$metadata,
      log2=TRUE, shift.min=TRUE)
  
  # Expression specificity computations
  ## Compute mean expression per tissue
  filtered.samples.common[[sp]]$mean.expr <-
    mean_expr_per_tissue(filtered.samples.common[[sp]]$log2.tmm.cpm.gf,
                         filtered.samples.common[[sp]]$metadata)
  
  if (params$randomize.mean.expr==TRUE) {
    # Shuffle mean expression vector per gene
    filtered.samples.common[[sp]]$mean.expr <-
      permute_by_rows(filtered.samples.common[[sp]]$mean.expr, seed=params$set.seed)
  }
  
  ## Scale by maximum expression
  filtered.samples.common[[sp]]$scaled.expr <-
    scale_by_max_expr(filtered.samples.common[[sp]]$mean.expr)
  
  ## Compute tau from mean expression
  filtered.samples.common[[sp]]$tau <-
    compute_tau(filtered.samples.common[[sp]]$mean.expr)
}
```

```{r tau-summary-common-conditions}
for (sp in species) {
  print(sp)
  print(summary(filtered.samples.exc.skin[[sp]]$tau))
}
```

### Plot - Tau distribution
```{r plot-tau-distribution-common-conditions}
for (sp in species) {
  print(sp)
  print(levels(as.factor(filtered.samples.common[[sp]]$metadata$Tissue)))
  
  summary(filtered.samples.exc.skin[[sp]]$tau)
  if (!params$all.nonzero.matrix) {
    plot_tau_distribution(filtered.samples.common[[sp]]$tau, prob=seq(0,1,0.25), ylim=c(0.0,2.4))    
  } else {
    plot_tau_distribution(filtered.samples.common[[sp]]$tau, prob=seq(0,1,0.25), ylim=c(0.0,3.6))    
  }
}
```


### Rank organs for each gene
```{r rank-organs-common-conditions}
## For each gene, rank conditions by expression level
for (sp in species) {
  filtered.samples.common[[sp]]$expr.rank <-
    t(apply(-filtered.samples.common[[sp]]$mean.expr, MARGIN=1, rank, ties.method="max")) %>%
    as.data.frame() %>%
    tibble::rownames_to_column(var="GeneID")
  
  filtered.samples.common[[sp]]$expr.rank.pivot <-
    tidyr::pivot_longer(as.data.frame(filtered.samples.common[[sp]]$expr.rank),
                        cols=any_of(colnames(filtered.samples.common[[sp]]$mean.expr)), 
                        names_to="Tissue", values_to="ExpRank")
  
  filtered.samples.common[[sp]]$expr.rank.first <- filtered.samples.common[[sp]]$expr.rank.pivot %>%
    dplyr::filter(ExpRank==1) %>%
    dplyr::inner_join(data.frame("GeneID"=names(filtered.samples.common[[sp]]$tau),
                                 "tau"=filtered.samples.common[[sp]]$tau),
                      by="GeneID")
  colnames(filtered.samples.common[[sp]]$expr.rank.first) <- c("GeneID","PrimaryTissue","ExpRank","tau")
  filtered.samples.common[[sp]]$expr.rank.first <- 
    filtered.samples.common[[sp]]$expr.rank.first[,c("GeneID","PrimaryTissue","tau")]
}
```

#### Barplots - Number of genes with top expression in organ

##### No tau cutoff
```{r distribution-of-top-tissues-common-conditions}
for (sp in species) {
  print(barplot_top_tissue(
    expr.rank.first=filtered.samples.common[[sp]]$expr.rank.first,
    apply.cutoff=FALSE,
    title=sp))
}
```


##### With tau cutoff
```{r distribution-of-top-tissues-tau-cutoff-common-conditions}
for (sp in species) {
  print(barplot_top_tissue(
    expr.rank.first=filtered.samples.common[[sp]]$expr.rank.first,
    apply.cutoff=TRUE, min.tau=params$tau.cutoff,
    title=sp))
}
```

# Save
```{r save-objects}
if (!dir.exists(params$output.rds.path)) { dir.create(params$output.rds.path, showWarnings=FALSE) }

# Separate per species
if (!params$all.nonzero.matrix) {
  ## All conditions within-species except skin
  if (!params$randomize.mean.expr) {
    for (sp in species) {
      saveRDS(filtered.samples.exc.skin[[sp]],
              file.path(params$output.rds.path, 
                        paste(sp, "filtered.samples.exc.skin.rds", sep="_")))    
    }    
  } else {
    for (sp in species) {
      saveRDS(filtered.samples.exc.skin[[sp]],
              file.path(params$output.rds.path,
                        paste(sp, "filtered.samples.exc.skin", paste0("rnd",params$set.seed, ".rds"), sep="_")))
    }
  }

  
  ## All common conditions
  if (!params$randomize.mean.expr) {
    for (sp in species) {
      saveRDS(filtered.samples.common[[sp]],
              file.path(params$output.rds.path, 
                        paste(sp, "filtered.samples.common.rds", sep="_")))    
    }    
  } else {
    for (sp in species) {
      saveRDS(filtered.samples.exc.skin[[sp]],
              file.path(params$output.rds.path,
                        paste(sp, "filtered.samples.common", paste0("rnd",params$set.seed, ".rds"), sep="_")))
    }    
  }

} else {
  ## All conditions within-species except skin
  if (!params$randomize.mean.expr) {
    for (sp in species) {
      saveRDS(filtered.samples.exc.skin[[sp]],
              file.path(params$output.rds.path, 
                        paste(sp, "filtered.samples.exc.skin.nonzero.rds", sep="_")))    
    }     
  } else {
    for (sp in species) {
      saveRDS(filtered.samples.exc.skin[[sp]],
              file.path(params$output.rds.path,
                        paste(sp, "filtered.samples.exc.skin.nonzero", paste0("rnd",params$set.seed, ".rds"), sep="_")))
    }    
  }

  
  ## All common conditions
  if (!params$randomize.mean.expr) {
    for (sp in species) {
      saveRDS(filtered.samples.common[[sp]],
              file.path(params$output.rds.path, 
                        paste(sp, "filtered.samples.common.nonzero.rds", sep="_")))    
    }    
  } else {
    for (sp in species) {
      saveRDS(filtered.samples.exc.skin[[sp]],
              file.path(params$output.rds.path,
                        paste(sp, "filtered.samples.common.nonzero", paste0("rnd",params$set.seed, ".rds"), sep="_")))
    }      
  }

}
```

