---
title: "Variability of organ-biased genes (*`r params$species.name`*)"
author: "Bucao, Christabel Floi"
output: html_document
date: "`r Sys.Date()`"
params:
  input.variability.rds.path: "../../results/03_Variability_Plots"
  input.tau.rds.path: "../../results/10_Expression_Specificity"
  output.rds.path: "../../results/12_Organ_Bias"
  species: "DRE"
  species.name: "Danio rerio"
  tau.cutoff: 0.30
  all.nonzero.matrix: FALSE
  randomized.mean.expr: FALSE
  seed.run: 12345
---

# Setup {.tabset}

## Parameters
```{r print-params}
print(params)
```

## Libraries
```{r import-libraries}
library(dplyr)
```

## Functions
```{r source-functions}
source("../functions/helper_functions.R")
source("../functions/expression_specificity_plots.R")
source("../functions/organ_bias.R")
source("../functions/organ_bias_plots.R")
#source("../functions/annotate_boxplots.R")
```

## Session Info
```{r print-session-info}
sessionInfo()
```

## Data

### Organ expression specificity (tau)
```{r import-tau}
# Load expression specificity computations from 10_Expression_Specificity.Rmd
## Conditions available in all species
if (!params$all.nonzero.matrix) {
  if (!params$randomized.mean.expr) {
    filtered.samples <- readRDS(
      file.path(params$input.tau.rds.path,
                paste(params$species, "filtered.samples.common.rds", sep="_")))     
  } else {
    filtered.samples <- readRDS(
      file.path(params$input.tau.rds.path,
                paste(params$species, "filtered.samples.common", 
                      paste0("rnd", params$seed.run, ".rds"), sep="_"))) 
  }
 
} else {
  if (!params$randomized.mean.expr) {
    filtered.samples <- readRDS(
      file.path(params$input.tau.rds.path,
                paste(params$species, "filtered.samples.common.nonzero.rds", sep="_")))      
  } else {
    filtered.samples <- readRDS(
      file.path(params$input.tau.rds.path,
                paste(params$species, "filtered.samples.common.nonzero", 
                      paste0("rnd", params$seed.run, ".rds"), sep="_")))     
  }
}

filtered.samples$conditions <- split_gonads_df(filtered.samples$conditions)
expr.rank.first <- filtered.samples$expr.rank.first
```

### Expression variability
```{r import-variability}
# Load processed expression variability estimates from 03_Variability_Plots.Rmd
if (!params$all.nonzero.matrix) {
  expr.var <- readRDS(
    file.path(params$input.variability.rds.path,
              paste(params$species, "jack.resid.cv.bind.rds", sep="_"))) 
} else {
  expr.var <- readRDS(
    file.path(params$input.variability.rds.path,
              paste(params$species, "jack.resid.cv.bind.nonzero.rds", sep="_"))) 
}

expr.var <- split_gonads_df(expr.var)

# Retain only conditions included in expression specificity computations
expr.var.common <- expr.var %>%
  dplyr::filter(Tissue %in% filtered.samples$conditions$Tissue)
```

# Main {.tabset}

## Expression specificity (tau)

### Plot - Tau distribution

#### All genes with tau
```{r plot-tau-distribution-all-genes}
if (!params$all.nonzero.matrix) {
  plot_tau_distribution(filtered.samples$tau,
                        tau.low=params$tau.cutoff, tau.high=params$tau.cutoff,
                        ylim=c(0.0,1.5))  
} else {
  plot_tau_distribution(filtered.samples$tau,
                        tau.low=params$tau.cutoff, tau.high=params$tau.cutoff,
                        ylim=c(0.0,3.6))    
}
```

#### With tau and expression variability data
```{r plot-tau-distribution-genes-with-variability-data}
tau.with.var <- 
  filtered.samples$tau[names(filtered.samples$tau) %in% levels(as.factor(expr.var.common$GeneID))]

if (!params$all.nonzero.matrix) {
  plot_tau_distribution(tau.with.var,
                        tau.low=params$tau.cutoff, tau.high=params$tau.cutoff,
                        ylim=c(0.0,1.5))    
} else {
  plot_tau_distribution(tau.with.var,
                        tau.low=params$tau.cutoff, tau.high=params$tau.cutoff,
                        ylim=c(0.0,3.6))    
}
```

## Variability as a function of tau

### Reorder organs for plotting
```{r reorder-conditions}
expr.var.rank.common <- expr.var.common %>%
  dplyr::inner_join(expr.rank.first, by="GeneID")

if ("Sex" %in% colnames(filtered.samples$metadata)) {
  levels.order <- 
    c("brain","eye","intestine","liver",
      "heart","muscle","gills","pectoral_fin",
      "ovary","testis")
} else {
  levels.order <-
    c("brain","eye","intestine","liver",
      "heart","muscle","gills","pectoral_fin","gonads")  
}

# Exclude missing conditions from factor levels
levels.exclude <- setdiff(levels.order, levels(as.factor(expr.var.rank.common$Tissue)))

# Reorder levels for plotting
expr.var.rank.common$Tissue <-
  factor(expr.var.rank.common$Tissue, levels=levels.order, exclude=levels.exclude)
```


### Density plots

#### Variability rank vs. mean expression
```{r density-plot-variability-vs-mean, fig.width=20, fig.height=10}
density_plot_variability_vs_mean(expr.var.rank.common, title=params$species.name)
```

#### Mean expression vs. tau
```{r density-plot-mean-vs-tau, fig.width=20, fig.height=10}
density_plot_mean_vs_tau(expr.var.rank.common, title=params$species.name)
```

#### Residual variation vs. tau
```{r density-plot-residual-variation-vs-tau, fig.width=20, fig.height=10}
density_plot_resid_var_vs_tau(expr.var.rank.common, title=params$species.name)
```

#### Variability rank vs. tau
```{r density-plot-variability-vs-tau, fig.width=20, fig.height=10}
density_plot_variability_vs_tau(expr.var.rank.common, title=params$species.name)
```

### Boxplots by tau quantiles

#### Split by tau quantiles
```{r tau-quantiles}
tau.with.var.quantiles <- quantile(tau.with.var)

expr.var.rank.common$Quantile <-
  apply(expr.var.rank.common, MARGIN=1, function(r) {
    if (r[["tau"]] <= tau.with.var.quantiles["25%"]) { "Q1" }
    else if (r[["tau"]] <= tau.with.var.quantiles["50%"]) { "Q2" }
    else if (r[["tau"]] <= tau.with.var.quantiles["75%"]) { "Q3" }
    else { "Q4" }
  })

print(tau.with.var.quantiles)  
```


#### Mean expression
```{r boxplot-mean-expression-tau-quantiles, fig.width=20, fig.height=10}
boxplot_mean_by_tau_quantiles(expr.var.rank.common, title=species.name)
```

#### Residual variation
```{r boxplot-resid-var-tau-quantiles, fig.width=20, fig.height=10}
boxplot_resid_var_by_tau_quantiles(expr.var.rank.common, title=species.name)
```

#### Variability rank
```{r boxplot-variability-tau-quantiles, fig.width=20, fig.height=10}
boxplot_variability_by_tau_quantiles(expr.var.rank.common, title=species.name)
```

## Variability of organ-biased genes

### Classify broadly expressed and organ-biased genes (with subsets)
```{r classify-genes-by-organ-bias}
Tissue <- as.factor(expr.var.rank.common$Tissue)

# Per organ, classify genes by organ bias
if ("Sex" %in% colnames(expr.var.rank.common)) {
  expr.var.bias.by.condition <- vector(mode="list", length=2)
  names(expr.var.bias.by.condition) <- c("F","M")
  
  for (s in c("F","M")) {
    expr.var.bias.by.condition[[s]] <- vector(mode="list", length=length(levels(Tissue)))
    names(expr.var.bias.by.condition[[s]]) <- levels(Tissue)
    
    for (t in levels(Tissue)) {
      expr.var.bias.by.condition[[s]][[t]] <- expr.var.rank.common %>%
        dplyr::filter(Sex==s) %>%
        dplyr::filter(Tissue!=t) %>%
        classify_genes_by_organ_bias(tau.cutoff=params$tau.cutoff)
      
      # Reorder factor levels
      expr.var.bias.by.condition[[s]][[t]]$TissueBias <-
        factor(expr.var.bias.by.condition[[s]][[t]]$TissueBias, levels=c("broad","focal",t,"other"))
    }
  }
} else {
  expr.var.bias.by.condition <- vector(mode="list", length=length(levels(Tissue)))
  names(expr.var.bias.by.condition) <- levels(Tissue)  
  
  for (t in levels(Tissue)) {
    expr.var.bias.by.condition[[t]] <- expr.var.rank.common %>%
      dplyr::filter(Tissue!=t) %>%
      classify_genes_by_organ_bias(tau.cutoff=params$tau.cutoff)
  
  # Reorder factor levels
    expr.var.bias.by.condition[[t]]$TissueBias <-
      factor(expr.var.bias.by.condition[[t]]$TissueBias, levels=c("broad","focal",t,"other"))    
  }
}
```

### Boxplots by organ bias {.tabset}

#### Mean expression
```{r boxplot-mean-expression-organ-bias, fig.width=18, fig.height=10}
if ("Sex" %in% colnames(expr.var.rank.common)) {
  for (s in c("F","M")) {
    for (t in levels(Tissue)) {
      print(boxplot_mean_by_organ_bias(expr.var.bias.by.condition[[s]][[t]], 
                                       species=params$species, tissue=t, sex=s))
    }
  }
} else {
  for (t in levels(Tissue)) { 
    print(boxplot_mean_by_organ_bias(expr.var.bias.by.condition[[t]],
                                     species=params$species, tissue=t))    
    }
}
```

#### Residual variation
```{r boxplot-resid-var-organ-bias, fig.width=18, fig.height=10}
if ("Sex" %in% colnames(expr.var.rank.common)) {
  for (s in c("F","M")) {
    for (t in levels(Tissue)) {
      print(boxplot_resid_var_by_organ_bias(expr.var.bias.by.condition[[s]][[t]], 
                                            species=params$species, tissue=t, sex=s))
    }
  }
} else {
  for (t in levels(Tissue)) {
    print(boxplot_resid_var_by_organ_bias(expr.var.bias.by.condition[[t]],
                                          species=params$species, tissue=t))    
  }
}
```

#### Variability rank
```{r boxplot-variability-organ-bias, fig.width=18, fig.height=10}
if ("Sex" %in% colnames(expr.var.rank.common)) {
  for (s in c("F","M")) {
    for (t in levels(Tissue)) {
      print(boxplot_variability_by_organ_bias(expr.var.bias.by.condition[[s]][[t]], 
                                              species=params$species, tissue=t, sex=s))
    }
  }
} else {
  for (t in levels(Tissue)) {
    print(boxplot_variability_by_organ_bias(expr.var.bias.by.condition[[t]],
                                          species=params$species, tissue=t))    
  }

}
```

#### Supplementary

##### Brain-biased genes
```{r boxplot-variability-brain-bias, fig.width=20, fig.height=10}
bias <- "brain"

if ("Sex" %in% colnames(expr.var.rank.common)) {
  expr.var.brain.bias <- 
    dplyr::bind_rows(expr.var.bias.by.condition[["F"]][[bias]], expr.var.bias.by.condition[["M"]][[bias]])
  
} else {
  expr.var.brain.bias <- expr.var.bias.by.condition[[bias]]
}

# Same, just different colors
if (params$randomized.mean.expr) {
  boxplot_variability_example(expr.var.brain.bias, tissue=bias, species=params$species.name, color="#FFFAFA")
} else {
  boxplot_variability_example(expr.var.brain.bias, tissue=bias, species=params$species.name, color="#67005D")
  boxplot_variability_example(expr.var.brain.bias, tissue=bias, species=params$species.name, color="#00BFC4")    
}
```

##### Liver-biased genes
```{r boxplot-variability-liver-bias, fig.width=20, fig.height=10}
bias <- "liver"

if ("Sex" %in% colnames(expr.var.rank.common)) {
  expr.var.liver.bias <- 
    dplyr::bind_rows(expr.var.bias.by.condition[["F"]][[bias]], expr.var.bias.by.condition[["M"]][[bias]])
  
} else {
  expr.var.liver.bias <- expr.var.bias.by.condition[[bias]]
}

# Same, just different colors
if (params$randomized.mean.expr) {
  boxplot_variability_example(expr.var.liver.bias, tissue=bias, species=params$species.name, color="#FFFAFA")  
} else {
  boxplot_variability_example(expr.var.liver.bias, tissue=bias, species=params$species.name, color="#FF8A00")
  boxplot_variability_example(expr.var.liver.bias, tissue=bias, species=params$species.name, color="#F8766D")  
}
```

##### Ovary- or gonads-biased genes
```{r boxplot-variability-ovary-bias, fig.width=20, fig.height=10}
if ("Sex" %in% colnames(expr.var.rank.common)) {
  bias <- "ovary"
  expr.var.ovary.bias <- 
    dplyr::bind_rows(expr.var.bias.by.condition[["F"]][[bias]], expr.var.bias.by.condition[["M"]][[bias]])
  
  if (params$randomized.mean.expr) {
    boxplot_variability_example(expr.var.ovary.bias, tissue=bias, species=params$species.name, color="#FFFAFA")
  } else {
    boxplot_variability_example(expr.var.ovary.bias, tissue=bias, species=params$species.name, color="#67005D")
    boxplot_variability_example(expr.var.ovary.bias, tissue=bias, species=params$species.name, color="#00BFC4")    
  }
  
} else {
  bias <- "gonads"
  expr.var.gonads.bias <- expr.var.bias.by.condition[[bias]]
  
  if (params$randomized.mean.expr) {
    boxplot_variability_example(expr.var.gonads.bias, tissue=bias, species=params$species.name, color="#FFFAFA")
  } else {
    boxplot_variability_example(expr.var.gonads.bias, tissue=bias, species=params$species.name, color="#67005D")
    boxplot_variability_example(expr.var.gonads.bias, tissue=bias, species=params$species.name, color="#00BFC4")    
  }
}
```

#### Examples

##### Brain-biased genes expressed in the pectoral fin
```{r brain-biased-in-pectoral-fin, fig.width=3.5, fig.height=5}
bias <- "brain"
focal <- "pectoral_fin"
  
if ("Sex" %in% colnames(expr.var.rank.common)) {
  sex <- "F"
  brain.biased.in.pecfin <- 
    expr.var.bias.by.condition[[sex]][[bias]] %>%
    dplyr::filter(Tissue==focal)

} else {
  brain.biased.in.pecfin <- 
    expr.var.bias.by.condition[[bias]] %>%
    dplyr::filter(Tissue==focal)
  
}

if (params$randomized.mean.expr) {
  print(boxplot_variability_example(
    brain.biased.in.pecfin, species=params$species.name, tissue=bias, ylab="Variability rank", color="#FFFAFA"))    
} else {
  print(boxplot_variability_example(
    brain.biased.in.pecfin, species=params$species.name, tissue=bias, ylab="Variability rank", color="#67005D"))
    
  print(boxplot_variability_example(
    brain.biased.in.pecfin, species=params$species.name, tissue=bias, ylab="Variability rank", color="#00BFC4"))    
}
```


##### Liver-biased genes expressed in the intestine
```{r liver-biased-in-intestine, fig.width=3.5, fig.height=5}
bias <- "liver"
focal <- "intestine"
  
if ("Sex" %in% colnames(expr.var.rank.common)) {
  sex <- "F"
  liver.biased.in.intestine <- 
    expr.var.bias.by.condition[[sex]][[bias]] %>%
    dplyr::filter(Tissue==focal)

} else {
  liver.biased.in.intestine <- 
    expr.var.bias.by.condition[[bias]] %>%
    dplyr::filter(Tissue==focal)
    
}

if (params$randomized.mean.expr) {
  print(boxplot_variability_example(
    liver.biased.in.intestine, species=params$species.name, tissue=bias, ylab="Variability rank", color="#FFFAFA"))    
} else {
  print(boxplot_variability_example(
    liver.biased.in.intestine, species=params$species.name, tissue=bias, ylab="Variability rank", color="#FF8A00"))
  
  print(boxplot_variability_example(
    liver.biased.in.intestine, species=params$species.name, tissue=bias, ylab="Variability rank", color="#F8766D"))    
}
```


##### Ovary-biased (or gonads-biased) genes expressed in the eye
```{r gonads-biased-in-eye, fig.width=3.5, fig.height=5}
focal <- "eye"
  
if ("Sex" %in% colnames(expr.var.rank.common)) {
  bias <- "ovary"
  sex <- "F"
  ovary.biased.in.eye <- 
    expr.var.bias.by.condition[[sex]][[bias]] %>%
    dplyr::filter(Tissue==focal)
  
  if (params$randomized.mean.expr) {
    print(boxplot_variability_example(
      ovary.biased.in.eye, species=params$species.name, tissue=bias, ylab="Variability rank", color="#FFFAFA"))      
  } else {
    print(boxplot_variability_example(
      ovary.biased.in.eye, species=params$species.name, tissue=bias, ylab="Variability rank", color="#67005D"))
  
    print(boxplot_variability_example(
      ovary.biased.in.eye, species=params$species.name, tissue=bias, ylab="Variability rank", color="#00BFC4"))     
  }
  
} else {
  bias <- "gonads"
  gonads.biased.in.eye <- 
    expr.var.bias.by.condition[[bias]] %>%
    dplyr::filter(Tissue==focal)

  if (params$randomized.mean.expr) {
    print(boxplot_variability_example(
      gonads.biased.in.eye, species=params$species.name, tissue=bias, ylab="Variability rank", color="#FFFAFA"))  
  } else {
    print(boxplot_variability_example(
      gonads.biased.in.eye, species=params$species.name, tissue=bias, ylab="Variability rank", color="#67005D"))
  
    print(boxplot_variability_example(
      gonads.biased.in.eye, species=params$species.name, tissue=bias, ylab="Variability rank", color="#00BFC4"))  
  }
}
```

### Multiple pairwise comparisons {.tabset}

#### Description

Variability of the set of _x_-biased genes expressed in organ _i_ (not _x_) vs. variability of other organ-biased genes in organ _i_ (excluding _i_-biased genes)

e.g., variability of brain-biased genes expressed in the ovary vs. variability of other organ-biased genes expressed in the ovary (excluding ovary-biased genes)

#### Exclude broadly expressed and focal-biased genes
```{r exclude-broad-and-focal-biased-genes}
if ("Sex" %in% colnames(expr.var.rank.common)) {
  expr.var.bias.comparisons <-
    lapply(c("F"="F","M"="M"), function(s) {
      dplyr::bind_rows(expr.var.bias.by.condition[[s]], .id="Bias") %>%
        dplyr::filter(!(TissueBias %in% c("broad","focal")))
      }) %>%
    dplyr::bind_rows()  
} else {
  expr.var.bias.comparisons <- expr.var.bias.by.condition %>%
    dplyr::bind_rows(.id="Bias") %>%
    dplyr::filter(!(TissueBias %in% c("broad","focal")))
}
```

#### Mean expression

##### Effect size
```{r compute-effect-size-mean-expression}
comparisons.mean.expr <- list()

comparisons.mean.expr$effect.size <- 
  compute_glass_delta_effect_size(expr.var.bias.comparisons, value="mean")
```

##### Pairwise Wilcoxon test
```{r pairwise-comparisons-mean-expression}
# Complete
comparisons.mean.expr$wilcox.complete <- 
  compute_pairwise_comparisons(
    comparisons.df=expr.var.bias.comparisons,
    effect.size.df=comparisons.mean.expr$effect.size,
    value="mean", apply.pvalue.cutoff=FALSE,
    order.levels=levels(Tissue)
  )

# Keep significant observations only
comparisons.mean.expr$wilcox.signif <- 
  compute_pairwise_comparisons(
    comparisons.df=expr.var.bias.comparisons, 
    effect.size.df=comparisons.mean.expr$effect.size,
    value="mean", apply.pvalue.cutoff=TRUE, adj.p=0.05, 
    order.levels=levels(Tissue))
```


##### Plot - Heatmap
```{r heatmap-pairwise-comparisons-mean-expression, fig.height=10, fig.width=14}
heatmap_effect_size_pairwise_comparisons_blue2red(
  comparisons.mean.expr$wilcox.complete, species=params$species.name, value="mean", order.levels=levels.order)

heatmap_effect_size_pairwise_comparisons(
  comparisons.mean.expr$wilcox.signif, species=params$species.name, value="mean", order.levels=levels.order)
```

#### Log-coefficient of variation

##### Effect size
```{r compute-effect-size-log2cv}
comparisons.log2cv <- list()

comparisons.log2cv$effect.size <- 
  compute_glass_delta_effect_size(expr.var.bias.comparisons, value="log2cv")
```

##### Pairwise Wilcoxon test
```{r pairwise-comparisons-log2cv}
# Complete
comparisons.log2cv$wilcox.complete <- 
  compute_pairwise_comparisons(
    comparisons.df=expr.var.bias.comparisons,
    effect.size.df=comparisons.log2cv$effect.size,
    value="log2cv", apply.pvalue.cutoff=FALSE,
    order.levels=levels(Tissue)
  )

# Keep significant observations only
comparisons.log2cv$wilcox.signif <- 
  compute_pairwise_comparisons(
    comparisons.df=expr.var.bias.comparisons, 
    effect.size.df=comparisons.log2cv$effect.size,
    value="log2cv", apply.pvalue.cutoff=TRUE, adj.p=0.05, 
    order.levels=levels(Tissue))
```

##### Plot - Heatmap
```{r heatmap-pairwise-comparisons-log2cv, fig.height=10, fig.width=14}
heatmap_effect_size_pairwise_comparisons_blue2red(
  comparisons.log2cv$wilcox.complete, species=params$species.name, value="log2cv", order.levels=levels.order)

heatmap_effect_size_pairwise_comparisons(
  comparisons.log2cv$wilcox.signif, species=params$species.name, value="log2cv", order.levels=levels.order)
```

#### Residual variation

##### Effect size
```{r compute-effect-size-resid-var}
comparisons.resid.var <- list()

comparisons.resid.var$effect.size <- 
  compute_glass_delta_effect_size(expr.var.bias.comparisons, value="residual_variation")
```

##### Pairwise Wilcoxon test
```{r pairwise-comparisons-resid-var}
# Complete
comparisons.resid.var$wilcox.complete <- 
  compute_pairwise_comparisons(
    comparisons.df=expr.var.bias.comparisons,
    effect.size.df=comparisons.resid.var$effect.size,
    value="residual_variation", apply.pvalue.cutoff=FALSE,
    order.levels=levels(Tissue)
  )

# Keep significant observations only
comparisons.resid.var$wilcox.signif <- 
  compute_pairwise_comparisons(
    comparisons.df=expr.var.bias.comparisons, 
    effect.size.df=comparisons.resid.var$effect.size,
    value="residual_variation", apply.pvalue.cutoff=TRUE, adj.p=0.05, 
    order.levels=levels(Tissue))
```

##### Plot - Heatmap
```{r heatmap-pairwise-comparisons-resid-var, fig.height=10, fig.width=14}
heatmap_effect_size_pairwise_comparisons_blue2red(
  comparisons.resid.var$wilcox.complete, species=params$species.name, value="residual_variation", order.levels=levels.order)

heatmap_effect_size_pairwise_comparisons(
  comparisons.resid.var$wilcox.signif, species=params$species.name, value="residual_variation", order.levels=levels.order)
```

#### Variability rank

##### Effect size
```{r compute-effect-size-variability}
comparisons.expr.var <- list()

comparisons.expr.var$effect.size <- 
  compute_glass_delta_effect_size(expr.var.bias.comparisons, value="variability")
```

##### Pairwise Wilcoxon test
```{r pairwise-comparisons-variability}
# Complete
comparisons.expr.var$wilcox.complete <- 
  compute_pairwise_comparisons(
    comparisons.df=expr.var.bias.comparisons,
    effect.size.df=comparisons.expr.var$effect.size,
    value="variability", apply.pvalue.cutoff=FALSE,
    order.levels=levels(Tissue)
  )

# Keep significant observations only
comparisons.expr.var$wilcox.signif <- 
  compute_pairwise_comparisons(
    comparisons.df=expr.var.bias.comparisons, 
    effect.size.df=comparisons.expr.var$effect.size,
    value="variability", apply.pvalue.cutoff=TRUE, adj.p=0.05, 
    order.levels=levels(Tissue))
```


##### Plot - Heatmap
```{r heatmap-pairwise-comparisons-variability, fig.height=10, fig.width=14}
heatmap_effect_size_pairwise_comparisons_blue2red(
  comparisons.expr.var$wilcox.complete, species=params$species.name, value="variability", order.levels=levels.order)

heatmap_effect_size_pairwise_comparisons(
  comparisons.expr.var$wilcox.signif, species=params$species.name, value="variability", order.levels=levels.order)
```

# Save

## Observed
```{r save-objects-1, eval=!params$randomized.mean.expr}
if (!dir.exists(params$output.rds.path)) { dir.create(params$output.rds.path, showWarnings=FALSE) }

if (!params$all.nonzero.matrix) {
  saveRDS(expr.var.rank.common,
          file.path(params$output.rds.path, 
                    paste(params$species, "expr.var.rank.common.rds", sep="_")))
  
  saveRDS(expr.var.bias.by.condition,
          file.path(params$output.rds.path, 
                    paste(params$species, "expr.var.bias.by.condition.rds", sep="_")))
  
  saveRDS(expr.var.brain.bias,
          file.path(params$output.rds.path, 
                    paste(params$species, "expr.var.brain.bias.rds", sep="_")))
  
  saveRDS(expr.var.liver.bias,
          file.path(params$output.rds.path, 
                    paste(params$species, "expr.var.liver.bias.rds", sep="_")))
  
  saveRDS(expr.var.bias.comparisons,
          file.path(params$output.rds.path, 
                    paste(params$species, "expr.var.bias.comparisons.rds", sep="_")))
  
  saveRDS(comparisons.mean.expr,
          file.path(params$output.rds.path, 
                    paste(params$species, "comparisons.mean.expr.rds", sep="_")))
  
  saveRDS(comparisons.log2cv,
          file.path(params$output.rds.path, 
                    paste(params$species, "comparisons.log2cv.rds", sep="_")))
  
  saveRDS(comparisons.resid.var,
          file.path(params$output.rds.path, 
                    paste(params$species, "comparisons.resid.var.rds", sep="_")))

  saveRDS(comparisons.expr.var,
          file.path(params$output.rds.path, 
                    paste(params$species, "comparisons.expr.var.rds", sep="_")))
  
} else {
  saveRDS(expr.var.rank.common,
          file.path(params$output.rds.path, 
                    paste(params$species, "expr.var.rank.common.nonzero.rds", sep="_")))
  
  saveRDS(expr.var.bias.by.condition,
          file.path(params$output.rds.path, 
                    paste(params$species, "expr.var.bias.by.condition.nonzero.rds", sep="_")))
  
  saveRDS(expr.var.brain.bias,
          file.path(params$output.rds.path, 
                    paste(params$species, "expr.var.brain.bias.nonzero.rds", sep="_")))
  
  saveRDS(expr.var.liver.bias,
          file.path(params$output.rds.path, 
                    paste(params$species, "expr.var.liver.bias.nonzero.rds", sep="_")))
  
  saveRDS(expr.var.bias.comparisons,
          file.path(params$output.rds.path, 
                    paste(params$species, "expr.var.bias.comparisons.nonzero.rds", sep="_")))
  
  saveRDS(comparisons.mean.expr,
          file.path(params$output.rds.path, 
                    paste(params$species, "comparisons.mean.expr.nonzero.rds", sep="_")))
  
  saveRDS(comparisons.log2cv,
          file.path(params$output.rds.path, 
                    paste(params$species, "comparisons.log2cv.nonzero.rds", sep="_")))
  
  saveRDS(comparisons.resid.var,
          file.path(params$output.rds.path, 
                    paste(params$species, "comparisons.resid.var.nonzero.rds", sep="_")))

  saveRDS(comparisons.expr.var,
          file.path(params$output.rds.path, 
                    paste(params$species, "comparisons.expr.var.nonzero.rds", sep="_")))
}
```

## Randomized mean expression
```{r save-objects-2, eval=params$randomized.mean.expr}
if (!dir.exists(params$output.rds.path)) { dir.create(params$output.rds.path, showWarnings=FALSE) }

if (!params$all.nonzero.matrix) {
  saveRDS(expr.var.rank.common,
          file.path(params$output.rds.path, 
                    paste(params$species, "expr.var.rank.common", 
                          paste0("rnd", params$seed.run, ".rds"), sep="_")))

  saveRDS(expr.var.bias.by.condition,
          file.path(params$output.rds.path, 
                    paste(params$species, "expr.var.bias.by.condition", 
                          paste0("rnd", params$seed.run, ".rds"), sep="_")))
  
  saveRDS(expr.var.brain.bias,
          file.path(params$output.rds.path, 
                    paste(params$species, "expr.var.brain.bias", 
                          paste0("rnd", params$seed.run, ".rds"), sep="_")))
  
  saveRDS(expr.var.liver.bias,
          file.path(params$output.rds.path, 
                    paste(params$species, "expr.var.liver.bias", 
                          paste0("rnd", params$seed.run, ".rds"), sep="_")))
  
  saveRDS(expr.var.bias.comparisons,
          file.path(params$output.rds.path, 
                    paste(params$species, "expr.var.bias.comparisons", 
                          paste0("rnd", params$seed.run, ".rds"), sep="_")))
  
  saveRDS(comparisons.mean.expr,
          file.path(params$output.rds.path, 
                    paste(params$species, "comparisons.mean.expr", 
                          paste0("rnd", params$seed.run, ".rds"), sep="_")))
  
  saveRDS(comparisons.log2cv,
          file.path(params$output.rds.path, 
                    paste(params$species, "comparisons.log2cv", 
                          paste0("rnd", params$seed.run, ".rds"), sep="_")))
  
  saveRDS(comparisons.resid.var,
          file.path(params$output.rds.path, 
                    paste(params$species, "comparisons.resid.var", 
                          paste0("rnd", params$seed.run, ".rds"), sep="_")))

  saveRDS(comparisons.expr.var,
          file.path(params$output.rds.path, 
                    paste(params$species, "comparisons.expr.var", 
                          paste0("rnd", params$seed.run, ".rds"), sep="_")))
  
} else {
  saveRDS(expr.var.rank.common,
          file.path(params$output.rds.path, 
                    paste(params$species, "expr.var.rank.common.nonzero", 
                          paste0("rnd", params$seed.run, ".rds"), sep="_")))

  saveRDS(expr.var.bias.by.condition,
          file.path(params$output.rds.path, 
                    paste(params$species, "expr.var.bias.by.condition.nonzero", 
                          paste0("rnd", params$seed.run, ".rds"), sep="_")))
  
  saveRDS(expr.var.brain.bias,
          file.path(params$output.rds.path, 
                    paste(params$species, "expr.var.brain.bias.nonzero", 
                          paste0("rnd", params$seed.run, ".rds"), sep="_")))
  
  saveRDS(expr.var.liver.bias,
          file.path(params$output.rds.path, 
                    paste(params$species, "expr.var.liver.bias.nonzero", 
                          paste0("rnd", params$seed.run, ".rds"), sep="_")))
  
  saveRDS(expr.var.bias.comparisons,
          file.path(params$output.rds.path, 
                    paste(params$species, "expr.var.bias.comparisons.nonzero", 
                          paste0("rnd", params$seed.run, ".rds"), sep="_")))
  
  saveRDS(comparisons.mean.expr,
          file.path(params$output.rds.path, 
                    paste(params$species, "comparisons.mean.expr.nonzero", 
                          paste0("rnd", params$seed.run, ".rds"), sep="_")))
  
  saveRDS(comparisons.log2cv,
          file.path(params$output.rds.path, 
                    paste(params$species, "comparisons.log2cv.nonzero", 
                          paste0("rnd", params$seed.run, ".rds"), sep="_")))
  
  saveRDS(comparisons.resid.var,
          file.path(params$output.rds.path, 
                    paste(params$species, "comparisons.resid.var.nonzero", 
                          paste0("rnd", params$seed.run, ".rds"), sep="_")))

  saveRDS(comparisons.expr.var,
          file.path(params$output.rds.path, 
                    paste(params$species, "comparisons.expr.var.nonzero", 
                          paste0("rnd", params$seed.run, ".rds"), sep="_")))
}
```

