---
title: "Variability of organ-biased genes (*`r params$species.name`*)"
author: "Bucao, Christabel Floi"
output: html_document
date: "`r Sys.Date()`"
params:
  input.variability.rds.path: "../../results/02b_Variability_NoJack"
  input.tau.rds.path: "../../results/10_Expression_Specificity"
  output.rds.path: "../../results/12b_Organ_Bias_NoJack"
  species: "DRE"
  species.name: "Danio rerio"
  tau.cutoff: 0.30
  all.nonzero.matrix: FALSE
  randomized.mean.expr: FALSE
  qc1.only: FALSE
  combat: TRUE
  seed.run: 12345
---

# Setup {.tabset}

## Parameters
```{r print-params}
print(params)
```

## Libraries
```{r import-libraries}
library(dplyr)
```

## Functions
```{r source-functions}
source("../functions/helper_functions.R")
source("../functions/expression_specificity_plots.R")
source("../functions/organ_bias.R")
source("../functions/organ_bias_plots.R")
#source("../functions/annotate_boxplots.R")
```

## Session Info
```{r print-session-info}
sessionInfo()
```

## Data

### Organ expression specificity (tau)
```{r import-tau}
if (!params$combat) {
  if (!params$qc1.only) {
    input.tau.path <- file.path(params$input.tau.rds.path, "no_combat", "full_qc")    
  } else {
    input.tau.path <- file.path(params$input.tau.rds.path, "no_combat", "qc1_only")        
  }

} else {
  if (!params$qc1.only) {
    input.tau.path <- file.path(params$input.tau.rds.path, "combat", "full_qc")    
  } else {
    input.tau.path <- file.path(params$input.tau.rds.path, "combat", "qc1_only")        
  }

}

# Load expression specificity computations from 10_Expression_Specificity.Rmd
## Conditions available in all species
if (!params$all.nonzero.matrix) {
  if (!params$randomized.mean.expr) {
    filtered.samples <- readRDS(
      file.path(input.tau.path,
                paste(params$species, "filtered.samples.common.rds", sep="_")))     
    
    } else {
      filtered.samples <- readRDS(
        file.path(input.tau.path,
                  paste(params$species, "filtered.samples.common", 
                        paste0("rnd", params$seed.run, ".rds"), sep="_"))) 
    }
 
} else {
  if (!params$randomized.mean.expr) {
    filtered.samples <- readRDS(
      file.path(input.tau.path,
                paste(params$species, "filtered.samples.common.nonzero.rds", sep="_")))      
    
    } else {
      filtered.samples <- readRDS(
        file.path(input.tau.path,
                  paste(params$species, "filtered.samples.common.nonzero", 
                        paste0("rnd", params$seed.run, ".rds"), sep="_")))     
    }
}  


filtered.samples$conditions <- split_gonads_df(filtered.samples$conditions)
expr.rank.first <- filtered.samples$expr.rank.first
```

### Expression variability
```{r import-variability}
if (!params$combat) {
  if (!params$qc1.only) {
    input.variability.path <- file.path(params$input.variability.rds.path, "no_combat", "full_qc")    
  } else {
    input.variability.path <- file.path(params$input.variability.rds.path, "no_combat", "qc1_only")        
  }

} else {
  if (!params$qc1.only) {
    input.variability.path <- file.path(params$input.variability.rds.path, "combat", "full_qc")    
  } else {
    input.variability.path <- file.path(params$input.variability.rds.path, "combat", "qc1_only")        
  }

}

# Load processed expression variability estimates from 02b_Variability_NoJack.Rmd
if (!params$all.nonzero.matrix) {
  expr.var <- readRDS(
    file.path(input.variability.path,
              paste(params$species, "var.summary.bind.rds", sep="_"))) 
  } else {
    expr.var <- readRDS(
      file.path(input.variability.path,
                paste(params$species, "var.summary.bind.nonzero.rds", sep="_")))
  }

expr.var <- split_gonads_df(expr.var)

# Retain only conditions included in expression specificity computations
expr.var.common <- expr.var %>%
  dplyr::filter(Tissue %in% filtered.samples$conditions$Tissue)
```


# Main {.tabset}

## Reorder organs for plotting
```{r reorder-conditions}
expr.var.rank.common <- expr.var.common %>%
  dplyr::inner_join(expr.rank.first, by="GeneID")

if ("Sex" %in% colnames(filtered.samples$metadata)) {
  levels.order <- 
    c("brain","eye","intestine","liver",
      "heart","muscle","gills","pectoral_fin",
      "ovary","testis")
} else {
  levels.order <-
    c("brain","eye","intestine","liver",
      "heart","muscle","gills","pectoral_fin","gonads")  
}

# Exclude missing conditions from factor levels
levels.exclude <- setdiff(levels.order, levels(as.factor(expr.var.rank.common$Tissue)))

# Reorder levels for plotting
expr.var.rank.common$Tissue <-
  factor(expr.var.rank.common$Tissue, levels=levels.order, exclude=levels.exclude)
```


## Variability of organ-biased genes

### Classify broadly expressed and organ-biased genes (with subsets)
```{r classify-genes-by-organ-bias}
Tissue <- as.factor(expr.var.rank.common$Tissue)

# Per organ, classify genes by organ bias
if ("Sex" %in% colnames(expr.var.rank.common)) {
  expr.var.bias.by.condition <- vector(mode="list", length=2)
  names(expr.var.bias.by.condition) <- c("F","M")
  
  for (s in c("F","M")) {
    expr.var.bias.by.condition[[s]] <- vector(mode="list", length=length(levels(Tissue)))
    names(expr.var.bias.by.condition[[s]]) <- levels(Tissue)
    
    for (t in levels(Tissue)) {
      expr.var.bias.by.condition[[s]][[t]] <- expr.var.rank.common %>%
        dplyr::filter(Sex==s) %>%
        dplyr::filter(Tissue!=t) %>%
        classify_genes_by_organ_bias(tau.cutoff=params$tau.cutoff)
      
      # Reorder factor levels
      expr.var.bias.by.condition[[s]][[t]]$TissueBias <-
        factor(expr.var.bias.by.condition[[s]][[t]]$TissueBias, levels=c("broad","focal",t,"other"))
    }
  }
} else {
  expr.var.bias.by.condition <- vector(mode="list", length=length(levels(Tissue)))
  names(expr.var.bias.by.condition) <- levels(Tissue)  
  
  for (t in levels(Tissue)) {
    expr.var.bias.by.condition[[t]] <- expr.var.rank.common %>%
      dplyr::filter(Tissue!=t) %>%
      classify_genes_by_organ_bias(tau.cutoff=params$tau.cutoff)
  
  # Reorder factor levels
    expr.var.bias.by.condition[[t]]$TissueBias <-
      factor(expr.var.bias.by.condition[[t]]$TissueBias, levels=c("broad","focal",t,"other"))    
  }
}
```


### Boxplots by organ bias {.tabset}

#### Residual MAD
```{r boxplot-resid-var-organ-bias, fig.width=18, fig.height=10}
if ("Sex" %in% colnames(expr.var.rank.common)) {
  for (s in c("F","M")) {
    for (t in levels(Tissue)) {
      print(boxplot_var_metric_by_organ_bias(
        expr.var.bias.by.condition[[s]][[t]], 
        metric="Resid_MAD", species=params$species, tissue=t, sex=s, ylab="Residual MAD"))
    }
  }
} else {
  for (t in levels(Tissue)) {
    print(boxplot_var_metric_by_organ_bias(
      expr.var.bias.by.condition[[t]],
      metric="Resid_MAD", species=params$species, tissue=t, ylab="Residual MAD"))    
  }
}
```

#### Examples

##### Brain-biased genes expressed in the pectoral fin
```{r brain-biased-in-pectoral-fin, fig.width=3.5, fig.height=5}
bias <- "brain"
focal <- "pectoral_fin"
  
if ("Sex" %in% colnames(expr.var.rank.common)) {
  sex <- "F"
  brain.biased.in.pecfin <- 
    expr.var.bias.by.condition[[sex]][[bias]] %>%
    dplyr::filter(Tissue==focal)

} else {
  brain.biased.in.pecfin <- 
    expr.var.bias.by.condition[[bias]] %>%
    dplyr::filter(Tissue==focal)
  
}

if (params$randomized.mean.expr) {
  print(boxplot_variability_example_v2(
    brain.biased.in.pecfin, species=params$species.name, metric="Resid_MAD", 
    tissue=bias, ylab="Residual MAD", color="#FFFAFA"))    
} else {
  print(boxplot_variability_example_v2(
    brain.biased.in.pecfin, species=params$species.name, metric="Resid_MAD",
    tissue=bias, ylab="Residual MAD", color="#67005D"))
    
  print(boxplot_variability_example_v2(
    brain.biased.in.pecfin, species=params$species.name, metric="Resid_MAD",
    tissue=bias, ylab="Residual MAD", color="#00BFC4"))    
}
```


##### Liver-biased genes expressed in the intestine
```{r liver-biased-in-intestine, fig.width=3.5, fig.height=5}
bias <- "liver"
focal <- "intestine"
  
if ("Sex" %in% colnames(expr.var.rank.common)) {
  sex <- "F"
  liver.biased.in.intestine <- 
    expr.var.bias.by.condition[[sex]][[bias]] %>%
    dplyr::filter(Tissue==focal)

} else {
  liver.biased.in.intestine <- 
    expr.var.bias.by.condition[[bias]] %>%
    dplyr::filter(Tissue==focal)
    
}

if (params$randomized.mean.expr) {
  print(boxplot_variability_example_v2(
    liver.biased.in.intestine, species=params$species.name, metric="Resid_MAD",
    tissue=bias, ylab="Residual MAD", color="#FFFAFA"))    
} else {
  print(boxplot_variability_example_v2(
    liver.biased.in.intestine, species=params$species.name, metric="Resid_MAD", 
    tissue=bias, ylab="Residual MAD", color="#FF8A00"))
  
  print(boxplot_variability_example_v2(
    liver.biased.in.intestine, species=params$species.name, metric="Resid_MAD",
    tissue=bias, ylab="Residual MAD", color="#F8766D"))    
}
```


##### Ovary-biased (or gonads-biased) genes expressed in the eye
```{r gonads-biased-in-eye, fig.width=3.5, fig.height=5}
focal <- "eye"
  
if ("Sex" %in% colnames(expr.var.rank.common)) {
  bias <- "ovary"
  sex <- "F"
  ovary.biased.in.eye <- 
    expr.var.bias.by.condition[[sex]][[bias]] %>%
    dplyr::filter(Tissue==focal)
  
  if (params$randomized.mean.expr) {
    print(boxplot_variability_example_v2(
      ovary.biased.in.eye, species=params$species.name, metric="Resid_MAD", 
      tissue=bias, ylab="Residual MAD", color="#FFFAFA"))      
  } else {
    print(boxplot_variability_example_v2(
      ovary.biased.in.eye, species=params$species.name, metric="Resid_MAD", 
      tissue=bias, ylab="Residual MAD", color="#67005D"))
  
    print(boxplot_variability_example_v2(
      ovary.biased.in.eye, species=params$species.name, metric="Resid_MAD",
      tissue=bias, ylab="Residual MAD", color="#00BFC4"))     
  }
  
} else {
  bias <- "gonads"
  gonads.biased.in.eye <- 
    expr.var.bias.by.condition[[bias]] %>%
    dplyr::filter(Tissue==focal)

  if (params$randomized.mean.expr) {
    print(boxplot_variability_example_v2(
      gonads.biased.in.eye, species=params$species.name, metric="Resid_MAD",
      tissue=bias, ylab="Residual MAD", color="#FFFAFA"))  
  } else {
    print(boxplot_variability_example_v2(
      gonads.biased.in.eye, species=params$species.name, metric="Resid_MAD",
      tissue=bias, ylab="Residual MAD", color="#67005D"))
  
    print(boxplot_variability_example_v2(
      gonads.biased.in.eye, species=params$species.name, metric="Resid_MAD",
      tissue=bias, ylab="Residual MAD", color="#00BFC4"))  
  }
}
```



## Multiple pairwise comparisons {.tabset}

### Description

Variability of the set of _x_-biased genes expressed in organ _i_ (not _x_) vs. variability of other organ-biased genes in organ _i_ (excluding _i_-biased genes)

e.g., variability of brain-biased genes expressed in the ovary vs. variability of other organ-biased genes expressed in the ovary (excluding ovary-biased genes)

### Exclude broadly expressed and focal-biased genes
```{r exclude-broad-and-focal-biased-genes}
if ("Sex" %in% colnames(expr.var.rank.common)) {
  expr.var.bias.comparisons <-
    lapply(c("F"="F","M"="M"), function(s) {
      dplyr::bind_rows(expr.var.bias.by.condition[[s]], .id="Bias") %>%
        dplyr::filter(!(TissueBias %in% c("broad","focal")))
      }) %>%
    dplyr::bind_rows()  
} else {
  expr.var.bias.comparisons <- expr.var.bias.by.condition %>%
    dplyr::bind_rows(.id="Bias") %>%
    dplyr::filter(!(TissueBias %in% c("broad","focal")))
}
```


### MAD

#### Effect size
```{r compute-effect-size-mad}
comparisons.mad <- list()

comparisons.mad$effect.size <- 
  compute_glass_delta_effect_size_v2(
    expr.var.bias.comparisons, metric="MAD")
```

#### Pairwise Wilcoxon test
```{r pairwise-comparisons-mad}
# Complete
comparisons.mad$wilcox.complete <- 
  compute_pairwise_comparisons_v2(
    comparisons.df=expr.var.bias.comparisons,
    effect.size.df=comparisons.mad$effect.size,
    metric="MAD",
    apply.pvalue.cutoff=FALSE,
    order.levels=levels(Tissue)
  )

# Keep significant observations only
comparisons.mad$wilcox.signif <- 
  compute_pairwise_comparisons_v2(
    comparisons.df=expr.var.bias.comparisons, 
    effect.size.df=comparisons.mad$effect.size,
    metric="MAD",
    apply.pvalue.cutoff=TRUE, adj.p=0.05, 
    order.levels=levels(Tissue))
```

#### Plot - Heatmap
```{r heatmap-pairwise-comparisons-mad, fig.height=10, fig.width=14}
heatmap_effect_size_pairwise_comparisons_blue2red(
  comparisons.mad$wilcox.complete, species=params$species.name, 
  title="Median absolute deviation", order.levels=levels.order)

heatmap_effect_size_pairwise_comparisons(
  comparisons.mad$wilcox.signif, species=params$species.name,
  title="Median absolute deviation", order.levels=levels.order)
```


### Residual MAD

#### Effect size
```{r compute-effect-size-resid-mad}
comparisons.resid.mad <- list()

comparisons.resid.mad$effect.size <- 
  compute_glass_delta_effect_size_v2(
    expr.var.bias.comparisons, metric="Resid_MAD")
```

#### Pairwise Wilcoxon test
```{r pairwise-comparisons-resid-mad}
# Complete
comparisons.resid.mad$wilcox.complete <- 
  compute_pairwise_comparisons_v2(
    comparisons.df=expr.var.bias.comparisons,
    effect.size.df=comparisons.resid.mad$effect.size,
    metric="Resid_MAD",
    apply.pvalue.cutoff=FALSE,
    order.levels=levels(Tissue)
  )

# Keep significant observations only
comparisons.resid.mad$wilcox.signif <- 
  compute_pairwise_comparisons_v2(
    comparisons.df=expr.var.bias.comparisons, 
    effect.size.df=comparisons.resid.mad$effect.size,
    metric="Resid_MAD",
    apply.pvalue.cutoff=TRUE, adj.p=0.05, 
    order.levels=levels(Tissue))
```

#### Plot - Heatmap
```{r heatmap-pairwise-comparisons-resid-mad, fig.height=10, fig.width=14}
heatmap_effect_size_pairwise_comparisons_blue2red(
  comparisons.resid.mad$wilcox.complete, species=params$species.name, 
  title="Residual MAD", order.levels=levels.order)

heatmap_effect_size_pairwise_comparisons(
  comparisons.resid.mad$wilcox.signif, species=params$species.name,
  title="Residual MAD", order.levels=levels.order)
```


# Save {.tabset}

## Observed
```{r save-objects-1}
if (!params$combat) {
  if (!params$qc1.only) {
    output.path <- file.path(params$output.rds.path, "no_combat", "full_qc")    
  } else {
    output.path <- file.path(params$output.rds.path, "no_combat", "qc1_only")        
  }

} else {
  if (!params$qc1.only) {
    output.path <- file.path(params$output.rds.path, "combat", "full_qc")    
  } else {
    output.path <- file.path(params$output.rds.path, "combat", "qc1_only")        
  }

}

if (!dir.exists(output.path)) { dir.create(output.path, recursive=TRUE, showWarnings=FALSE) }

if (!params$all.nonzero.matrix) {
  
  saveRDS(expr.var.bias.comparisons,
          file.path(output.path,
                    paste(params$species, "expr.var.bias.comparisons",
                          paste(paste0("tau", params$tau.cutoff), "rds", sep="."), sep="_")))
  
  saveRDS(comparisons.mad,
          file.path(output.path, 
                    paste(params$species, "comparisons.mad",
                          paste(paste0("tau", params$tau.cutoff), "rds", sep="."), sep="_")))
  
  saveRDS(comparisons.resid.mad,
          file.path(output.path, 
                    paste(params$species, "comparisons.resid.mad",
                          paste(paste0("tau", params$tau.cutoff), "rds", sep="."), sep="_")))
  

  
} else {
  
  saveRDS(expr.var.bias.comparisons,
          file.path(output.path,
                    paste(params$species, "expr.var.bias.comparisons.nonzero",
                          paste(paste0("tau", params$tau.cutoff), "rds", sep="."), sep="_")))
  
  saveRDS(comparisons.mad,
          file.path(output.path, 
                    paste(params$species, "comparisons.mad.nonzero",
                          paste(paste0("tau", params$tau.cutoff), "rds", sep="."), sep="_")))
  
  saveRDS(comparisons.resid.mad,
          file.path(output.path, 
                    paste(params$species, "comparisons.resid.mad.nonzero",
                          paste(paste0("tau", params$tau.cutoff), "rds", sep="."), sep="_")))        
}  

```

## Randomized mean expression
```{r save-objects-2, eval=params$randomized.mean.expr}
if (!params$combat) {
  if (!params$qc1.only) {
    output.path <- file.path(params$output.rds.path, "no_combat", "full_qc")    
  } else {
    output.path <- file.path(params$output.rds.path, "no_combat", "qc1_only")        
  }

} else {
  if (!params$qc1.only) {
    output.path <- file.path(params$output.rds.path, "combat", "full_qc")    
  } else {
    output.path <- file.path(params$output.rds.path, "combat", "qc1_only")
  }

}

if (!dir.exists(output.path)) { dir.create(output.path, recursive=TRUE, showWarnings=FALSE) }

if (!params$all.nonzero.matrix) {

  saveRDS(expr.var.bias.comparisons,
          file.path(output.path, 
                    paste(params$species, "expr.var.bias.comparisons", 
                          paste(paste0("tau", params$tau.cutoff),
                                paste0("rnd", params$seed.run), "rds", sep="."), sep="_")))

  saveRDS(comparisons.mad,
          file.path(output.path, 
                    paste(params$species, "comparisons.mad",
                          paste(paste0("tau", params$tau.cutoff),
                                paste0("rnd", params$seed.run), "rds", sep="."), sep="_")))
  
  saveRDS(comparisons.resid.mad,
          file.path(output.path, 
                    paste(params$species, "comparisons.resid.mad", 
                          paste0("rnd", params$seed.run, ".rds"), sep="_")))
  
} else {
  saveRDS(expr.var.bias.comparisons,
          file.path(output.path, 
                    paste(params$species, "expr.var.bias.comparisons.nonzero", 
                          paste(paste0("tau", params$tau.cutoff),
                                paste0("rnd", params$seed.run), "rds", sep="."), sep="_")))

  saveRDS(comparisons.mad,
          file.path(output.path, 
                    paste(params$species, "comparisons.mad.nonzero", 
                          paste(paste0("tau", params$tau.cutoff),
                                paste0("rnd", params$seed.run), "rds", sep="."), sep="_")))
  
  saveRDS(comparisons.resid.mad,
          file.path(output.path, 
                    paste(params$species, "comparisons.resid.mad.nonzero", 
                          paste(paste0("tau", params$tau.cutoff),
                                paste0("rnd", params$seed.run), "rds", sep="."), sep="_")))
  }  

```