---
title: "WGCNA (*`r params$species.name`*)"
author: "Bucao, Christabel Floi"
output: html_document
date: "`r Sys.Date()`"
params:
  input.samples.rds.path: "../../results/01_Filter_Samples"
  input.variability.rds.path: "../../results/03_Variability_Plots"
  input.bias.rds.path: "../../results/12_Organ_Bias"
  output.rds.path: "../../results/15_WGCNA"
  species: "DRE"
  species.name: "Danio rerio"
  tau.cutoff: 0.30
  min.replicates: 10
  qc1.only: FALSE
  combat: FALSE
  set.seed: 12345
  power: 10
---

# Setup {.tabset}

## Parameters
```{r print-params}
print(params)
```

## Libraries
```{r install-libraries, eval=FALSE}
# For WGCNA installation

## Install Hmisc dependencies
install.packages(c("latticeExtra","data.table","htmlTable"))

## Install archived version of Hmisc
install.packages(
    "https://cran.r-project.org/src/contrib/Archive/Hmisc/Hmisc_4.6-0.tar.gz",
    repos = NULL,
    type = "source"
)

# Install WGCNA
BiocManager::install('WGCNA')
BiocManager::install('flashClust')

renv::snapshot()
```

```{r import-libraries}
packages <- c("dplyr","tibble","matrixStats","corpcor","coseq","corrplot","RColorBrewer","ggplot2","ggpubr","pheatmap","WGCNA","flashClust","viridis")

for (p in packages) {
  if (!require(p, character.only=TRUE)) {
    print(paste("Installing", p))
    install.packages(p, character.only=TRUE)
    renv::snapshot()
  }
  print(paste("Loading", p))
  library(p, character.only=TRUE)
}

allowWGCNAThreads() 
```

## Functions
```{r source-functions}
source("../functions/helper_functions.R")
source("../functions/organ_bias.R")
#source("../functions/bimodality_test.R")
```

## Session Info
```{r print-session-info}
sessionInfo()
```

## Data

### Expression levels
```{r import-expression-levels}
if (!params$combat) {
  if (!params$qc1.only) {
    input.samples.path <- file.path(params$input.samples.rds.path, "no_combat", "full_qc")    
  } else {
    input.samples.path <- file.path(params$input.samples.rds.path, "no_combat", "qc1_only")        
  }

} else {
  if (!params$qc1.only) {
    input.samples.path <- file.path(params$input.samples.rds.path, "combat", "full_qc")    
  } else {
    input.samples.path <- file.path(params$input.samples.rds.path, "combat", "qc1_only")        
  }

}

# Load saved data from 01_Filter_Samples.Rmd
# Normalized within-condition
if (!params$qc1.only) {
  filtered.samples <- readRDS(
    file.path(input.samples.path,
              paste(params$species, "4_reps_data.conditions.rds", sep="_")))    
} else {
  filtered.samples <- readRDS(
    file.path(input.samples.path,
              paste(params$species, "4_reps_data.qc1.conditions.rds", sep="_")))      
}

```

### Expression variability
```{r import-variability}
if (!params$combat) {
  if (!params$qc1.only) {
    input.variability.path <- file.path(params$input.variability.rds.path, "no_combat", "full_qc")    
  } else {
    input.variability.path <- file.path(params$input.variability.rds.path, "no_combat", "qc1_only")        
  }

} else {
  if (!params$qc1.only) {
    input.variability.path <- file.path(params$input.variability.rds.path, "combat", "full_qc")    
  } else {
    input.variability.path <- file.path(params$input.variability.rds.path, "combat", "qc1_only")            
  }

}

# Load processed expression variability estimates from 03_Variability_Plots.Rmd
jack.bind <- readRDS(
  file.path(input.variability.path,
            paste(params$species, "jack.resid.cv.bind.rds", sep="_")))
```

### Expression variability - Organ bias
```{r import-variability-organ-bias}
if (!params$combat) {
  if (!params$qc1.only) {
    input.bias.path <- file.path(params$input.bias.rds.path, "no_combat", "full_qc")    
  } else {
    input.bias.path <- file.path(params$input.bias.rds.path, "no_combat", "qc1_only")        
  }

} else {
  if (!params$qc1.only) {
    input.bias.path <- file.path(params$input.bias.rds.path, "combat", "full_qc")    
  } else {
    input.bias.path <- file.path(params$input.bias.rds.path, "combat", "qc1_only")            
  }

}

# Load expression variability - organ bias data from 12_Organ_Bias.Rmd
expr.var.rank.common <- readRDS(
  file.path(input.bias.path,
            paste(params$species, "expr.var.rank.common", 
                  paste0("tau", params$tau.cutoff, ".rds"), sep="_"))) %>%
  dplyr::filter(Biotype=="protein_coding")

# Set order levels for plotting
if ("Sex" %in% colnames(filtered.samples$metadata)) {
  levels.order <- 
    c("broad","brain","eye","intestine","liver",
      "heart","muscle","gills","pectoral_fin",
      "ovary","testis")
} else {
  levels.order <-
    c("broad","brain","eye","intestine","liver",
      "heart","muscle","gills","pectoral_fin","gonads")  
}

levels.exclude <- 
  setdiff(levels.order, c("broad",levels(as.factor(expr.var.rank.common$PrimaryTissue))))
```


# Main {.tabset}

## 0. Select conditions with minimum replicates
```{r select-conditions}
filtered.samples.retained <- list()

filtered.samples.retained$conditions <- filtered.samples$conditions %>%
  dplyr::filter(n>=params$min.replicates) %>%
  dplyr::filter(Tissue %in% c(levels.order, "gonads"))

filtered.samples.retained$metadata <- filtered.samples$metadata %>%
  dplyr::filter(grepl(paste(filtered.samples.retained$conditions$regex,collapse="|"),.$SampleName))

if ("Sex" %in% colnames(filtered.samples.retained$conditions)) {
  conditions <- 
    paste(filtered.samples.retained$conditions$Tissue, 
          filtered.samples.retained$conditions$Sex, sep="_")
} else {
  conditions <- filtered.samples.retained$conditions$Tissue
}

filtered.samples.retained$counts.gf <- vector(mode="list", length=length(conditions))
names(filtered.samples.retained$counts.gf) <- conditions

filtered.samples.retained$log2.tmm.cpm.gf <- vector(mode="list", length=length(conditions))
names(filtered.samples.retained$log2.tmm.cpm.gf) <- conditions

jack.condition.retained <- vector(mode="list", length=length(conditions))
names(jack.condition.retained) <- conditions

for (c in 1:length(conditions)) {
  if ("Sex" %in% colnames(filtered.samples.retained$conditions)) {
    t <- filtered.samples.retained$conditions$Tissue[[c]]
    s <- filtered.samples.retained$conditions$Sex[[c]]
    
    ts <- paste(t,s,sep="_")
    
    filtered.samples.retained$counts.gf[[ts]] <- filtered.samples$counts.gf[[t]][[s]]
    filtered.samples.retained$log2.tmm.cpm.gf[[ts]] <- filtered.samples$log2.tmm.cpm.gf[[t]][[s]]
    
    jack.condition.retained[[ts]] <- jack.bind %>%
      dplyr::filter(Biotype=="protein_coding") %>%
      dplyr::filter(paste(.$Tissue, .$Sex, sep="_")==ts)
    
  } else {
    ts <- filtered.samples.retained$conditions$Tissue[[c]]
    
    filtered.samples.retained$log2.tmm.cpm.gf[[ts]] <- filtered.samples$log2.tmm.cpm.gf[[ts]]
    
    jack.condition.retained[[ts]] <- jack.bind %>%
      dplyr::filter(Biotype=="protein_coding") %>%
      dplyr::filter(Tissue==ts)
  }
  
  # Filter gene list
  filtered.samples.retained$counts.gf[[ts]] <- filtered.samples.retained$counts.gf[[ts]] %>%
    .[rownames(.) %in% jack.condition.retained[[ts]]$GeneID,]
  
  filtered.samples.retained$log2.tmm.cpm.gf[[ts]] <- filtered.samples.retained$log2.tmm.cpm.gf[[ts]] %>%
    .[rownames(.) %in% jack.condition.retained[[ts]]$GeneID,]
}

filtered.samples.retained$conditions <- split_gonads_df(filtered.samples.retained$conditions)
filtered.samples.retained$metadata <- split_gonads_df(filtered.samples.retained$metadata)

if ("Sex" %in% colnames(filtered.samples.retained$metadata)) {
  conditions <- 
    paste(filtered.samples.retained$conditions$Tissue, 
          filtered.samples.retained$conditions$Sex, sep="_") 
} else {
  conditions <- filtered.samples.retained$conditions$Tissue
}

```


## 1. WGCNA - By organ bias & variability

### 1.0 Pre-processing
```{r preprocess-observed}
expr.var.rank.common$TissueBias3 <- 
  apply(expr.var.rank.common, MARGIN=1, FUN=function(r) {
    if (r[["tau"]]<params$tau.cutoff) { "broad" }
    else if (r[["Tissue"]]==r[["PrimaryTissue"]]) { "focal" }
    else { "other" }
  })

TissueBias3 <- levels(as.factor(expr.var.rank.common$TissueBias3))

# Group variability ranks into bins of size 0.1
expr.var.rank.common$VarRankBin <- 
  ceiling(expr.var.rank.common$Mean_Local_Rank_Log2CV*10)/10
expr.var.rank.common$VarRankBin[round(expr.var.rank.common$VarRankBin,1)==0.0] <- 0.1

# Group variability ranks into bins of size 0.2
expr.var.rank.common$VarRankBin2 <- expr.var.rank.common$VarRankBin
expr.var.rank.common$VarRankBin2[expr.var.rank.common$VarRankBin<=0.2] <- 0.2
expr.var.rank.common$VarRankBin2[
  expr.var.rank.common$VarRankBin>0.2 & expr.var.rank.common$VarRankBin>=0.4] <- 0.4
expr.var.rank.common$VarRankBin2[
  expr.var.rank.common$VarRankBin>0.4 & expr.var.rank.common$VarRankBin>=0.6] <- 0.6
expr.var.rank.common$VarRankBin2[
  expr.var.rank.common$VarRankBin>0.6 & expr.var.rank.common$VarRankBin>=0.8] <- 0.8
expr.var.rank.common$VarRankBin2[
  expr.var.rank.common$VarRankBin>0.8 & expr.var.rank.common$VarRankBin>=1.0] <- 1.0

VarRankBins2 <- levels(as.factor(seq(0.2,1,0.2)))

# Define expression matrix subsets by organ bias - variability
expr.subsets <- list()

for (c in conditions) {
  expr.subsets[[c]] <- vector(mode="list", length=length(TissueBias3))
  names(expr.subsets[[c]]) <- TissueBias3
  
  for (b in TissueBias3) {
    
    expr.subsets[[c]][[b]] <- vector(mode="list", length=length(VarRankBins2))
    names(expr.subsets[[c]][[b]]) <- VarRankBins2
    
    if ("Sex" %in% colnames(filtered.samples.retained$metadata)) {
      t <- unlist(strsplit(c, split="_[FM]"))[1]
      s <- substr(c,nchar(c),nchar(c))
  
      ts <- c
      if (t=="ovary") { ts <- "gonads_F" }
      if (t=="testis") { ts <- "gonads_M" }
      
      for (v in VarRankBins2) {
        gene.set <- 
          unique(expr.var.rank.common$GeneID[
            expr.var.rank.common$VarRankBin2==as.numeric(v) &
              expr.var.rank.common$TissueBias3==b &
              expr.var.rank.common$Tissue==t & 
              expr.var.rank.common$Sex==s])
    
        expr.subsets[[c]][[b]][[v]] <-
          filtered.samples.retained$log2.tmm.cpm.gf[[ts]][
            rownames(filtered.samples.retained$log2.tmm.cpm.gf[[ts]]) %in% gene.set,]      
      }      
    } else {
      for (v in VarRankBins2) {
        gene.set <- 
          unique(expr.var.rank.common$GeneID[
          expr.var.rank.common$VarRankBin2==as.numeric(v) &
            expr.var.rank.common$TissueBias3==b &
            expr.var.rank.common$Tissue==c])
    
        expr.subsets[[c]][[b]][[v]] <-
          filtered.samples.retained$log2.tmm.cpm.gf[[c]][
            rownames(filtered.samples.retained$log2.tmm.cpm.gf[[c]]) %in% gene.set,]      
      }    
    }

  }
}
```


### 1.1 Pick soft threshold
```{r pick-threshold-observed, eval=FALSE}
# Choose a set of soft-threshold powers
powers <- c(c(1:10), seq(from=12, to=30, by=2))

thresh.subsets <- list()

for (c in conditions) {
  thresh.subsets[[c]] <- vector(mode="list", length=length(TissueBias3))
  names(thresh.subsets[[c]]) <- TissueBias3
  
  for (b in TissueBias3) {
    thresh.subsets[[c]][[b]] <- vector(mode="list", length=length(VarRankBins2))
    names(thresh.subsets[[c]][[b]]) <- VarRankBins2
    
    for (v in VarRankBins2) {
      thresh.subsets[[c]][[b]][[v]] <-
        pickSoftThreshold(t(expr.subsets[[c]][[b]][[v]]), powerVector=powers)
    }
  }
}
```


### 1.2 Find modules
```{r find-modules-observed}
superpower <- params$power
temp.cor <- cor
cor <- WGCNA::cor

module.subsets <- list()

for (c in conditions) {
  module.subsets[[c]] <- vector(mode="list", length=length(TissueBias3))
  names(module.subsets[[c]]) <- TissueBias3
  
  for (b in TissueBias3) {
    module.subsets[[c]][[b]] <- vector(mode="list", length=length(VarRankBins2))
    names(module.subsets[[c]][[b]]) <- VarRankBins2
    
    for (v in VarRankBins2) {
      module.subsets[[c]][[b]][[v]] <- 
        blockwiseModules(t(expr.subsets[[c]][[b]][[v]]),
                         power = superpower,
                         networkType = "signed",
                         deepSplit = 2,
                         pamRespectsDendro = F,
                         minModuleSize = 30,
                         maxBlockSize = 4000,
                         reassignThreshold = 0,
                         mergeCutHeight = 0.25,
                         saveTOMs = T,
                         saveTOMFileBase = "ER",
                         numericLabels = T)      
    }
  }
}

cor <- temp.cor # Return cor function to original namespace
```


### 1.3 Network adjacency
```{r adjacency-observed}
network.adjacency <- list()
connectivity <- list()

for (c in conditions) {
  network.adjacency[[c]] <- vector(mode="list", length=length(TissueBias3))
  names(network.adjacency[[c]]) <- TissueBias3
  
  connectivity[[c]] <- vector(mode="list", length=length(TissueBias3))
  names(connectivity[[c]]) <- TissueBias3
  
  for (b in TissueBias3) {
    network.adjacency[[c]][[b]] <- vector(mode="list", length=length(VarRankBins2))
    names(network.adjacency[[c]][[b]]) <- VarRankBins2
    
    connectivity[[c]][[b]] <- vector(mode="list", length=length(VarRankBins2))
    names(connectivity[[c]][[b]]) <- VarRankBins2
    
    for (v in VarRankBins2) {
      network.adjacency[[c]][[b]][[v]] <- 
        WGCNA::adjacency(t(expr.subsets[[c]][[b]][[v]]),
                         type="signed",
                         power=superpower,
                         corFnc = "cor",
                         corOptions = "use='p'")
    
      connectivity[[c]][[b]][[v]] <- intramodularConnectivity(
        network.adjacency[[c]][[b]][[v]], module.subsets[[c]][[b]][[v]]$colors
      )
    
      connectivity[[c]][[b]][[v]]$module <- module.subsets[[c]][[b]][[v]]$colors
      connectivity[[c]][[b]][[v]]$moduleColor <- labels2colors(module.subsets[[c]][[b]][[v]]$colors)      
    }
    
  }
}
```


### 1.4 Modularity metrics
```{r modularity-metrics-observed}
modularity.metrics <- data.frame(
  "Condition"=rep(conditions, each=length(VarRankBins2)*length(TissueBias3)),
  "TissueBias3"=rep(rep(TissueBias3, each=length(VarRankBins2)), length(conditions)),
  "VarRankBin2"=rep(VarRankBins2, length(conditions)*length(TissueBias3)))

for (c in conditions) {
  for (b in TissueBias3) {
    for (v in VarRankBins2) {
      ## Proportion of genes assigned to dummy module (gray; 0)
      modularity.metrics$PropDummy[modularity.metrics$Condition==c & 
                                              modularity.metrics$TissueBias3==b &
                                              modularity.metrics$VarRankBin2==v] <- 
        nrow(connectivity[[c]][[b]][[v]][connectivity[[c]][[b]][[v]]$module==0,]) /
        nrow(connectivity[[c]][[b]][[v]])
    
    ## Proportion of genes assigned to top module (1)
    modularity.metrics$PropTop[modularity.metrics$Condition==c &
                                          modularity.metrics$TissueBias3==b &
                                          modularity.metrics$VarRankBin2==v] <- 
      nrow(connectivity[[c]][[b]][[v]][connectivity[[c]][[b]][[v]]$module==1,]) /
      nrow(connectivity[[c]][[b]][[v]])    
    
    ## Distribution of kDiff for genes assigned to other modules
    modularity.metrics$Mean_kDiff[modularity.metrics$Condition==c &
                                             modularity.metrics$TissueBias3==b &
                                             modularity.metrics$VarRankBin2==v] <- 
      mean(connectivity[[c]][[b]][[v]]$kDiff[connectivity[[c]][[b]][[v]]$module!=0])
    
    modularity.metrics$Median_kDiff[modularity.metrics$Condition==c &
                                               modularity.metrics$TissueBias3==b &
                                               modularity.metrics$VarRankBin2==v] <- 
      median(connectivity[[c]][[b]][[v]]$kDiff[connectivity[[c]][[b]][[v]]$module!=0])
    
    ## Distribution of kDiff for genes assigned to dummy module
    modularity.metrics$Mean_kDiff_Dummy[modularity.metrics$Condition==c &
                                                   modularity.metrics$TissueBias3==b &
                                                   modularity.metrics$VarRankBin2==v] <- 
      mean(connectivity[[c]][[b]][[v]]$kDiff[connectivity[[c]][[b]][[v]]$module==0])
    
    modularity.metrics$Median_kDiff_Dummy[modularity.metrics$Condition==c &
                                                     modularity.metrics$TissueBias3==b &
                                                     modularity.metrics$VarRankBin2==v] <- 
      median(connectivity[[c]][[b]][[v]]$kDiff[connectivity[[c]][[b]][[v]]$module==0])
    
    ## Distribution of kWithin for genes assigned to other modules
    modularity.metrics$Mean_kWithin[modularity.metrics$Condition==c &
                                               modularity.metrics$TissueBias3==b &
                                               modularity.metrics$VarRankBin2==v] <- 
      mean(connectivity[[c]][[b]][[v]]$kWithin[connectivity[[c]][[b]][[v]]$module!=0])
    
    modularity.metrics$Median_kWithin[modularity.metrics$Condition==c &
                                                 modularity.metrics$TissueBias3==b & 
                                                 modularity.metrics$VarRankBin2==v] <- 
      median(connectivity[[c]][[b]][[v]]$kWithin[connectivity[[c]][[b]][[v]]$module!=0])
    
    ## Distribution of kWithin for genes assigned to dummy module
    modularity.metrics$Mean_kWithin_Dummy[modularity.metrics$Condition==c &
                                                     modularity.metrics$TissueBias3==b &
                                                     modularity.metrics$VarRankBin2==v] <- 
      mean(connectivity[[c]][[b]][[v]]$kWithin[connectivity[[c]][[b]][[v]]$module==0])
    
    modularity.metrics$Median_kWithin_Dummy[modularity.metrics$Condition==c &
                                                       modularity.metrics$TissueBias3==b &
                                                       modularity.metrics$VarRankBin2==v] <- 
      median(connectivity[[c]][[b]][[v]]$kWithin[connectivity[[c]][[b]][[v]]$module==0])
          
    }

  }
}
```


### 1.5 Summary Plots
```{r plots-observed}
ggplot(modularity.metrics, 
       aes(x=VarRankBin2, y=Condition, fill=PropDummy)) + 
  geom_tile() +
  facet_wrap(vars(TissueBias3)) +
  colorspace::scale_fill_continuous_diverging(mid=0, na.value="#A9A9A9") +
  labs(x="Variability rank bin", fill="Prop(no module)") +
  theme_bw()

ggplot(modularity.metrics, 
       aes(x=VarRankBin2, y=Condition, fill=PropTop)) + 
  geom_tile() +
  facet_wrap(vars(TissueBias3)) +
  colorspace::scale_fill_continuous_diverging(mid=0, na.value="#A9A9A9") +
  labs(x="Variability rank bin", fill="Prop(top module)") +
  theme_bw()

ggplot(modularity.metrics, 
       aes(x=VarRankBin2, y=Condition, fill=log2(Mean_kDiff - min(Mean_kDiff, na.rm=TRUE) + 1))) + 
  geom_tile() +
  facet_wrap(vars(TissueBias3)) +  
  colorspace::scale_fill_continuous_diverging(mid=0, na.value="#A9A9A9") +
  labs(x="Variability rank bin", fill="Mean kDiff") +
  theme_bw()

ggplot(modularity.metrics, 
       aes(x=VarRankBin2, y=Condition, fill=log2(Median_kDiff - min(Median_kDiff, na.rm=TRUE) + 1))) + 
  geom_tile() +
  facet_wrap(vars(TissueBias3)) +
  colorspace::scale_fill_continuous_diverging(mid=0, na.value="#A9A9A9") +
  labs(x="Variability rank bin", fill="Median kDiff") +
  theme_bw()

ggplot(modularity.metrics, 
       aes(x=VarRankBin2, y=Condition, fill=log2(Mean_kWithin +1))) + 
  geom_tile() +
  facet_wrap(vars(TissueBias3)) +
  colorspace::scale_fill_continuous_diverging(mid=0, na.value="#A9A9A9") +
  labs(x="Variability rank bin", fill="Mean kWithin") +
  theme_bw()

ggplot(modularity.metrics, 
       aes(x=VarRankBin2, y=Condition, fill=log2(Median_kWithin +1))) + 
  geom_tile() +
  facet_wrap(vars(TissueBias3)) +
  colorspace::scale_fill_continuous_diverging(mid=0, na.value="#A9A9A9") +
  labs(x="Variability rank bin", fill="Median kWithin") +
  theme_bw()
```


## 2. WGCNA - Variability - Organ bias - Rowwise permutation

### 2.0 Permute expression per row
```{r permute-subsets}
expr.subsets.perm <- lapply(setNames(conditions, conditions), function(c) {
  lapply(setNames(TissueBias3, TissueBias3), function(b) {
    lapply(setNames(VarRankBins2, VarRankBins2), function(v) {
      permute_by_rows(expr.subsets[[c]][[b]][[v]], seed=params$set.seed)
      })    
  })
})
```


### 2.1 Pick soft threshold
```{r pick-threshold-permuted, eval=FALSE}
# Choose a set of soft-threshold powers
powers <- c(c(1:10), seq(from=12, to=30, by=2))

thresh.subsets.perm <- list()

for (c in conditions) {
  thresh.subsets.perm[[c]] <- vector(mode="list", length=length(TissueBias3))
  names(thresh.subsets.perm[[c]]) <- TissueBias3
  
  for (b in TissueBias3) {
    thresh.subsets.perm[[c]][[b]] <- vector(mode="list", length=length(VarRankBins2))
    names(thresh.subsets.perm[[c]][[b]]) <- VarRankBins2
    
    for (v in VarRankBins2) {
      thresh.subsets.perm[[c]][[b]][[v]] <-
        pickSoftThreshold(t(expr.subsets.perm[[c]][[b]][[v]]), powerVector=powers)
    }
  }
}
```


### 2.2 Find modules
```{r find-modules-permuted}
superpower <- params$power
temp.cor <- cor
cor <- WGCNA::cor

module.subsets.perm <- list()

for (c in conditions) {
  module.subsets.perm[[c]] <- vector(mode="list", length=length(TissueBias3))
  names(module.subsets.perm[[c]]) <- TissueBias3
  
  for (b in TissueBias3) {
    module.subsets.perm[[c]][[b]] <- vector(mode="list", length=length(VarRankBins2))
    names(module.subsets.perm[[c]][[b]]) <- VarRankBins2
    
    for (v in VarRankBins2) {
      module.subsets.perm[[c]][[b]][[v]] <- 
        blockwiseModules(t(expr.subsets.perm[[c]][[b]][[v]]),
                         power = superpower,
                         networkType = "signed",
                         deepSplit = 2,
                         pamRespectsDendro = F,
                         minModuleSize = 30,
                         maxBlockSize = 4000,
                         reassignThreshold = 0,
                         mergeCutHeight = 0.25,
                         saveTOMs = T,
                         saveTOMFileBase = "ER",
                         numericLabels = T)
    }
  }
}

cor <- temp.cor # Return cor function to original namespace
```


### 2.3 Network adjacency
```{r adjacency-permuted}
network.adjacency.perm <- list()
connectivity.perm <- list()

for (c in conditions) {
  network.adjacency.perm[[c]] <- vector(mode="list", length=length(TissueBias3))
  names(network.adjacency.perm[[c]]) <- TissueBias3
  
  connectivity.perm[[c]] <- vector(mode="list", length=length(TissueBias3))
  names(connectivity.perm[[c]]) <- TissueBias3
  
  for (b in TissueBias3) {
    network.adjacency.perm[[c]][[b]] <- vector(mode="list", length=length(VarRankBins2))
    names(network.adjacency.perm[[c]][[b]]) <- VarRankBins2
    
    connectivity.perm[[c]][[b]] <- vector(mode="list", length=length(VarRankBins2))
    names(connectivity.perm[[c]][[b]]) <- VarRankBins2
    
    for (v in VarRankBins2) {
      network.adjacency.perm[[c]][[b]][[v]] <- 
        WGCNA::adjacency(t(expr.subsets.perm[[c]][[b]][[v]]),
                         type="signed",
                         power=superpower,
                         corFnc = "cor",
                         corOptions = "use='p'")
    
      connectivity.perm[[c]][[b]][[v]] <- intramodularConnectivity(
        network.adjacency.perm[[c]][[b]][[v]], module.subsets.perm[[c]][[b]][[v]]$colors
      )
    
      connectivity.perm[[c]][[b]][[v]]$module <- module.subsets.perm[[c]][[b]][[v]]$colors
      connectivity.perm[[c]][[b]][[v]]$moduleColor <- labels2colors(module.subsets.perm[[c]][[b]][[v]]$colors)      
    }
    
  }
}
```


### 2.4 Modularity metrics
```{r modularity-metrics-permuted}
modularity.metrics.perm <- data.frame(
  "Condition"=rep(conditions, each=length(VarRankBins2)*length(TissueBias3)),
  "TissueBias3"=rep(rep(TissueBias3, each=length(VarRankBins2)), length(conditions)),
  "VarRankBin2"=rep(VarRankBins2, length(conditions)*length(TissueBias3)))

for (c in conditions) {
  for (b in TissueBias3) {
    for (v in VarRankBins2) {
      ## Proportion of genes assigned to dummy module (gray; 0)
      modularity.metrics.perm$PropDummy[modularity.metrics.perm$Condition==c & 
                                              modularity.metrics.perm$TissueBias3==b &
                                              modularity.metrics.perm$VarRankBin2==v] <- 
        nrow(connectivity.perm[[c]][[b]][[v]][connectivity.perm[[c]][[b]][[v]]$module==0,]) /
        nrow(connectivity.perm[[c]][[b]][[v]])
    
    ## Proportion of genes assigned to top module (1)
    modularity.metrics.perm$PropTop[modularity.metrics.perm$Condition==c &
                                          modularity.metrics.perm$TissueBias3==b &
                                          modularity.metrics.perm$VarRankBin2==v] <- 
      nrow(connectivity.perm[[c]][[b]][[v]][connectivity.perm[[c]][[b]][[v]]$module==1,]) /
      nrow(connectivity.perm[[c]][[b]][[v]])    
    
    ## Distribution of kDiff for genes assigned to other modules
    modularity.metrics.perm$Mean_kDiff[modularity.metrics.perm$Condition==c &
                                             modularity.metrics.perm$TissueBias3==b &
                                             modularity.metrics.perm$VarRankBin2==v] <- 
      mean(connectivity.perm[[c]][[b]][[v]]$kDiff[connectivity.perm[[c]][[b]][[v]]$module!=0])
    
    modularity.metrics.perm$Median_kDiff[modularity.metrics.perm$Condition==c &
                                               modularity.metrics.perm$TissueBias3==b &
                                               modularity.metrics.perm$VarRankBin2==v] <- 
      median(connectivity.perm[[c]][[b]][[v]]$kDiff[connectivity.perm[[c]][[b]][[v]]$module!=0])
    
    ## Distribution of kDiff for genes assigned to dummy module
    modularity.metrics.perm$Mean_kDiff_Dummy[modularity.metrics.perm$Condition==c &
                                                   modularity.metrics.perm$TissueBias3==b &
                                                   modularity.metrics.perm$VarRankBin2==v] <- 
      mean(connectivity.perm[[c]][[b]][[v]]$kDiff[connectivity.perm[[c]][[b]][[v]]$module==0])
    
    modularity.metrics.perm$Median_kDiff_Dummy[modularity.metrics.perm$Condition==c &
                                                     modularity.metrics.perm$TissueBias3==b &
                                                     modularity.metrics.perm$VarRankBin2==v] <- 
      median(connectivity.perm[[c]][[b]][[v]]$kDiff[connectivity.perm[[c]][[b]][[v]]$module==0])
    
    ## Distribution of kWithin for genes assigned to other modules
    modularity.metrics.perm$Mean_kWithin[modularity.metrics.perm$Condition==c &
                                               modularity.metrics.perm$TissueBias3==b &
                                               modularity.metrics.perm$VarRankBin2==v] <- 
      mean(connectivity.perm[[c]][[b]][[v]]$kWithin[connectivity.perm[[c]][[b]][[v]]$module!=0])
    
    modularity.metrics.perm$Median_kWithin[modularity.metrics.perm$Condition==c &
                                                 modularity.metrics.perm$TissueBias3==b & 
                                                 modularity.metrics.perm$VarRankBin2==v] <- 
      median(connectivity.perm[[c]][[b]][[v]]$kWithin[connectivity.perm[[c]][[b]][[v]]$module!=0])
    
    ## Distribution of kWithin for genes assigned to dummy module
    modularity.metrics.perm$Mean_kWithin_Dummy[modularity.metrics.perm$Condition==c &
                                                     modularity.metrics.perm$TissueBias3==b &
                                                     modularity.metrics.perm$VarRankBin2==v] <- 
      mean(connectivity.perm[[c]][[b]][[v]]$kWithin[connectivity.perm[[c]][[b]][[v]]$module==0])
    
    modularity.metrics.perm$Median_kWithin_Dummy[modularity.metrics.perm$Condition==c &
                                                       modularity.metrics.perm$TissueBias3==b &
                                                       modularity.metrics.perm$VarRankBin2==v] <- 
      median(connectivity.perm[[c]][[b]][[v]]$kWithin[connectivity.perm[[c]][[b]][[v]]$module==0])
          
    }

  }
}
```


### 2.5 Summary Plots
```{r plots-permuted}
ggplot(modularity.metrics.perm, 
       aes(x=VarRankBin2, y=Condition, fill=PropDummy)) + 
  geom_tile() +
  facet_wrap(vars(TissueBias3)) +
  colorspace::scale_fill_continuous_diverging(mid=0, na.value="#A9A9A9") +
  labs(x="Variability rank bin", fill="Prop(no module)") +
  theme_bw()

ggplot(modularity.metrics.perm, 
       aes(x=VarRankBin2, y=Condition, fill=PropTop)) + 
  geom_tile() +
  facet_wrap(vars(TissueBias3)) +
  colorspace::scale_fill_continuous_diverging(mid=0, na.value="#A9A9A9") +
  labs(x="Variability rank bin", fill="Prop(top module)") +
  theme_bw()

ggplot(modularity.metrics.perm, 
       aes(x=VarRankBin2, y=Condition, fill=log2(Mean_kDiff - min(Mean_kDiff, na.rm=TRUE) + 1))) + 
  geom_tile() +
  facet_wrap(vars(TissueBias3)) +  
  colorspace::scale_fill_continuous_diverging(mid=0, na.value="#A9A9A9") +
  labs(x="Variability rank bin", fill="Mean kDiff") +
  theme_bw()

ggplot(modularity.metrics.perm, 
       aes(x=VarRankBin2, y=Condition, fill=log2(Median_kDiff - min(Median_kDiff, na.rm=TRUE) + 1))) + 
  geom_tile() +
  facet_wrap(vars(TissueBias3)) +
  colorspace::scale_fill_continuous_diverging(mid=0, na.value="#A9A9A9") +
  labs(x="Variability rank bin", fill="Median kDiff") +
  theme_bw()

ggplot(modularity.metrics.perm, 
       aes(x=VarRankBin2, y=Condition, fill=log2(Mean_kWithin +1))) + 
  geom_tile() +
  facet_wrap(vars(TissueBias3)) +
  colorspace::scale_fill_continuous_diverging(mid=0, na.value="#A9A9A9") +
  labs(x="Variability rank bin", fill="Mean kWithin") +
  theme_bw()

ggplot(modularity.metrics.perm, 
       aes(x=VarRankBin2, y=Condition, fill=log2(Median_kWithin +1))) + 
  geom_tile() +
  facet_wrap(vars(TissueBias3)) +
  colorspace::scale_fill_continuous_diverging(mid=0, na.value="#A9A9A9") +
  labs(x="Variability rank bin", fill="Median kWithin") +
  theme_bw()
```


## 3.0 Compare observed vs. randomized
```{r modularity-enrichment}
modularity.enrichment <- modularity.metrics[,c("Condition","TissueBias3","VarRankBin2")]

modularity.enrichment$LogFC_PropDummy <- 
  log2((modularity.metrics$PropDummy + 1) / (modularity.metrics.perm$PropDummy + 1))

modularity.enrichment$LogFC_PropTop <-
  log2((modularity.metrics$PropTop + 1) / (modularity.metrics.perm$PropTop + 1))

modularity.enrichment[,6:13] <- 
  modularity.metrics[,6:13] - modularity.metrics.perm[,6:13]

colnames(modularity.enrichment)[6:13] <-
  levels(interaction("Resid",colnames(modularity.metrics)[6:13], sep="_"))
```


### 3.1 Summary plots 
```{r plots-enrichment}
ggplot(modularity.enrichment, 
       aes(x=VarRankBin2, y=Condition, fill=LogFC_PropDummy)) + 
  geom_tile() +
  facet_wrap(vars(TissueBias3)) +
  colorspace::scale_fill_continuous_diverging(mid=0, na.value="#A9A9A9") +
  labs(x="Variability rank bin", fill="LogFC prop(no module)") +
  theme_bw()

ggplot(modularity.enrichment, 
       aes(x=VarRankBin2, y=Condition, fill=LogFC_PropTop)) + 
  geom_tile() +
  facet_wrap(vars(TissueBias3)) +
  colorspace::scale_fill_continuous_diverging(mid=0, na.value="#A9A9A9") +
  labs(x="Variability rank bin", fill="LogFC prop(top module)") +
  theme_bw()

ggplot(modularity.enrichment, 
       aes(x=VarRankBin2, y=Condition, fill=log2(Resid_Mean_kDiff))) + 
  geom_tile() +
  facet_wrap(vars(TissueBias3)) +  
  colorspace::scale_fill_continuous_diverging(mid=0, na.value="#A9A9A9") +
  labs(x="Variability rank bin", fill="Residual mean kDiff") +
  theme_bw()

ggplot(modularity.enrichment, 
       aes(x=VarRankBin2, y=Condition, fill=log2(Resid_Median_kDiff))) + 
  geom_tile() +
  facet_wrap(vars(TissueBias3)) +  
  colorspace::scale_fill_continuous_diverging(mid=0, na.value="#A9A9A9") +
  labs(x="Variability rank bin", fill="Residual median kDiff") +
  theme_bw()

ggplot(modularity.enrichment, 
       aes(x=VarRankBin2, y=Condition, fill=log2(Resid_Mean_kWithin))) + 
  geom_tile() +
  facet_wrap(vars(TissueBias3)) +  
  colorspace::scale_fill_continuous_diverging(mid=0, na.value="#A9A9A9") +
  labs(x="Variability rank bin", fill="Residual mean kWithin") +
  theme_bw()

ggplot(modularity.enrichment, 
       aes(x=VarRankBin2, y=Condition, fill=log2(Resid_Median_kWithin))) + 
  geom_tile() +
  facet_wrap(vars(TissueBias3)) +  
  colorspace::scale_fill_continuous_diverging(mid=0, na.value="#A9A9A9") +
  labs(x="Variability rank bin", fill="Residual median kWithin") +
  theme_bw()
```


# Save {.tabset}
```{r save-objects}
if (!params$combat) {
  if (!params$qc1.only) {
    output.path <- file.path(params$output.rds.path, "no_combat", "full_qc")    
  } else {
    output.path <- file.path(params$output.rds.path, "no_combat", "qc1_only")        
  }

} else {
  if (!params$qc1.only) {
    output.path <- file.path(params$output.rds.path, "combat", "full_qc")    
  } else {
    output.path <- file.path(params$output.rds.path, "combat", "qc1_only")        
  }

}

if (!dir.exists(output.path)) { dir.create(output.path, recursive=TRUE, showWarnings=FALSE) }

saveRDS(connectivity,
        file.path(output.path, 
                  paste(params$species, "connectivity", 
                        paste(paste0("tau", params$tau.cutoff),
                              paste0("rnd", params$set.seed), "rds", sep="."), sep="_")))

saveRDS(modularity.metrics,
        file.path(output.path, 
                  paste(params$species, "modularity.metrics", 
                        paste(paste0("tau", params$tau.cutoff),
                              paste0("rnd", params$set.seed), "rds", sep="."), sep="_")))

saveRDS(modularity.enrichment,
        file.path(output.path, 
                  paste(params$species, "modularity.enrichment", 
                        paste(paste0("tau", params$tau.cutoff),
                              paste0("rnd", params$set.seed), "rds", sep="."), sep="_")))
```

