---
title: "Extract supplementary tables"
author: "Bucao, Christabel Floi"
output: html_document
date: "`r Sys.Date()`"
params:
  input.var.path: "../../results/03_Variability_Plots"
  input.go.path: "../../results/07_GO_Enrichment"
  input.camera.path: "../../results/07b_GO_cameraPR"
  input.selectome.path: "../../results/08_Selection"
  input.tau.path: "../../results/10_Expression_Specificity"
  input.organ.var.path: "../../results/12_Organ_Bias"
  species: "DRE"
  species.name: "Danio rerio"
  tau.cutoff: 0.30
  qc1.only: FALSE
  combat: FALSE
---

# Libraries
```{r}
library(dplyr)
```


# Supplemental Table 4 - Variability metric

## Import rds
```{r}
if (!params$combat) {
  if (!params$qc1.only) {
    input.path <- file.path(params$input.var.path, "no_combat", "full_qc")    
  } else {
    input.path <- file.path(params$input.var.path, "no_combat", "qc1_only")        
  }

} else {
  if (!params$qc1.only) {
    input.path <- file.path(params$input.var.path, "combat", "full_qc")    
  } else {
    input.path <- file.path(params$input.var.path, "combat", "qc1_only")
  }

}

species <- c("DRE"="Danio_rerio", "ELU"="Esox_lucius", "LOC"="Lepisosteus_oculatus")

jack.resid.cv.bind <- lapply(setNames(names(species), species), function(sp) {
  readRDS(file.path(input.path, paste(sp, "jack.resid.cv.bind.rds", sep="_")))
})
```

## Organize df
```{r}
jack.resid.cv.bind[["Lepisosteus_oculatus"]]$Sex <- NA
jack.resid.cv.bind <- dplyr::bind_rows(jack.resid.cv.bind, .id="Species")
```

## Export df as tsv
```{r}
write.table(jack.resid.cv.bind, 
            file=file.path(input.path, "Supplemental_Table_S4.tsv"),
            quote=TRUE, row.names=FALSE, col.names=TRUE, sep="\t")
```


# Supplemental Table 5 - GOStats

## Import rds
```{r}
if (!params$combat) {
  if (!params$qc1.only) {
    input.path <- file.path(params$input.go.path, "no_combat", "full_qc")    
  } else {
    input.path <- file.path(params$input.go.path, "no_combat", "qc1_only")        
  }

} else {
  if (!params$qc1.only) {
    input.path <- file.path(params$input.go.path, "combat", "full_qc")    
  } else {
    input.path <- file.path(params$input.go.path, "combat", "qc1_only")
  }

}

go.aggregate <- readRDS(file.path(input.path, paste(params$species, "go.aggregate.rds", sep="_"))) 
```

## Organize df
```{r}
EV <- c("LV","MV","HV")

# Enriched
go.aggregate.over.bind <- lapply(setNames(EV, EV), function(ev) {
  go.aggregate[[ev]][["over"]]
}) %>%
  dplyr::bind_rows(.id="VarCategory")
go.aggregate.over.bind$EnrichmentTest <- rep("GOStats",nrow(go.aggregate.over.bind))
go.aggregate.over.bind$TestDirection <- rep("enrichment",nrow(go.aggregate.over.bind))

# Depleted
go.aggregate.under.bind <- lapply(setNames(EV, EV), function(ev) {
  go.aggregate[[ev]][["under"]]
}) %>%
  dplyr::bind_rows(.id="VarCategory")
go.aggregate.under.bind$EnrichmentTest <- rep("GOStats",nrow(go.aggregate.under.bind))
go.aggregate.under.bind$TestDirection <- rep("depletion",nrow(go.aggregate.under.bind))


go.aggregate.bind <- go.aggregate.over.bind %>%
  dplyr::bind_rows(go.aggregate.under.bind)

rm(go.aggregate.over.bind, go.aggregate.under.bind)
```

## Export df as tsv
```{r}
write.table(go.aggregate.bind, 
            file=file.path(input.path, "Supplemental_Table_S5.tsv"),
            quote=TRUE, row.names=FALSE, col.names=TRUE, sep="\t")
```


# Supplemental Table 6 - cameraPR

## Import rds
```{r}
if (!params$combat) {
  if (!params$qc1.only) {
    input.path <- file.path(params$input.camera.path, "no_combat", "full_qc")    
  } else {
    input.path <- file.path(params$input.camera.path, "no_combat", "qc1_only")        
  }

} else {
  if (!params$qc1.only) {
    input.path <- file.path(params$input.camera.path, "combat", "full_qc")    
  } else {
    input.path <- file.path(params$input.camera.path, "combat", "qc1_only")
  }

}

camera.bind <- readRDS(file.path(input.path, paste(params$species, "camera.bind.rds", sep="_")))
```

## Organize df
```{r}
camera.bind <- dplyr::bind_rows(camera.bind)
camera.bind$EnrichmentTest <- rep("cameraPR", nrow(camera.bind))
```

## Export df as tsv
```{r}
write.table(camera.bind, 
            file=file.path(input.path, "Supplemental_Table_S6.tsv"),
            quote=TRUE, row.names=FALSE, col.names=TRUE, sep="\t")
```


# Supplemental Tables 7 & 8 - Selectome

## Import rds
```{r}
if (!params$combat) {
  if (!params$qc1.only) {
    input.path <- file.path(params$input.selectome.path, "no_combat", "full_qc")    
  } else {
    input.path <- file.path(params$input.selectome.path, "no_combat", "qc1_only")        
  }

} else {
  if (!params$qc1.only) {
    input.path <- file.path(params$input.selectome.path, "combat", "full_qc")    
  } else {
    input.path <- file.path(params$input.selectome.path, "combat", "qc1_only")
  }

}


selectome.sub <- 
  readRDS(file.path(input.path, paste(params$species, "selectome.sub.rds", sep="_")))
selectome.sub <- selectome.sub[["Bgd"]]


selectome.groupby.ev <- 
  readRDS(file.path(input.path, paste(params$species, "selectome.bgd.groupby.gene.ev.median.rds", sep="_")))
```

## Export df as tsv
```{r}
write.table(selectome.sub, 
            file=file.path(input.path, "Supplemental_Table_S7.tsv"),
            quote=TRUE, row.names=FALSE, col.names=TRUE, sep="\t")

write.table(selectome.groupby.ev, 
            file=file.path(input.path, "Supplemental_Table_S8.tsv"),
            quote=TRUE, row.names=FALSE, col.names=TRUE, sep="\t")
```


# Supplemental Table S10 - Expression specificity

## Import rds
```{r}
if (!params$combat) {
  if (!params$qc1.only) {
    input.path <- file.path(params$input.tau.path, "no_combat", "full_qc")    
  } else {
    input.path <- file.path(params$input.tau.path, "no_combat", "qc1_only")        
  }

} else {
  if (!params$qc1.only) {
    input.path <- file.path(params$input.tau.path, "combat", "full_qc")    
  } else {
    input.path <- file.path(params$input.tau.path, "combat", "qc1_only")
  }

}

species <- c("DRE"="Danio_rerio", "ELU"="Esox_lucius", "LOC"="Lepisosteus_oculatus")

filtered.samples.common <- lapply(setNames(names(species), species), function(sp) {
      readRDS(file.path(input.path, paste(sp, "filtered.samples.common.rds", sep="_")))   
})
```

## Organize df
```{r}
expr.rank.first <- lapply(setNames(species, species), function(sp) {
  filtered.samples.common[[sp]]$expr.rank.first
}) %>%
  dplyr::bind_rows(.id="Species")
```

## Export df as tsv
```{r}
write.table(expr.rank.first, 
            file=file.path(input.path, "Supplemental_Table_S10.tsv"),
            quote=TRUE, row.names=FALSE, col.names=TRUE, sep="\t")
```


# Supplemental Table S11 - Organ bias & variability

## Import rds
```{r}
if (!params$combat) {
  if (!params$qc1.only) {
    input.path <- file.path(params$input.organ.var.path, "no_combat", "full_qc")    
  } else {
    input.path <- file.path(params$input.organ.var.path, "no_combat", "qc1_only")        
  }

} else {
  if (!params$qc1.only) {
    input.path <- file.path(params$input.organ.var.path, "combat", "full_qc")    
  } else {
    input.path <- file.path(params$input.organ.var.path, "combat", "qc1_only")
  }

}

species <- c("DRE"="Danio_rerio", "ELU"="Esox_lucius", "LOC"="Lepisosteus_oculatus")

comparisons.expr.var <- lapply(setNames(names(species), species), function(sp) {
  readRDS(file.path(input.path, 
                    paste(sp, "comparisons.expr.var",
                          paste(paste0("tau", params$tau.cutoff), "rds", sep="."), sep="_")))  
})
```

## Organize df
```{r}
comparisons.wilcox.complete <- lapply(setNames(species, species), function(sp) {
  comparisons.expr.var[[sp]]$wilcox.complete
}) %>%
  dplyr::bind_rows(.id="Species")
```

## Export df as tsv
```{r}
write.table(comparisons.wilcox.complete, 
            file=file.path(input.path, "Supplemental_Table_S11.tsv"),
            quote=TRUE, row.names=FALSE, col.names=TRUE, sep="\t")
```


